---
title: "29 Uncertainties Monochrome"
author: "Merritt Harlan"
date: "2023-04-19"
output: html_document
---

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

Define libraries here. 

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse)
library(readr) # best for reading and writing csvs
library(downloader)
library(dataRetrieval)
library(raster)
library(sf)
library(dplyr)

# These are used for the layout of the viz
library(ggpubr)
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo
library(scales)
```



## Load files

```{r load}
# load Tanana data from data release
tanana_zip <- download("https://alaska.usgs.gov/data/hydrology/streamflow_satelliteComputed_alaska/tanana_fairbanks.zip", destfile = "in/tanana_fairbanks.zip")

unzip("in/tanana_fairbanks.zip", exdir = "in/tanana_fairbanks")

tanana_flow <- read_csv("in/tanana_fairbanks/tanana_fairbanks_Zone 6_008051_Tanana_TPJOJ_2_5_1_RSQ_DATA.csv")

tanana_dswe <- read_csv("in/tanana_fairbanks/tanana_fairbanks_Zone 6_008051_Tanana_TPJOJ_2_5_1_altim_DSWE_relation_DATA.csv")[c(2:6),]#remove first row, which contains a Landsat 7 image, with known scanline corrector issues

tanana_shp <- read_sf("in/tanana_fairbanks/shapefile/tanana_fairbank_zone7.shp")

# load Tanana Fairbanks gage data
tanana_gage <- readNWISdv(siteNumbers = c("15485500"), parameterCd = "00060")

tanana_site <- readNWISsite(siteNumbers = c("15485500"))

# read in satellite imagery files
DSWE_files <- list.files("in/DSWE/", full.names = TRUE)

```

## Get data ready for plotting

```{r processing}
# Add your processing code here
tanana_dswe_cleaned = as.data.frame(tanana_dswe) %>%
  mutate(Date = as.Date(date, format = "%m/%d/%Y")) %>%
  mutate(width_ft = DSWE12/length_km * 3.2808399 * 1000) %>%
  mutate(wse_ft = wse_m * 3.2808399) %>%
  dplyr::select(Date, width_ft, wse_ft, tileName)

tanana_flow_cleaned = data.frame(tanana_flow) %>% 
  mutate(Date = as.Date(datetime, format = "%m/%d/%Y %H:%M")) %>%
  mutate(width_ft = width_m * 3.2808399) %>%
  mutate(wse_ft = wse_m * 3.2808399) %>% 
  mutate(Q_satellite = Q_cfs) %>%
  right_join(tanana_gage) %>% 
  mutate(Q_gage = X_00060_00003) %>%
  filter(Date > min(tanana_dswe_cleaned$Date)) %>%
  filter(Date < max(tanana_dswe_cleaned$Date)) %>% 
  dplyr::select(Date, wse_ft, Q_satellite, Q_gage)

DSWE_crop <- function(DSWE_file){
  r = raster(DSWE_file)
  date = as.Date(substr(DSWE_file, nchar(DSWE_file)-33, nchar(DSWE_file)-26), format = "%Y%m%d")
  shp = tanana_shp %>% st_transform(crs(r)) %>% st_zm()
  masked = mask(crop(r, shp), shp)
  spdf = as.data.frame(as(masked, "SpatialPixelsDataFrame"))
  dswe_meaning = data.frame(value = c(0:4), DSWE = c("", "water - high confidence", "water - moderate conference", "potential wetland", "water/wetland - low confidence"))
  colnames(spdf) = c("value", "x", "y")
  spdf$Date = date
  spdf = spdf %>% left_join(dswe_meaning)
  return(spdf)
}

dswe_images <- lapply(DSWE_files, DSWE_crop)

```

## Set up main plot

```{r plotting}
# Load some custom fonts and set some custom settings
supporting_font <- "Source Sans Pro"
sysfonts::font_add_google("Source Sans Pro")
showtext::showtext_opts(dpi = 300, regular.wt = 200, bold.wt = 700)
showtext::showtext_auto(enable = TRUE)

title_fontsize = 40
subtitle_fontsize = 18
plot_fontsize = 16

# Define colors
background_color = "white"
font_color = "dodgerblue4"

# The background canvas for your viz
canvas <- grid::rectGrob(
  x = 0, y = 0, 
  width = 16, height = 9,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color)
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_black.png") %>% magick::image_colorize(100, font_color)

# Plots
width_plot <- ggplot(tanana_dswe_cleaned, aes(x = wse_ft, y = width_ft)) + 
  geom_point(size = 3, alpha = 0.5, color = font_color) + 
  theme_bw() + 
  scale_y_continuous(labels = comma) +
  xlab("Satellite River Elevation (ft)") + 
  ylab("Satellite River Width (ft)") +
  theme(
    text=element_text(size=plot_fontsize, color = font_color, family = supporting_font),
    axis.text = element_text(color = font_color),
    axis.line = element_line(color = font_color),
    axis.title.x = element_text(size=plot_fontsize, face="bold", family = supporting_font),
    axis.title.y = element_text(size=plot_fontsize, face="bold", family = supporting_font),
    panel.border = element_blank(),
    panel.background = element_rect(fill='transparent', color = NA), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
    panel.grid.major = element_blank(), #remove major gridlines
    panel.grid.minor = element_blank(), #remove minor gridlines
    legend.background = element_rect(fill='transparent'), #transparent legend bg
  )

altim_plot <- ggplot(tanana_flow_cleaned, aes(x = wse_ft, y = Q_satellite)) + 
  geom_point(size = 3, alpha = 0.5, color = font_color) + 
  theme_bw() + 
  scale_y_continuous(labels = comma) +
  xlab("Satellite River Elevation (ft)") + 
  ylab(expression(bold(paste("Satellite Streamflow (", ft^{"3"},"/s)"))))+
  theme(
    text = element_text(size=plot_fontsize, color = font_color, family = supporting_font), 
    axis.text = element_text(color = font_color),
    axis.line = element_line(color = font_color),
    axis.title.x = element_text(size=plot_fontsize, face="bold", family = supporting_font),
    axis.title.y = element_text(size=plot_fontsize, face="bold", family = supporting_font),
    panel.border = element_blank(),
    panel.background = element_rect(fill='transparent', color = NA), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
    panel.grid.major = element_blank(), #remove major gridlines
    panel.grid.minor = element_blank(), #remove minor gridlines
    legend.background = element_rect(fill='transparent'), #transparent legend bg
  )

flow_plot <- ggplot(tanana_flow_cleaned) + 
  geom_line(aes(x = Date, y = Q_gage, color = "Gage"), size = 1) +
  geom_point(aes(x = Date, y = Q_satellite, color= "Satellite"), size = 3, alpha = 0.5) + 
  theme_bw() + xlab("Year") + 
  ylab(expression(bold(paste("Satellite River Flow (", ft^{"3"},"/s)")))) + 
  scale_y_continuous(labels = comma) +
  scale_color_manual(name = NULL, breaks = c("Satellite", "Gage"), values = c("Satellite" = font_color, "Gage" = "lightskyblue")) +
  theme(
    legend.position = "right",
    legend.text = element_text(size = plot_fontsize, family = supporting_font),
    text = element_text(size=plot_fontsize, color = font_color), 
    axis.text = element_text(color = font_color),
    axis.line = element_line(color = font_color),
    axis.title.x = element_text(size=plot_fontsize, face="bold", family = supporting_font),
    axis.title.y = element_text(size=plot_fontsize, face="bold", family = supporting_font),
    panel.border = element_blank(),
    panel.background = element_rect(fill='transparent', color = NA), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
    panel.grid.major = element_blank(), #remove major gridlines
    panel.grid.minor = element_blank(), #remove minor gridlines
    legend.background = element_rect(fill='transparent'), #transparent legend bg
  ) +
  guides(color = guide_legend(override.aes = list(linetype = c(NA, 1), shape = c(19, NA))))

dswe_image = dswe_images[[1]]
dswe_image$DSWE = factor(dswe_image$DSWE, levels = c("", "water - high confidence", "water - moderate conference", "potential wetland", "water/wetland - low confidence"))

dswe_plot <- ggplot() +  
  geom_tile(data=dswe_image, aes(x=x, y=y, fill= DSWE), alpha=0.8) + 
  coord_equal() +
  scale_fill_manual(name = "Classification", values = c("white", "dodgerblue4", "dodgerblue", "lightskyblue", "lightblue1")) +
  theme_map() +
  theme(legend.position="right",
        legend.text = element_text(size = subtitle_fontsize, color = font_color, family = supporting_font),
        legend.title = element_text(size = subtitle_fontsize, color = font_color, face = "bold", family = supporting_font))
        
#Get outline of AK for basemap
AK_shp <- spData::alaska %>% st_as_sf()

AK_inset <- ggplot() + 
  geom_sf(data = AK_shp, color = "lightskyblue", fill = "lightskyblue") +
  geom_sf(data = st_as_sf(tanana_site, coords = c("dec_long_va", "dec_lat_va"), crs = 4267) %>% st_transform(st_crs(AK_shp)),
          color = "dodgerblue4", linewidth = 2) + 
  theme_map()

```


## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

```{r cowplot, fig.width = 16, fig.height = 9}
#Static Image
ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 9, width = 16,
            hjust = 0, vjust = 1) +
  # satellite image
  draw_plot(dswe_plot,
            x = 0.00,
            y = 0.03,
            width = 0.96) +
  # width height plot
  draw_plot(width_plot,
            x = 0.01,
            y = 0.01,
            height = 0.34,
            width = 0.22) +
  # height flow plot
  draw_plot(altim_plot,
            x = 0.25,
            y = 0.01,
            height = 0.34,
            width = 0.22) +
  # height flow plot
  draw_plot(flow_plot,
            x = 0.49,
            y = 0.01,
            height = 0.34,
            width = 0.34) +
  # Alaska inset map
  draw_plot(AK_inset,
            x = 0.75,
            y = 0.72,
            height = 0.25,
            width = 0.25) +
  # explainer text
  draw_label(paste('Merritt Harlan, USGS', 
                   'Data: Smith, T. L., et al., 2022, Computed streamflow using',
                   'satellite data for the Copper, Nushagak, Tanana, Yukon,', 
                   'Susitna, and Knik, Koyukuk Rivers, Alaska, 2008-2021:',
                   'USGS data release, https://doi.org/10.5066/P94LLG4R.',
                   'Tanana River in Fairbanks, AK- USGS Site 15485500', sep = '\n'),
             fontfamily = supporting_font,
             fontface = "italic",
             x = 0.96,
             y = 0.03,
             size = 10,
             hjust = 1,
             vjust = 0,
             color = font_color)+
 # Title
  draw_label(stringr::str_wrap("Estimating streamflow from satellites",60),
             x = 0.03,
             y = 0.92,
             hjust = 0,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = supporting_font,
             fontface = 'bold',
             color = font_color,
             size = title_fontsize) +
  # Annotation
  draw_label(paste("By relating optical satellite imagery, which can give us an estimate of river width,",
                   "to satellite altimetry, which gives us an estimate of river elevation, we can approximate",
                   "streamflow in large rivers. Although this data can have high uncertainty, combined with ",
                   "knowledge of rivers and ground data, satellites can help us improve monitoring.", sep = '\n'),
             x = 0.05,
             y = 0.77,
             hjust = 0,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = supporting_font,
             color = font_color,
             size = subtitle_fontsize) +
  # Add logo
  draw_image(usgs_logo,
             x = 0.96,
             y = 0.15,
             width = 0.10,
             hjust = 1, vjust = 0,
             halign = 0, valign = 0)

# Save the final image in Twitter's 16 by 9 format
# !! Use format for saving with the date of your prompt: 
#         YYYYMMDD_prompt_name ()
# e.g. `20230101_part-to-whole-cnell.png`
ggsave(filename = "out/20230429_monochrome_mharlan.png", 
       width = 14, height = 9, dpi = 300)
```

## Final GIF

```{r gif}
##### Create 3D GIF for output #####

# Set up parameters for loop to create GIF
n_frames <- length(dswe_images)
img_frames <- paste0("out/cache/dswe_uncertainty_", seq_len(n_frames), ".png")


for(i in 1:n_frames){
  dswe_index = arrange(tanana_dswe_cleaned, width_ft)[i,]
  date = dswe_index$Date
  flow_index = tanana_flow_cleaned %>% filter(Date == date)
  
  
  filtered_image = lapply(dswe_images, function(x) filter(x, Date == date))
  dswe_image = data.frame()
  for(x in 1:length(filtered_image)){
    if(nrow(filtered_image[[x]])>0){
      dswe_image= filtered_image[[x]]
    }
  }
  
  dswe_image$DSWE = factor(dswe_image$DSWE, levels = c("", "water - high confidence", "water - moderate conference", "potential wetland", "water/wetland - low confidence"))
  
  dswe_plot <- ggplot() +  
    geom_tile(data=dswe_image, aes(x=x, y=y, fill= DSWE), alpha=0.8) + 
    coord_equal() +
    scale_fill_manual(name = "Classification", values = c("white", "dodgerblue4", "dodgerblue", "lightskyblue", "lightblue1")) +
    theme_map() +
  theme(legend.position="right",
        legend.text = element_text(size = subtitle_fontsize, 
                                   color = font_color, 
                                   family = supporting_font),
        legend.title = element_text(size = subtitle_fontsize, 
                                    color = font_color, 
                                    face = "bold", 
                                    family = supporting_font))
  
  width_plot_new <- width_plot + geom_point(data = dswe_index, aes(x = wse_ft, y = width_ft), 
                                            size = 7, shape = 19, color = "dodgerblue")

  ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 9, width = 16,
            hjust = 0, vjust = 1) +
  draw_plot(dswe_plot,
            x = 0.00,
            y = 0.03,
            width = 0.96) +
  # width height plot
  draw_plot(width_plot_new,
            x = 0.01,
            y = 0.01,
            height = 0.34,
            width = 0.22) +
  # height flow plot
  draw_plot(altim_plot,
            x = 0.25,
            y = 0.01,
            height = 0.34,
            width = 0.22) +
  # height flow plot
  draw_plot(flow_plot,
            x = 0.49,
            y = 0.01,
            height = 0.34,
            width = 0.34) +
  # Alaska inset map
  draw_plot(AK_inset,
            x = 0.75,
            y = 0.72,
            height = 0.25,
            width = 0.25) +
  # explainer text
  draw_label(paste('Merritt Harlan, USGS', 
                   'Data: Smith, T. L., et al., 2022, Computed streamflow using',
                   'satellite data for the Copper, Nushagak, Tanana, Yukon,', 
                   'Susitna, and Knik, Koyukuk Rivers, Alaska, 2008-2021:',
                   'USGS data release, https://doi.org/10.5066/P94LLG4R.',
                   'Tanana River in Fairbanks, AK- USGS Site 15485500', sep = '\n'),
             fontfamily = supporting_font,
             fontface = "italic",
             x = 0.96,
             y = 0.03,
             size = 10,
             hjust = 1,
             vjust = 0,
             color = font_color)+
 # Title
  draw_label(stringr::str_wrap("Estimating streamflow from satellites",60),
             x = 0.03,
             y = 0.92,
             hjust = 0,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = supporting_font,
             fontface = 'bold',
             color = font_color,
             size = title_fontsize) +
  # Annotation
  draw_label(paste("By relating optical satellite imagery, which can give us an estimate of river width,",
                   "to satellite altimetry, which gives us an estimate of river elevation, we can approximate",
                   "streamflow in large rivers. Although this data can have high uncertainty, combined with ",
                   "knowledge of rivers and ground data, satellites can help us improve monitoring.", sep = '\n'),
             x = 0.05,
             y = 0.77,
             hjust = 0,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = supporting_font,
             color = font_color,
             size = subtitle_fontsize) +
  # Add logo
  draw_image(usgs_logo,
             x = 0.96,
             y = 0.15,
             width = 0.10,
             hjust = 1, vjust = 0,
             halign = 0, valign = 0)


  # Save the final image in Twitter's 16 by 9 format
  # !! Use format for saving with the date of your prompt: 
  #         YYYYMMDD_prompt_name ()
  # e.g. `20230101_part-to-whole-cnell.png`
  ggsave(filename = img_frames[i], 
         width = 14, height = 9, dpi = 300)
}

# Build GIF using cached images - add labels here to control size better
magick::image_read(img_frames) %>%
  magick::image_write_gif(path="out/20230429_monochrome_mharlan.gif",delay=12/n_frames) # Number here is duration

```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. Satellite optical and altimetry data can be combined to estimate streamflow, as shown here for the Tanana River in Fairbanks Alaska. We can use the classified water extent to estimate river width by dividing the water extent area and the length of the river, and match this data with overlapping satellite altimeters to build relationships between river width, elevation, and streamflow. 

2. Future data from the Surface Water Ocean Topography (SWOT) satellite will provide simultaneous water extent and elevation, providing even more data!

### Data source(s)

Landsat Collection 2 Level-3 Dynamic Surface Water Extent (DSWE) Science Products courtesy of the U.S. Geological Survey.

Jones, J.W., 2019. Improved Automated Detection of Subpixel-Scale Inundation—Revised Dynamic Surface Water Extent (DSWE) Partial Surface Water Tests. Remote Sens., 11, 374 https://doi.org/10.3390/rs11040374.

Smith, T. L., et al., 2022, Computed streamflow using satellite data for the Copper, Nushagak, Tanana, Yukon, Susitna, and Knik, Koyukuk Rivers, Alaska, 2008-2021: United States Geological Survey data release, https://doi.org/10.5066/P94LLG4R.

Tanana River in Fairbanks, AK- USGS Site 15485500

DSWE data can be downloaded from https://earthexplorer.usgs.gov/
Instructions: 
- draw a bounding box around Tanana Fairbanks (64 43' 36'' N, 146 51' 59'' W) OR upload shapefile downloaded from Smith et al., 2022 data repository
- select dates corresponding to imagery (09/18/2014, 07/03/2015, 06/06/2017, 07/06/2017, 07/06/2011)
- filter data to Landsat 4-9 C2 Dynamic Surface Water Extent
- download INTR scene for each date

