---
title: "28 Trends - How much water is entering Lake Tahoe?"
author: "Margaux Sleckman"
date: "2023-04-28"
output: html_document
---
## Set up: 

```{r setup, include=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

```{r libraries,include=FALSE, warning=FALSE, message=FALSE}

# Load libraries
## main packages
library(tidyverse) # includes ggplot
library(readr) # best for reading and writing csvs
library(lubridate)
library(glue)
library(zoo)
library(purrr)

## spatial packages
library(sf)
library(mapdata)
library(maps)
library(spData)
library(ggspatial)
library(ggrepel)
library(mapview) # for interactive map viz

# viz packages
library(gghighlight)
library(scales)
library(colorfindr)

# These are used for the layout of the viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo

```

```{r options_dir_creation, include=FALSE, warning=FALSE, message=FALSE}

options(tidyverse.quiet = TRUE, timeout = 500)
mapviewOptions(fgb = FALSE)

## Creating local folders for downloaded data
dir.create('in/R', showWarnings = F)
dir.create('in/R/nhdPlusTools', showWarnings = F)

```

## Fetch geospatial sata for viz

### A. Fetching nhdplus spatial data w/ nhdplustools

_Note: code hidden in markdown output_

```{r generic_nhdPlus_data_fetch, include=FALSE, warning=FALSE, message=FALSE}

# state boundarie --------------------------------------------------------

CA_sf <- us_states |> filter(NAME == 'California')

# Grabbing Huc 8 to create the AOI for USGS NHD data extraction ----------

## Grab huc 8s with nhdplus tools
CA_huc8 <- nhdplusTools::get_huc8(AOI = CA_sf) 

## Filter to Lake Tahoe's Huc area
LT_huc8 <- CA_huc8|> 
  filter(name == 'Lake Tahoe')

## Create huc4 string to allow data download for AOI (HR NHD data downloads only at Huc4 level)
LT_huc4 <- LT_huc8$huc8 |> substr(1,4) |> unique()

# NHD Data Download (conditional Run only if not yet downloaded) ----------

if(length(list.files('in/R/nhdPlusTools')) == 0){
  
  print(paste('Downloading NHD data for the following selected CA HUC4s', LT_huc4))

  download_nhdplushr('in/R/nhdPlusTools', LT_huc4)
}

## paths for geopackage
huc_gpkg_path <- 'in/R/nhdPlusTools/nhdplustools_HU_data.gpkg'
wtbdy_gpkg_path <- 'in/R/nhdPlusTools/nhdplustools_wtrbdy_data.gpkg'

if(length(list.files('in/R/nhdPlusTools', full.names = T, pattern = '_data.gpkg')) == 0){

  # Store HUC shps into own geopackage
  print(paste('... Storing HUCs shps into a local geopackage here:', huc_gpkg_path))
  nhdplusTools::get_nhdplushr(hr_dir = 'in/R/nhdPlusTools/',
                              out_gpkg = huc_gpkg_path,
                              layer = c('WBDHU6','WBDHU8', 'WBDHU10'))

  # Store Waterbody shps into own geopackage
  print(paste('... Storing Waterbody shps into a local geopackage here:', wtbdy_gpkg_path))
  nhdplusTools::get_nhdplushr(hr_dir = 'in/R/nhdPlusTools/',
                              out_gpkg = wtbdy_gpkg_path,
                              layer = 'NHDWaterbody')
  }
  
## Display gpkg layers
print(st_layers(huc_gpkg_path))
print(st_layers(wtbdy_gpkg_path))

# Load nhdPlus features for AOI ----------------------------------------------------------------
## HUCS
huc10 <- st_read(huc_gpkg_path, layer = 'WBDHU10')
huc8 <- st_read(huc_gpkg_path, layer = 'WBDHU8') 
huc6 <- st_read(huc_gpkg_path, layer = 'WBDHU6')

## Waterbodies
lakes <- st_read(wtbdy_gpkg_path, layer = 'NHDWaterbody') |> 
  st_make_valid()

lakes_in_hucs <- st_join(lakes, huc6) |>
  ## filtering out NA HUC6 because this ensures that we only have Lakes within the CA Huc6 boundaries
  filter(!is.na(HUC6),
  ## rm-ing NA AreaSqKM
  !is.na(AreaSqKM.x))

```

### B. Fetch geospatial data for Lake Tahoe Basin

_Note: code hidden in markdown output_

```{r LT_data_pull, include=FALSE, warning=FALSE, message=FALSE}

# NHDPLUS Tools data - Lake Tahoe subset ----------------------------------

# HUC
Lake_Tahoe_huc6 <- filter(huc6, Name == 'Truckee') 
Lake_Tahoe_huc8 <- filter(huc8, Name == 'Lake Tahoe')

# Water features
LT_lakes <- st_join(lakes_in_hucs , Lake_Tahoe_huc8) |>
  filter(!is.na(HUC8))

## Grabbing just LT
LT <- LT_lakes |> filter(GNIS_Name == "Lake Tahoe")

## Flowlines - this is pulled from NHDplus tools integrated API vs. from local geopackage
LT_flines <- nhdplusTools::get_nhdplus(AOI = Lake_Tahoe_huc8,
                         realization = 'flowline',
                         streamorder = 1)

# NWIS Gauges --------------------------------------------------------------
## for more information on USGS NWIS data, please visit: https://waterdata.usgs.gov/nwis and/or USGS dataRetrieval package documentation 

# ID Active NWIS Sites in LT Basin

## ID active sites
### choosing active sites as active over last last 90 days 
### selected daily nwis daily values (dv) 

prev90days <- Sys.Date() - 90

active_sites_data_2023 <- dataRetrieval::readNWISdata(huc = Lake_Tahoe_huc8$HUC8,
                                       service = "dv",
                                       parameterCd = "00060",
                                       startDate = prev90days) |> 
  ## Filtering out this site, determined ot be too far upstream (along a Southern tributary)
  filter(site_no != "103366092")

active_sites_2023 <- unique(active_sites_data_2023$site_no)

## Order sites now for viz after 
site_lst_ordered <- c("10336676",
                      "10336660",
                      "10336645",
                      "10336610",
                      "10336780",
                      "10336730",
                      "10336715",
                      "10336700",
                      "10336698")

active_sites_2023 <- factor(active_sites_2023, site_lst_ordered) |> sort()

## Final spatial data frame of active sites for Viz
active_nwis_sites_lake_tahoe <- get_nwis(AOI = Lake_Tahoe_huc8,
                                         t_srs = NULL,
                                         buffer = 0) |>
  filter(site_no %in% active_sites_2023) |> 
  mutate(site_no = factor(site_no, active_sites_2023)) |>
  arrange(site_no)

```

#### C. Obs. LT spatial data in mapview

_Note: code hidden in markdown output_

```{r looking_at_spatial_data_w_mapview, echo = FALSE,  warning=FALSE, message=FALSE}

# Quick interim Mapview of spatial object fetched -----------------------------------------------------------

mapview(LT_flines, color = 'darkblue')+
  mapview(LT_lakes, col.regions = 'darkblue')+
  mapview(active_nwis_sites_lake_tahoe, col.regions = 'red')+
  mapview(Lake_Tahoe_huc8, col.regions = NA, color = 'black', alpha.regions = 0)

```

## Fetch streamflow data for viz 

### A. Extract surface water gauge data for plots

_Note: code hidden in markdown output_

```{r pull_nwis_data, include= FALSE, warning= FALSE, message=FALSE}

# Pull all NWIS data for active sites ---------------------------------

## Get Data with readNWISdv 
LT_dv_data <- readNWISdv(siteNumbers = active_sites_2023,
           parameterCd = '00060') |> 
  renameNWISColumns()

```

## Process data

_Note: code hidden in markdown output_

```{r flow_data_processing, include= FALSE, warning= FALSE, message=FALSE}

# Dates --------------------------------------------------------------

LT_dv_data_dates <- LT_dv_data |>
  mutate(year = year(Date),
         month = month(Date, label = T),
         day = day(Date),
         day_of_week = lubridate::wday(Date, label = T),
         ## grabbing month_day only, as I hope this can be the X axis - however this is not considered Date format. 
         month_day = format(Date,"%m-%d"),
         ## making a fake date column  so that each year can be plotted as it's own line on the line plot 
         fake_date = as.Date(paste0('2020-',month_day))
         )

# Moving Avg ----------------------------------------------------------

# Calc Moving Average Values into a new column
LT_dv_data_w_MA <- LT_dv_data_dates |>
  ## calc MA with zoo::rollmean() . moving average period (k) chosen as 5 
  mutate(MA = rollmean(x = Flow, 
                       k = 5,
                       na.pad = T,
                       align = 'center')
         )

# Focus on Spring months -------------------------------------------------

LT_dv_data_viz <- LT_dv_data_w_MA |> 
  filter(month %in% c('Mar','Apr','May'),
         fake_date >= '2020-03-01',
         fake_date <= '2020-05-15'
         ) |>
  ## reorder table to factor in active_sites_2023 
  mutate(site_no = factor(site_no, active_sites_2023)) |> 
  arrange(site_no) |> 
  ## group by to apply group-map
  group_by(site_no) 

```

## Vizualize

### A. Map

```{r LT_map, warning=FALSE, message=FALSE}

# Map Viz Variables -----------------------------------------------------------

## bbox
extent_bbox <- st_bbox(Lake_Tahoe_huc8)
extent_lt_huc8 <- st_as_sfc(extent_bbox)

map_bbox <- st_bbox(Lake_Tahoe_huc8)

## colors

lt_all_colors <- c("#000000", "#0046A5", "#005DBA", "#0082CD", "#001E54","#00427C",
                "#FFFEFF", "#3DA6DD", "#574435", "#C7BBAD", "#4492DA", "#0079BE",
                "#005E94", "#95C2E1", "#B0CFE3", "#171310", "#354658", "#97D4AB",
                "#6B5E56", "#647100", "#A8D0F4", "#007FD0", "#63AEEF", "#00599F",
                "#00B79F", "#00B3A0", "#00506B", "#001920", "#2EE5C6", "#889756"
                )

teal_1 <- "#00B79F"
selected_teal <- '#008F81'
selected_grey <- "#E3E3E3"

```

```{r extent_map}

# Extent map -----------------------------------------------------------

(
  extent_map <- ggplot() + 
    geom_sf(data = CA_sf,
            fill = "white",
            color = alpha('#6D6D6D', 0.5)
            ) + 
    geom_sf(data = extent_lt_huc8,
            color = 'forest green',
            fill = NA
            )+
    geom_sf(data = LT,
            fill = "#A8D0F4",
            color = "#A8D0F4",
            alpha = 0.8
            )+
   coord_sf(
     ylim = c(extent_bbox$ymin - 6,
              ## just +4 otherwise it's too upper distance from centroid to top is too large  
              extent_bbox$ymax + 4
              ),
     xlim = c(extent_bbox$xmin - 6,
              extent_bbox$xmax + 6
              )
     )+
  theme_void()
)

```

```{r main_map}

# Main map -----------------------------------------------------------

(
  lt_map <- ggplot()+
   geom_sf(data = CA_sf,
           fill = 'white',
           color = alpha('#6D6D6D', 0.5),
           size = 0.5
           )+
   geom_sf(data = LT,
           fill = "#A8D0F4", 
           color = "#A8D0F4",
           alpha = 1
           )+
  ## Basin flowlines
   geom_sf(data = LT_flines,
           color = "#A8D0F4",
           size  = 0.5,
           alpha = 0.7
           )+
  ## Selected nwis st stations around LT
   geom_sf(data = active_nwis_sites_lake_tahoe |>
             ## removing sites that are too close together - added below
             filter(!site_no %in% c('10336698','10336700')),
           color = "#97D4AB",
           fill = selected_teal,
           size = 3, 
           shape = 21
           )+
   geom_sf(data = active_nwis_sites_lake_tahoe |>
             filter(site_no %in% c('10336698', '10336700')) |> 
             ## slight jitter to ensure not completely overlapping
             st_jitter(amount = 0.005),
           color = "#00B79F", 
           fill = selected_teal,
           size = 3,
           shape = 21
           )+
   coord_sf(
     ylim = c(map_bbox$ymin,
              map_bbox$ymax),
     xlim = c(map_bbox$xmin,
              map_bbox$xmax)
     )+
   theme_classic()+
  ## adding state labels at top of map
  annotate('text',
           x = -120.10, y = 39.34, 
           label = c('California'),
           hjust = 'left', size = 3.5,
           colour = '#6D6D6D'
           )+ 
  annotate('text',
           x = -119.975, y = 39.34,
           label = c('Nevada'),
           hjust = 'left', size = 3.5,
           colour = '#6D6D6D'
           )
)

```

```{r segments}

(
  lt_map_segs <- lt_map + 
    # top left
    geom_segment(aes(x = active_nwis_sites_lake_tahoe$geometry[[1]][1] - 0.015, xend = active_nwis_sites_lake_tahoe$geometry[[1]][1] - 0.1, 
                     y = active_nwis_sites_lake_tahoe$geometry[[1]][2] + 0.005, yend = active_nwis_sites_lake_tahoe$geometry[[1]][2] + 0.06),
                 arrow = arrow(length = unit(0.03, "npc")),
                 arrow.fill = "#354658", alpha = 0.4)+
    # 2nd to top left
      geom_segment(aes(x = active_nwis_sites_lake_tahoe$geometry[[2]][1] - 0.015, xend = active_nwis_sites_lake_tahoe$geometry[[2]][1] - 0.1, 
                     y = active_nwis_sites_lake_tahoe$geometry[[2]][2], yend = active_nwis_sites_lake_tahoe$geometry[[2]][2] - 0.03), 
                 arrow = arrow(length = unit(0.03, "npc")),
                 arrow.fill = "#354658", alpha = 0.4)+
    # 2nd to bottom left 
      geom_segment(aes(x = active_nwis_sites_lake_tahoe$geometry[[3]][1] - 0.015, xend = active_nwis_sites_lake_tahoe$geometry[[3]][1] - 0.15, 
                     y = active_nwis_sites_lake_tahoe$geometry[[3]][2] - 0.003, yend = active_nwis_sites_lake_tahoe$geometry[[3]][2] - 0.08), 
                 arrow = arrow(length = unit(0.03, "npc")),
                 arrow.fill = "#354658", alpha = 0.4)+
    # bottom left
      geom_segment(aes(x = active_nwis_sites_lake_tahoe$geometry[[4]][1]-0.015, xend = active_nwis_sites_lake_tahoe$geometry[[4]][1] - 0.13, 
                     y = active_nwis_sites_lake_tahoe$geometry[[4]][2]- 0.006, yend = active_nwis_sites_lake_tahoe$geometry[[4]][2] - 0.07 ), 
                 arrow = arrow(length = unit(0.03, "npc")),
                 arrow.fill = "#354658", alpha = 0.4)+
    # bottom right
     geom_segment(aes(x = active_nwis_sites_lake_tahoe$geometry[[5]][1] + 0.015, xend = active_nwis_sites_lake_tahoe$geometry[[5]][1] + 0.11, 
                     y = active_nwis_sites_lake_tahoe$geometry[[5]][2] - 0.01, yend = active_nwis_sites_lake_tahoe$geometry[[5]][2] - 0.1), 
                 arrow = arrow(length = unit(0.03, "npc")),
                 arrow.fill = "#354658", alpha = 0.4)+
    # 2nd to bottom on right
      geom_segment(aes(x = active_nwis_sites_lake_tahoe$geometry[[6]][1] + 0.015, xend = active_nwis_sites_lake_tahoe$geometry[[6]][1] + 0.07, 
                     y = active_nwis_sites_lake_tahoe$geometry[[6]][2] - 0.01, yend = active_nwis_sites_lake_tahoe$geometry[[6]][2] - 0.08), 
                 arrow = arrow(length = unit(0.03, "npc")),
                 arrow.fill = "#354658", alpha = 0.4)+
    # 3rd to top on right
      geom_segment(aes(x = active_nwis_sites_lake_tahoe$geometry[[7]][1] + 0.015, xend = active_nwis_sites_lake_tahoe$geometry[[7]][1] + 0.04, 
                     y = active_nwis_sites_lake_tahoe$geometry[[7]][2] - 0.01, yend = active_nwis_sites_lake_tahoe$geometry[[7]][2] - 0.04), 
                 arrow = arrow(length = unit(0.03, "npc")),
                 arrow.fill = "#354658", alpha = 0.4)+
    # top-right-right
      geom_segment(aes(x = active_nwis_sites_lake_tahoe$geometry[[8]][1] + 0.015, xend = active_nwis_sites_lake_tahoe$geometry[[8]][1] + 0.085, 
                     y = active_nwis_sites_lake_tahoe$geometry[[8]][2] - 0.005, yend = active_nwis_sites_lake_tahoe$geometry[[8]][2] - 0.02), 
                 arrow = arrow(length = unit(0.03, "npc")),
                 arrow.fill = "#354658", alpha = 0.4)+
    ## top-right
      geom_segment(aes(x = active_nwis_sites_lake_tahoe$geometry[[9]][1] + 0.01, xend = active_nwis_sites_lake_tahoe$geometry[[9]][1] + 0.08, 
                       y = active_nwis_sites_lake_tahoe$geometry[[9]][2] + 0.01, yend = active_nwis_sites_lake_tahoe$geometry[[9]][2] + 0.08), 
                   arrow = arrow(length = unit(0.03, "npc")),
                   arrow.fill = "#354658", alpha = 0.4)
)

```

```{r final_formatting_map}
  
## Using generic save_map function to generate final map (in 'src/funs.R')
(
  lt_map_for_viz <- final_map_formatting(lt_map_segs,
                                         scale_arrow_color = "#00427C",
                                         out_file = 'out/LT_map_gauges.png')
)
```

### B. Streamflow plots

```{r flow_data_line_plot, warning=FALSE, message=FALSE, results = 'hide'}

## Generate line plots
## Note - function pulled from 'src/funs.R'
line_plot_lst <- LT_dv_data_viz |>
  group_map( ~time_series_plot(data = .x,
                               x_var = 'fake_date', ## note from above - created a fake date with same year to get each year as different line
                               y_var = 'MA',
                               date_range =  c('1975-01-01', '2023-05-15'),
                               label_site = '10336676',
                               string_format_title = 'USGS Station ID: %s',
                               top_year_num = 4,
                               out_folder = 'out',
                               color_palette_highlight = selected_teal,
                               color_palette_unhighlight = selected_grey
                               ),
             .keep = TRUE) ## note - setting keep = TRUE ensure group col is kept.

line_plot_grid_1 <- cowplot::plot_grid(plotlist = line_plot_lst[1:5])
line_plot_grid_2 <-cowplot::plot_grid(plotlist = line_plot_lst[6:10])

ggsave(plot = line_plot_grid_1, filename = 'out/line_plot_grid_1.png',
       width = 9, height = 9, dpi = 300)
ggsave(plot = line_plot_grid_2, filename = 'out/line_plot_grid_2.png',
       width = 9, height = 9, dpi = 300)

```

**Final Viz Pieces:**

1. lt_map_for_viz
2. line_plot_lst

### C. Viz formatting

```{r plotting_vars}

# Custom fonts, custom settings
font_legend <- "Vollkorn"
sysfonts::font_add_google("Vollkorn")
supporting_font <- "Source Sans Pro"
sysfonts::font_add_google("Source Sans Pro")
showtext::showtext_opts(dpi = 300, regular.wt = 200, bold.wt = 700)
showtext::showtext_auto(enable = TRUE)

# Colors
background_color = "#FFFEFF"
font_color_1 = "#0046A5"
font_color_2 = "#63AEEF" 

# The background canvas 
canvas <- grid::rectGrob(
  x = 0,
  y = 0, 
  width = 16,
  height = 9,
  gp = grid::gpar(fill = background_color,
                  alpha = 1,
                  col = background_color
                  )
)

# Loading USGS logo (black selected)
usgs_logo <- magick::image_read("../usgs_logo_black.png") 

# Fixed X and Y position to sample from if needed
## taken from @aarcher's Day4 viz : https://code.usgs.gov/wma/vizlab/chart-challenge-23/-/blob/main/04_historical_aaarcher/04_historical_aaarcher_script.Rmd
Ymargin <- 0.04
Xmargin <- 0.04
Xbase <- 1-Xmargin # from right
Ybase <- Ymargin*2 # from bottom
Yjitter <- (0.40 - 2*Ymargin)/2
Xjitter <- (1 - 2*Xmargin)/5 # total jitter
Xwidth <- 0.15

```

```{r cowplotting_line_plots_arrangement}

# Line plot arrangement --------------------------------------------------------
## each plot placement individually due to to unique placement  and aligment with arrows on map 

draw_line_plots <- purrr::map(1:length(line_plot_lst), function(x){ ## Note - purrr must be placed in front of map or loop won't work (`coerce of type `closure` error`)
  
                          if(x == 1){
                          X_position <- 0.4
                          Y_position <- 0.65
                        } else if(x == 2){
                          X_position <- 0.36
                          Y_position <- 0.48  
                        } else if(x == 3){
                          X_position <- 0.4
                          Y_position <- 0.30
                        } else if(x == 4){
                          X_position <- 0.49
                          Y_position <- 0.11  
                        } else if(x == 5){
                          X_position <- 0.80
                          Y_position <- 0.1  
                        } else if(x == 6){
                          X_position <- 0.84
                          Y_position <- 0.27   
                        } else if(x == 7){
                          X_position <- 0.84
                          Y_position <- 0.45
                        } else if(x == 8){
                          X_position <- 0.84
                          Y_position <- 0.62
                        } else if(x == 9){
                          X_position <- 0.8
                          Y_position <- 0.78
                        } else { 
                          X_position <- 0.35
                          Y_position <- 0.08
                          }
  
                       draw_plot(line_plot_lst[[x]],
                                 x = X_position,
                                 y = Y_position,
                                 hjust = 1,
                                 height = Yjitter,
                                 width = Xjitter
                                 ) 

})

```

```{r final_plot_generation_and_save, fig.width = 16, fig.height = 9, warning=FALSE, message=FALSE}

ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 9, width = 16,
            hjust = 0, vjust = 1) +
  # the main plot
  draw_plot(extent_map,
            x = 0.03,
            y = 0.15,
            height = 0.4,
            width = 0.2
            )+
  draw_plot(lt_map_for_viz,
            x = 0.01,
            y = 0.01,
            height = 1,
            width = 1,
            scale = 0.8) +
  draw_line_plots +
    # Title 1
  draw_label("The 'Big Melt' has begun.",
             x = Xmargin,
             y = 0.96,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = supporting_font,
             fontface = 'bold',
             color = font_color_1,
             size = 36) +
    # Title 2 2023 spring flows into Lake Tahoe compared to the historical record
  draw_label("2023 spring flows",
             x = Xmargin,
             y = 0.89,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = supporting_font,
             color = selected_teal,
             fontface = 'bold',
             size = 24) +
  draw_label("into Lake Tahoe",
             x = 0.207,
             y = 0.89,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = supporting_font,
             color = '#6D6D6D',
             size = 24) +
  draw_label("compared to the historical record.",
             x = Xmargin,
             y = 0.85,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = supporting_font,
             color = '#6D6D6D',
             size = 24) +
  # Some explainer text
  draw_label("Margaux Sleckman, USGS\nFlow data from the USGS National Water Information System: https://waterdata.usgs.gov/nwis/rt",
             fontfamily = supporting_font,
             x = Xbase,   
             y = 0.05,
             size = 14,
             hjust = 1,
             vjust = 0,
             color = font_color_1,
             fontface = 'italic'
             )+
  # Add logo
  draw_image(usgs_logo, 
             x = Xmargin,
             y = 0.05,
             width = 0.1, 
             hjust = 0, vjust = 0, 
             halign = 0, valign = 0)

# Save the final image in Twitter's 16 by 9 format
ggsave(filename = "out/20230428_trends_msleckman.png", 
       width = 16,
       height = 9,
       dpi = 300)

```

## Supporting information

The Big Melt has begun; 2023 spring flows into Lake Tahoe compared to the historical record. Nine timeseries plots show daily streamflow (cubic feet / second) from March 2023 to present, highlighted in green, compared to historical record, shown in grey that date back to 1975.  Streamgraphs show the 2023 snowmelt runoff in the basin trending upwards across all stream gauges, aiming to surpass previous years dating back to 1975. 

### Key takeaways of this viz (1-2 sentences each)

Following an unprecedented Winter, the 2023 snowpack in California’s Sierra Nevada mountain range is generating significant runoff through it’s rivers and lakes. This data visualization shows daily streamflow measurements at USGS stream gauges surrounding Lake Tahoe for the months of March to the start of May in 2023, as well as all historical years dating back to 1975 (in grey). 

* While it is still early in the Spring season, the 2023 snowmelt runoff in the Lake Tahoe basin is trending upwards across all stream gauges, aiming to surpass previous years dating back to 1975.  

* All gauges are found in tributaries that are uniquely influenced by environmental factors that drive snow melt such as temperature or topography. This explains the variability in flow across each gauges. 

### Data source(s)

NHDPlus High Resolution - geospatial dataset capturing water basins and flow of water across the Nation's landscapes. https://www.usgs.gov/national-hydrography/nhdplus-high-resolution

USGS National Water Information System - water data collected at water gauges.
https://waterdata.usgs.gov/nwis/rt


