---
title: "28 trends"
author: "Margaux Sleckman"
date: "2023-03-10"
output: html_document
---

## Instructions for using this template

> Make sure you have copied the template subdirectory and named it with your prompt's date, prompt's name and your name (e.g., `/01_part-to-whole_cnell`) before editing this document!

1. Put all comments either in the html text (like this) or in the `r` chunks (delineated with 3 "`" marks) to explain your process to Vizlab reviewers
2. Put all R code in the chunks. They're organized for you to (in order): do some set-up steps (load libraries), load in any data (manually or with built in packages), process your data for plotting, create plot using ggplot, and finally compose the twitter image with cowplot.
3. You can run individual code chunks while developing code and plots. Note: the plots that are created in the RStudio coding window *do not* match the plot as created through the ggsave function. 
4. To preview the final plot, use the `Knit` button above. That will create a preview of the final plot in the output `html` and will save the `png` in the `out/` folder. On many systems, the preview in the script document and the html will not match the output `png`, so please refer to the `png` for final proofing.
5. When you're happy with the final plot, fill out the supporting information at the bottom and push to gitlab for review. Note: only commit input data if it can't be downloaded directly with code (e.g., `sbtools` or `dataRetrieval`).

This website has great information on how to use RMarkdown: https://yihui.org/knitr/

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```



Define libraries here. 

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) # includes ggplot
library(readr) # best for reading and writing csvs
library(lubridate)
library(glue)

library(sf)
library(mapdata)
library(maps)
library(spData)

library(nhdplusTools)
library(dataRetrieval)

# These are used for the layout of the viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo
library(mapview) # for interactive map viz

```

```{r options}

options(tidyverse.quiet = TRUE, timeout = 500)
mapviewOptions(fgb = FALSE)

dir.create('in/R/nhdPlusTools', showWarnings = F)

```


## Load files

Save any files that you are using directly in the `in` folder. Then read them in to the environment with this chunk of code. Remember, use the RStudio project and relative file pathways, and never include any personal computer file pathways.

This is also a good place to load any data from other built-in packages such as `dataRetrieval` or `spData`, etc.

```{r load}

source('src/funs.R')

```

#### 1. Fetching nhdplus spatial data w/ nhdplustools

```{r 1_generic_nhdPlus_data_fetch}

# state boundaries --------------------------------------------------------

CA_sf <- us_states |> filter(NAME == 'California')
West_sf <- us_states |> filter(NAME %in% c('California','Oregon',
                                           'Nevada','Utah',
                                           'Arizona',
                                           'Idaho','Washington'))

# AOI for NHD data extraction ---------------------------------------------

## grab huc 8s with nhdplus tools
CA_huc8 <- nhdplusTools::get_huc8(AOI = CA_sf)

## define all huc4 to allow data download
CA_huc4 <- CA_huc8$huc8 |> substr(1,4) |> unique()

# NHD Data Download (conditional Run only if not downloaded yet) ----------------------------------

if(is.null(list.files('in/R/nhdPlusTools'))){
  
  print(paste('Downloading NHD data for the following CA HUC4s',CA_huc4))

  download_nhdplushr('in/', CA_huc4)
}

huc_gpkg_path <- 'in/R/nhdPlusTools/nhdplustools_HU_CA_data.gpkg'
wtbdy_gpkg_path <- 'in/R/nhdPlusTools/nhdplustools_CA_wtrbdy_data.gpkg'

if(is.null(list.files('in/R/nhdPlusTools', full.names = T, pattern = '_data.gpkg'))){
  # Store HUCs shps into geopackage
  print(paste('... Storing HUCs shps into a local geopackage here:', huc_gpkg_path))
  nhdplusTools::get_nhdplushr(hr_dir = 'in/R/nhdPlusTools/',
                              out_gpkg = huc_gpkg_path,
                              layer = c('WBDHU6','WBDHU8', 'WBDHU10'))

  # Store Waterbodies shps into geopackage
  print(paste('... Storing Waterbody shps into a local geopackage here:', wtbdy_gpkg_path))
  nhdplusTools::get_nhdplushr(hr_dir = 'in/R/nhdPlusTools/',
                              out_gpkg = 'in/R/nhdplustools_CA_wtrbdy_data.gpkg',
                              layer = 'NHDWaterbody')

  # # Store Waterbodies shps into geopackage
  # # This one takes A-WHILE so defaulted to using nhdplustools::get_flines()
  # nhdplusTools::get_nhdplushr(hr_dir = 'in/R/nhdPlusTools/',
  #                            out_gpkg = 'in/R/nhdplustools_CA_flines_data.gpkg',
  #                            layer = 'NHDFlowline')

  }
  
## Display Layers
print(st_layers(huc_gpkg_path))
print(st_layers(wtbdy_gpkg_path))

# Load nhdPlus features for AOI ----------------------------------------------------------------
## HUCS
huc10_CA <- st_read(huc_gpkg_path, layer = 'WBDHU10')
huc8_CA <- st_read(huc_gpkg_path, layer = 'WBDHU8') 
huc6_CA <- st_read(huc_gpkg_path, layer = 'WBDHU6')

## Waterbodies
lakes_CA <- st_read(wtbdy_gpkg_path, layer = 'NHDWaterbody') |> 
  st_make_valid()

lakes_in_CA_hucs <- st_join(lakes_CA, huc6_CA) |>
  ## filtering out NA HUC6 because this ensures that we only have Lakes within the CA Huc6 boundaries
  filter(!is.na(HUC6),
  ## rm-ing NA AreaSqKM
  !is.na(AreaSqKM.x))

```

#### 2. Fetch geospatial data for Lake Tahoe Basin

```{r 2_LT_data_pull}

# NHDPLUS Tools data - Lake Tahoe subset ----------------------------------

# HUC
Lake_Tahoe_huc6 <- filter(huc6_CA, Name == 'Truckee') 
Lake_Tahoe_huc8 <- filter(huc8_CA, Name == 'Lake Tahoe')

# Water features
LT_lakes <- st_join(lakes_in_CA_hucs , Lake_Tahoe_huc8) |> filter(!is.na(HUC8))

LT_flines <- get_nhdplus(AOI = Lake_Tahoe_huc8,
                         realization = 'flowline',
                         streamorder = 1)

# NWIS Gages --------------------------------------------------------------

# ID Active NWIS Sites in LT Basin

## ID active sites from last 90 days (dv)
prev90days <- Sys.Date() - 90
active_sites_data_2023 <- readNWISdata(huc = Lake_Tahoe_huc8$HUC8,
                                       service="iv",parameterCd="00060",
                                       startDate=prev90days)

## Vector of currently active sites
active_sites_2023 <- active_sites_data_2023$site_no |> unique()

active_nwis_sites_lake_tahoe <- get_nwis(AOI = Lake_Tahoe_huc8, t_srs = NULL, buffer = 0) |>
  filter(site_no %in% active_sites_2023)

# Pull all NWIS data for active sites

view_data_summary <- readNWISstat(siteNumbers = active_sites_2023,
                                  parameterCd="00060",
                                  statReportType="annual",
                                  statType="mean") |>
  group_by(site_no,year_nu) |>
  summarize(n()) |>
  arrange(year_nu)

dailyDataAvailable <- whatNWISdata(
  siteNumber = active_sites_2023,
  service = "dv",
  statCd = "00003"
)

LT_dv_data <- readNWISdv(siteNumbers = active_sites_2023,
           parameterCd = '00060') |> 
  renameNWISColumns()

# readNWISdata(site_no = active_sites_2023, service = 'iv')

```


## Get data ready for plotting
  
This next section is for doing any pre-processing steps, data joins, etc, to get your data ready for plotting with ggplot. Remember that most of the time, ggplot prefers data in a "wide" format.

### 3. Process Gauge Data for plots
```{r data_processing}

# Dates --------------------------------------------------------------

LT_dv_data_dates <- LT_dv_data |>
  mutate(year = year(Date),
         month = month(Date, label = T),
         day = day(Date),
         day_of_week = lubridate::wday(Date, label = T),
         ## grabing month_day only, as I hope this can be the X axis - however this is not considered Date format. 
         month_day = format(Date,"%m-%d")
         ) |> 
  # placing month day at end
  select(!month_day,month_day)

# Moving Avg --------------------------------------------------------------

# Calc Moving average Values
LT_dv_data_w_MA <- LT_dv_data_dates |>
  ## calc MA with zoo::rollmean() . Check what k (2nd param) is for 
  mutate(MA = rollmean(Flow, 2, na.pad = T, align = 'right'))

```

```{r }

# Quick Mapview -----------------------------------------------------------

## VIZ
mapview(LT_flines_3, color = 'darkblue')+
  mapview(LT_lakes, col.regions = 'darkblue')+
  mapview(active_nwis_sites_lake_tahoe, col.regions = 'red')+
  mapview(Lake_Tahoe_huc8, col.regions = NA, color = 'black', alpha.regions = 0)



```

## Set up main plot

This chunk is where the main ggplot grob definition and set-up occurs. You may have several ggplot grobs, depending on your approach. For each one, if you define it with the `<-` assignment operator and surround the whole code statement in parentheses, you'll be able to preview here what the plot looks like before the next cowplot step.

> Note: The hardest part of this process is getting text to be the optimum size in the final `png` for Twitter. The font sizes here will **not** be the same as in the final image after the cowplot step. Make sure to check the output `png` for true font sizing and plot composition. 

```{r plotting_vars}
# Load some custom fonts and set some custom settings
font_legend <- "Pirata One"
sysfonts::font_add_google("Pirata One")
supporting_font <- "Source Sans Pro"
sysfonts::font_add_google("Source Sans Pro")
showtext::showtext_opts(dpi = 300, regular.wt = 200, bold.wt = 700)
showtext::showtext_auto(enable = TRUE)

# Define colors
background_color = "#0A1927"
font_color = "#ffffff"

# The background canvas for your viz
canvas <- grid::rectGrob(
  x = 0, y = 0, 
  width = 16, height = 9,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color)
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_white.png") 


```

```{r }

# Viz Variables -----------------------------------------------------------

map_bbox <- st_bbox(Lake_Tahoe_huc8)

```

```{r }


# Main plot
(main_map <- ggplot()+
   geom_sf(data = West_sf, fill = 'white', color = 'grey', alpha = 0.1, size = 0.5)+
   geom_sf(data = LT_lakes, fill = '#3792cb', color = '#3792cb', alpha = 0.8)+
   geom_sf_text(data = West_sf, color = 'grey', aes(label = NAME),
                size =2, fontface = 'italic', nudge_x = 0.1) +
   geom_sf(data = Lake_Tahoe_huc8, fill = 'transparent', color = 'firebrick', size = 0.8, linetype = 'dotted')+
   geom_sf(data = LT_flines_3, color = '#3792cb', size  = 0.5, alpha = 0.6)+
   geom_sf(data = active_nwis_sites_lake_tahoe,
           aes(geometry = geometry),
           color = '#006200',
           size = 1.3, alpha = 0.9)+
   geom_sf_text(data = active_nwis_sites_lake_tahoe, color = 'black',
                aes(label = site_no), size =3, nudge_x = 0.05) +
   coord_sf(ylim = c(map_bbox$ymin, map_bbox$ymax),
            xlim = c(map_bbox$xmin, map_bbox$xmax))+
   theme_classic()+
   theme(plot.title = element_text(size = 10, face= 'bold'),
         legend.text = element_text (size = 10),
         legend.title = element_text (size = 12),
         axis.title.x = element_blank(),
         axis.line = element_blank(),
         panel.background = element_rect(color= 'black'),
         axis.title.y = element_blank(),
         legend.position = 'bottom',
         axis.text = element_blank(),
         axis.ticks = element_blank()
         )

# ggsave('out/LT_map_gauges.png', width = 9, height = 9, dpi = 300)

 
 
 
 
 
 
 )



```


## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

**Make sure to use the format for saving your png with the date of the prompt that you've been assigned to!** (e.g., `20230401_part-to-whole_cnell.png`)

```{r cowplot, fig.width = 16, fig.height = 9}
ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 9, width = 16,
            hjust = 0, vjust = 1) +
  # the main plot
  draw_plot(main_plot,
            x = 0.01,
            y = 0.01,
            height = 1) +
  # explainer text
  draw_label("Explainer text",
             fontfamily = supporting_font,
             x = 0.96,   
             y = 0.05,
             size = 14,
             hjust = 1,
             vjust = 0,
             color = font_color)+
  # Title
  draw_label("Title",
             x = 0.04,
             y = 0.285,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = font_color,
             size = 55) +
  # Add logo
  draw_image(usgs_logo, 
             x = 0.04,
             y = 0.05,
             width = 0.1, 
             hjust = 0, vjust = 0, 
             halign = 0, valign = 0)

# Save the final image in Twitter's 16 by 9 format
# !! Use format for saving with the date of your prompt: 
#         YYYYMMDD_prompt_name ()
# e.g. `20230101_part-to-whole-cnell.png`
ggsave(filename = "out/20230428_trends_msleckman.png", 
       width = 16, height = 9, dpi = 300)
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. Key takeaways here

### Data source(s)

Data sources here.

