---
title: "April 7 - Hazards - Flooding animation of Norristown, Pennsylvania during hurricane Ida"
author: "Matthew Conlon and James Colgin"
date: "2023-03-29"
output: html_document
---

## Instructions for using this template

> Make sure you have copied the template subdirectory and named it with your prompt's date, prompt's name and your name (e.g., `/01_part-to-whole_cnell`) before editing this document!

1. Put all comments either in the html text (like this) or in the `r` chunks (delineated with 3 "`" marks) to explain your process to Vizlab reviewers
2. Put all R code in the chunks. They're organized for you to (in order): do some set-up steps (load libraries), load in any data (manually or with built in packages), process your data for plotting, create plot using ggplot, and finally compose the twitter image with cowplot.
3. You can run individual code chunks while developing code and plots. Note: the plots that are created in the RStudio coding window *do not* match the plot as created through the ggsave function. 
4. To preview the final plot, use the `Knit` button above. That will create a preview of the final plot in the output `html` and will save the `png` in the `out/` folder. On many systems, the preview in the script document and the html will not match the output `png`, so please refer to the `png` for final proofing.
5. When you're happy with the final plot, fill out the supporting information at the bottom and push to gitlab for review. Note: only commit input data if it can't be downloaded directly with code (e.g., `sbtools` or `dataRetrieval`).

This website has great information on how to use RMarkdown: https://yihui.org/knitr/

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```



Define libraries here. 

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries

# rayshader/render/image all used for final visualization
library(rayshader)
library(rayrender)
library(rayimage)

library(raster) # for elevation files
library(leaflet) # plot area to check extent
library(magick) # for saving GIF
library(gifski) # for saving GIF
library(webshot2)
library(osmdata) # imagery overlay 
library(sf) # spatial processing
library(ggplot2) # plotting
library(dataRetrieval) # streamflow data from event
library(readr)
library(dplyr) # organize data
library(imager) # for organizing image

library(rgl) # added 
```



## Load files

Save any files that you are using directly in the `in` folder. Then read them in to the environment with this chunk of code. Remember, use the RStudio project and relative file pathways, and never include any personal computer file pathways.

This is also a good place to load any data from other built-in packages such as `dataRetrieval` or `spData`, etc.


```{r load}
source("in/rayshade_functions.R") # This contains functions for pulling, organizing, and visualizing data. From Bishop, 2019 (https://wcmbishop.github.io/rayshader-demo/)

```

## Get data ready for plotting
  
This next section is for doing any pre-processing steps, data joins, etc, to get your data ready for plotting with ggplot. Remember that most of the time, ggplot prefers data in a "wide" format.


```{r processing}
##### Stream information #####
# First, gather information from streamgage during event

start_str <- "2021-09-01 12:00:00" # Ida - prior to event
end_str <- "2021-09-03 22:00:00"   # Ida - conclusion of event
river_data <- readNWISuv(siteNumbers = "01473500", parameterCd = c("00060","00065"),
                         startDate = as.Date(start_str), endDate = as.Date(end_str)) %>%
  renameNWISColumns() # site number controls gathering of DEM and overlay data

# Get site info
site_info <- attr(river_data, "siteInfo")
site_lat <- site_info$dec_lat_va
site_lon <- site_info$dec_lon_va

# dplyr commands after extracting siteinfo since resulting dfs drop dataRetrieval attributes
river_data <- river_data  %>% 
  filter(dateTime >= as.POSIXct(start_str), dateTime <= as.POSIXct(end_str)) %>% 
  mutate(minute = as.numeric(format(dateTime, "%M"))) %>% filter(minute %in% c(0, 30)) %>% 
  mutate(waterdepth = (GH_Inst * 1) + 7) # number is in feet <- need better solution

# Plot a quick hydrograph of this event
plot(river_data$dateTime, river_data$GH_Inst, xlab="Time", ylab="Water Level, ft")

##### Identify plotting area #####

# Bounding box of location - this will change with each site
bbox <- list(
  p1 = list(long = site_lon - 0.015, lat = site_lat - 0.015), # right and bottom
  p2 = list(long = site_lon + 0.015, lat = site_lat + 0.015)) # left and top

# Scale site location to image dimensions
site_y_img <- convert_to_img_dim(site_lat, bbox$p2$lat, bbox$p1$lat)
site_x_img <- convert_to_img_dim(site_lon, bbox$p2$long, bbox$p1$long)

# Leaflet to see area selection
leaflet() %>%
  addTiles() %>% 
  addRectangles(
    lng1 = bbox$p1$long, lat1 = bbox$p1$lat,
    lng2 = bbox$p2$long, lat2 = bbox$p2$lat,
    fillColor = "transparent") %>%
  fitBounds(
    lng1 = bbox$p1$long, lat1 = bbox$p1$lat,
    lng2 = bbox$p2$long, lat2 = bbox$p2$lat)

# Define the image size
image_size <- define_image_size(bbox, major_dim = 400)


# Download elevation data
elev_file <- file.path("in/elevation_files/norristown.tif")
#get_usgs_elevation_data(bbox, size = "400, 400", file = elev_file,
#                        sr_bbox = 4326, sr_image = 4326)


# Load elevation data
elev_img <- raster::raster(elev_file)
elev_matrix <- matrix(raster::extract(elev_img, raster::extent(elev_img), 
                                      buffer = 1000), nrow = ncol(elev_img), 
                      ncol = nrow(elev_img))

# Exaggerate z axis
elev_matrix <- elev_matrix * 3 # Arbitrary for now
zscale = 10

# Define 3 points on the plane that you want to flatten.
# Format is (x, y, z) where x is grid row, y is grid column, and z is elevation.
pt.up <- c(180, 1, 90)#, elev_matrix[1,40])          # 435
pt.gage <- c(200, 200, 90)#, elev_matrix[200,200])  # 418
pt.down <- c(220, 400, 90)#, elev_matrix[400,210])  # 398

# Set up the 3 points in matrix formation ax + by + cz = d
plane.points <- matrix(c(pt.up,
                         pt.gage,
                         pt.down), 
                       nrow = 3, ncol = 3, byrow=TRUE)

# Solve for the equation of a plane through the 3 points. 
# The output is a, b, c, and d = 1.
plane.coeffs <- solve(plane.points, c(1, 1, 1))

a <- plane.coeffs[1]
b <- plane.coeffs[2]
c <- plane.coeffs[3]
d <- 1

# Build a plane going through the 3 points. Format is the same as the original 
# elevation matrix with x, y, z grid row, column, elevation. We know the x and 
# the y (the grid coordinates), and d = 1. We are calculating a new z, the 
# elevation on the tilted plane, for all x, y points. Format is ax + by + cz = d
# Column 1 calculation: (d - a*rep(1:400) - b*1) / c
elev_adjust <- matrix(data = matrix(rep(1:400,400)), nrow = 400, ncol = 400)
for (i in 1:400) {
  elev_adjust[ , i] <- (d - a*rep(1:400) - b*i) / c
}

# Calculate the distance between the tilted plane and a flat plane at the gage 
# elevation, for each grid cell (x, y).
elev_adjust <- elev_adjust - elev_matrix[200,200]

# Subtract the distance from the real elevations. This tilts the real topology
# to a flat plane going through the 3 points.
elev.new_matrix <- elev_matrix - elev_adjust

elev_matrix <- elev.new_matrix

# Cleanup
rm(a,b,c,d,i,elev.new_matrix,elev_adjust,plane.coeffs,plane.points,pt.up,pt.gage,pt.down,start_str,end_str)

# Calculate rayshader layers
ambmat <- ambient_shade(elev_matrix, zscale = zscale)
raymat <- ray_shade(elev_matrix, zscale = zscale, lambert = TRUE)

# label location - should be (200,200)
label <- list(text = "USGS 01473500")
label$pos <- find_image_coordinates(long = site_lon, lat = site_lat, bbox = bbox,
                                    image_width = image_size$width, image_height = image_size$height)

# Download overlay image
overlay_file <- "in/imagery_files/norristown_map.png"
#get_arcgis_map_image(bbox, map_type = "World_Imagery", file = overlay_file,
#                     width = "400", height = "400", sr_bbox = 4326)

overlay_img <- png::readPNG(overlay_file)

# Open Street Map data - stream lines 
lat_range <- c(site_lat - 0.015, site_lat + 0.015)
long_range <- c(site_lon - 0.015, site_lon + 0.015)

utm_bbox <- convert_coords(lat = lat_range, long=long_range, to = crs(elev_img))

extent_zoomed <- extent(utm_bbox[1], utm_bbox[2], utm_bbox[3], utm_bbox[4])
loc_zoom <- crop(elev_img, extent_zoomed)
loc_zoom_mat <- raster_to_matrix(loc_zoom)

osm_bbox <- c(long_range[1],lat_range[1], long_range[2],lat_range[2])

loc_water <- opq(osm_bbox) %>% add_osm_feature("waterway") %>% osmdata_sf() 
loc_water_lines <- st_transform(loc_water$osm_lines)

loc_water_lines <- loc_water_lines %>% filter(name != "Schuylkill River") %>% filter(name != "Norristown Dam")

# Get logo onto view
logo_file <- "in/usgs_sfacw_logo_black.png"
logo <- imager::load.image(logo_file)
logo <- as.data.frame(logo)
logo$z <- 400
```

## Set up main plot

This chunk is where the main ggplot grob definition and set-up occurs. You may have several ggplot grobs, depending on your approach. For each one, if you define it with the `<-` assignment operator and surround the whole code statement in parentheses, you'll be able to preview here what the plot looks like before the next cowplot step.

> Note: The hardest part of this process is getting text to be the optimum size in the final `png` for Twitter. The font sizes here will **not** be the same as in the final image after the cowplot step. Make sure to check the output `png` for true font sizing and plot composition. 

```{r plotting}
##### Plotting #####

# 2D Map
elev_matrix %>%
  sphere_shade(texture = "imhof4") %>%
  #add_water(watermap, color = "blue") %>%
  add_shadow(raymat, max_darken = 0.5) %>%
  add_shadow(ambmat, max_darken = 0.5) %>%
  add_overlay(overlay_img, alphalayer = 0.7) %>%
  add_overlay(generate_line_overlay(loc_water_lines, extent = extent_zoomed, heightmap = loc_zoom_mat, 
                                    linewidth = 2, color = "dodgerblue")) %>% 
  plot_map()

# 3D Map
rgl::clear3d()
elev_matrix %>% 
  sphere_shade(texture = "desert") %>% 
  add_overlay(overlay_img, alphalayer = 0.8) %>%
  add_shadow(raymat, max_darken = 0.1) %>%
  add_shadow(ambmat, max_darken = 0.1) %>%
  add_overlay(generate_line_overlay(loc_water_lines, extent = extent_zoomed, heightmap = loc_zoom_mat, 
                                    linewidth = 2, color = "dodgerblue")) %>%
  plot_3d(elev_matrix, zscale = zscale, windowsize = c(1200, 1000),
          water = TRUE, waterdepth = 34, wateralpha = 0.8, watercolor = "dodgerblue",
          waterlinecolor = "white", waterlinealpha = 0.5, soliddepth = -max(elev_matrix)/zscale,
          theta = 30, phi = 35, zoom = 0.7, fov = 50, background = "#00264c", solidcolor = "black")

# Add site name
render_label(elev_matrix, x = site_x_img, y = site_y_img, z = 250, zscale = zscale, 
             text = site_info$station_nm, freetype = F, textcolor = "white")

# Timestamp
render_label(elev_matrix, x = 100, y = 300, z = 1500, zscale=zscale, adjustvec = c(0.5,0.5),
             text = paste0("Date-time: ",river_data$dateTime[30]), freetype = F, alpha=0, textcolor = "white")

# Discharge
render_label(elev_matrix, x = 100, y = 300, z = 1400, zscale=zscale, adjustvec = c(0.5,0.5),
             text = paste0("Discharge: ",river_data$Flow_Inst[30], " cfs"), freetype = F, alpha=0, textcolor = "white")

# Add logo
#rgl::plot3d(x = xx$z-200, y = rev(xx$y)*0.1+10, z = rev(xx$x)*0.1 +100, col = xx$value, add = TRUE) # original
rgl::plot3d(x = (logo$x)*0.035+160, y = rev(logo$y)*0.05+3, z = (logo$z)*0.035 +190, col = logo$value, add = TRUE) # move to front face


```


## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

**Make sure to use the format for saving your png with the date of the prompt that you've been assigned to!** (e.g., `20230401_part-to-whole_cnell.png`)

```{r}
##### Create 3D GIF for output #####

# Set up parameters for loop to create GIF
n_frames <- nrow(river_data)
img_frames <- paste0("out/cache/ida_norristown_flood_", seq_len(n_frames), ".png")

# Loop statement to create each image
for (i in seq_len(n_frames)){#seq_len(n_frames)) {
  message(paste(" - image", i, "of", n_frames))
  
  elev_matrix %>%
    sphere_shade(texture = "imhof1") %>%
    add_shadow(ambmat, max_darken = 0.4) %>%
    add_shadow(raymat, max_darken = 0.4) %>%
    add_overlay(overlay_img, alphalayer = 0.7) %>%
    add_overlay(generate_line_overlay(loc_water_lines, extent = extent_zoomed, heightmap = loc_zoom_mat, 
                                      linewidth = 2, color = "dodgerblue")) %>%
    plot_3d(elev_matrix, solid = TRUE, shadow = TRUE, zscale = zscale, 
            windowsize = c(1200, 1000),
            water = TRUE, watercolor = "dodgerblue", 
            wateralpha = 0.7, waterlinealpha = 0.5,
            waterdepth = river_data$waterdepth[i], # pulls gage data
            phi = 30, theta = 35, fov = 50, zoom = 0.7, background = "#00264c", solidcolor = "black")
  
  # Add logo
  #rgl::plot3d(x = xx$z - 200, y = rev(xx$y) * 0.1 + 30, z = rev(xx$x) * 0.1 + 100, col = xx$value, add=TRUE) # original
  rgl::plot3d(x = (logo$x)*0.035+160, y = rev(logo$y)*0.05+3, z = (logo$z)*0.035 +190, col = logo$value, add = TRUE) # move to front face 
  
  # Add site name
  render_label(elev_matrix, x = site_x_img, y = site_y_img, z = 300, zscale=zscale, 
               text = "", freetype = F, linecolor = "white")
  
  # Timestamp
  render_label(elev_matrix, x = 100, y = 300, z = 1500, zscale=zscale, adjustvec = c(0.5,0.5),
               text = paste0("Date-time: ", river_data$dateTime[i]), freetype = F, alpha=0, textcolor = "white")
  
  # Discharge
  render_label(elev_matrix, x = 100, y = 300, z = 1400, zscale=zscale, adjustvec = c(0.5,0.5),
               text = paste0("Discharge: ",river_data$Flow_Inst[i], " cfs"), freetype = F, alpha=0, textcolor = "white")
  
  # Save as image
  render_snapshot(filename = img_frames[i], clear=TRUE)
}

# Build GIF using cached images - add labels here to control size better
magick::image_read(img_frames) %>%
  magick::image_annotate('Flooding in Norristown, Pennsylvania during Hurricane Ida \nSeptember 2021', 
                         location = "+35+35", color = "white", size = 40) %>%
  magick::image_annotate('Schuylkill River at Norristown, PA \n USGS 01473500',location="+460+420", 
                         color='#00264c',boxcolor = "gray80",size=22) %>%
  magick::image_write_gif(path="20230407_hazards_conlon_colgin.gif",delay=12/n_frames) # Number here is duration

### End
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. Hurricane Ida impacted southeastern Pennsylvania in September 2021, causing major flooding along rivers including Brandywine Creek, Perkiomen Creek, and Schuylkill River.

2. This animation uses gage height data to create an inundation map for each timestep that data was collected at the Schuylkill River at Norristown (USGS 01473500) throughout the event.

3. The animation was created using the rayshader R package, which allows for 2- and 3D rendering of geospatial data.

### Data source(s)

Elevation data: https://elevation.nationalmap.gov/arcgis/rest/services/3DEPElevation/ImageServer/exportImage 
Imagery data: https://utility.arcgisonline.com/arcgis/rest/services/Utilities/PrintingTools/GPServer/Export%20Web%20Map%20Task/execute 
Streamflow data: https://waterdata.usgs.gov/monitoring-location/01473500/


