---
title: 'Fish In Hot Water: Made for Chart Challenge'
author: "Ellie White"
date: "2023-04-03"
output:
  pdf_document: default
  html_document: default
---

## Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) 
library(readr) 
library(scales)
library(gganimate)
library(scico)
library(gifski)
```

## Load files
Copy-pasted data from paper to a csv. Paper is here: https://afspubs.onlinelibrary.wiley.com/doi/full/10.1002/mcf2.10076

```{r load}
fish_data <- read_csv("in/fish_data.csv")
```

This extra data came from the author in personal communications 04/03/2023

I modified the spreadsheet to be more R friendly

It's a time series record spanning 1950-2099 of fish spawning onset and cessation dates (modeled values)

```{r load_extra_data}
fish_data_ts <- read_csv("in/fish_data_from_nack.csv")

# pulled the 1951 values as origin dates and saved to this file for easy reference
# note that the year date is set to 2015 to be consistent with the rest of the reference dates but the year in the date doesn't matter; we care about month and day. 
fish_data_ts[fish_data_ts$year == 1951, ]
fish_data_origin_dates <- read_csv("in/fish_data_origin_dates.csv")
```

## Get data ready for plotting
```{r processing}
# change to factors 
fish_data <- fish_data |>
  mutate(species = factor(species, levels = c("American Shad", "Striped Bass")), 
         variable = factor(variable, levels = c("Onset", "Cessation", "Duration")),
         period = factor(period, levels = c("Historical", "Future")))

# take out duration and confidence intervals
fish_data <- fish_data[fish_data$variable %in% c("Onset", "Cessation"), ]
fish_data_uncertainty <- select(fish_data, -c("RCP_26", "RCP_45", "RCP_60", "RCP_85")) # save this to bring back in later
fish_data <- select(fish_data, -c("RCP_26_SE", "RCP_45_SE", "RCP_60_SE", "RCP_85_SE"))

# prep origin dates 
fish_data_origin_dates <- pivot_longer(fish_data_origin_dates, RCP_26:RCP_85, values_to = "origin_date", names_to = "condition")
fish_data_origin_dates$origin_date <- as.Date(fish_data_origin_dates$origin_date, format = "%m/%d/%Y")

# add in origin dates
fish_data_long <- pivot_longer(fish_data, RCP_26:RCP_85, values_to = "value", names_to = "condition")
fish_data_long <- full_join(fish_data_long, fish_data_origin_dates, by = c("species", "variable", "condition"))
fish_data_long$end_date <- fish_data_long$origin_date + fish_data_long$value

# prep uncertainty in dates to be brought in later
fish_data_uncertainty_long <- pivot_longer(fish_data_uncertainty, RCP_26_SE:RCP_85_SE, names_to = "condition", values_to = "standard_error")
fish_data_uncertainty_long$condition <- substr(fish_data_uncertainty_long$condition, 1, 6)

# make into wide format 
fish_data_long <- fish_data_long |> 
  select(-c(origin_date, value))
fish_data_wide <- pivot_wider(fish_data_long, names_from = "variable", values_from = "end_date")
names(fish_data_wide)[names(fish_data_wide)=="Onset"] <- "onset"
names(fish_data_wide)[names(fish_data_wide)=="Cessation"] <- "cessation"

# bring in uncertainty that was prepped before
fish_data_uncertainty_wide <- pivot_wider(fish_data_uncertainty_long, names_from = "variable", values_from = "standard_error")
names(fish_data_uncertainty_wide)[names(fish_data_uncertainty_wide)=="Onset"] <- "onset_se"
names(fish_data_uncertainty_wide)[names(fish_data_uncertainty_wide)=="Cessation"] <- "cessation_se"

fish_data_wide <- full_join(fish_data_wide, fish_data_uncertainty_wide, by = c("species", "period", "condition"))
```

```{r processing_extra_data, include = FALSE}
# change to factors
fish_data_ts <- fish_data_ts|>
  mutate(species = factor(species, levels = c("American Shad", "Striped Bass")),
         condition = factor(condition, levels = c("RCP_26", "RCP_45", "RCP_60", "RCP_85", labels = c("RCP 2.6", "RCP 4.5", "RCP 6.0", "RCP 8.5"))),
         period = factor(ifelse(year <= 2012, "Historical", "Future")),
         )

# fix date column, the minus one is to match the numbers Nack gave in his spreadsheet.
fish_data_ts$Onset <- as.Date(fish_data_ts$start_date_julianday, origin = as.Date("2015-01-01"))-1
fish_data_ts$Cessation <- as.Date(fish_data_ts$end_date_julianday, origin = as.Date("2015-01-01"))-1

# get rid of duration and extra dates
fish_data_ts <- fish_data_ts |>
  select(-c("duration", "start_date_julianday", "end_date_julianday", "start_date", "end_date"))

# make long format
fish_data_ts_long <- pivot_longer(fish_data_ts, Onset:Cessation, names_to = "spawning", values_to = "value")
fish_data_ts_long$spawning <- factor(fish_data_ts_long$spawning, levels = c("Onset", "Cessation"))
```

Now to add extra things for plotting purposes only:
```{r plotting_add_extra_info}
# add in y location for segment plot
y_location <- tibble(condition = c("RCP_26", "RCP_45", "RCP_60", "RCP_85"),
                     y = rep(c(1:4)*2 + 5))
fish_data_wide <- full_join(fish_data_wide, y_location, by = c("condition"))

# offset the y locations and mess with spacing for historical vs. future so they slightly overlap but don't overlap across the RCPs
fish_data_wide <- fish_data_wide |>
  mutate(y_offset = ifelse(period == "Historical", y - 0.5, y))
```

## Set up main plot
Theme:
```{r plotting_setup}
theme_usgs <- function(legend.position = "right"){
  theme(
    plot.title = element_text(vjust = 3, size = 14, face = "bold", family="sans"),
    plot.subtitle = element_text(vjust = 3, size = 12,family="sans"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white"),
    legend.background = element_blank(),
    legend.justification=c(0, 0),
    legend.position = legend.position,
    legend.key = element_blank(),
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    axis.title.x = element_text(size = 10, family="sans"),
    axis.title.y = element_text(vjust = 1, angle = 90, size = 9, family="sans"),
    axis.text.x = element_text(size = 10, vjust = -0.25, colour = "black", 
                               family="sans", margin=margin(10,5,20,5,"pt")),
    axis.text.y = element_text(size = 10, hjust = 1, colour = "black", 
                               family="sans", margin=margin(5,10,10,5,"pt")),
    axis.ticks = element_line(colour = "black", linewidth = 0.1),
    axis.ticks.length = unit(-0.25 , "cm")
  )
}
```

## Produce plots
```{r plotting_simple, include = FALSE}
# first, a dot chart to see what the data basically looks like
ggplot(data = fish_data_long, aes(x = end_date, y = variable)) +
  geom_point(aes(col = condition)) + 
  facet_wrap(~species) +
  theme_usgs()
```

```{r plotting_main}
# main base plot
library(ggnewscale)
ggplot() +
  geom_segment(data = fish_data_wide[fish_data_wide$period == "Historical", ], aes(x = onset, xend = cessation, y = y_offset, yend = y_offset), alpha = 0.7, linewidth = 8 , show.legend = FALSE, col = "grey") + # main hisotrical bar
  geom_segment(data = fish_data_wide[fish_data_wide$period == "Historical", ], aes(x = onset-onset_se, xend = cessation + cessation_se, y = y_offset, yend = y_offset), alpha = 0.25, linewidth = 8 , show.legend = FALSE, col = "grey") + # bar with standard error end bits
  new_scale_colour() +
  geom_segment(data = fish_data_wide[fish_data_wide$period == "Future", ], aes(x = onset, xend = cessation, y = y_offset, yend = y_offset), alpha = 0.7, linewidth = 8 , show.legend = FALSE, col = "#815aa5") + # main future bar
  geom_segment(data = fish_data_wide[fish_data_wide$period == "Future", ], aes(x = onset-onset_se, xend = cessation + cessation_se, y = y_offset, yend = y_offset), alpha = 0.25, linewidth = 8 , show.legend = FALSE, col = "#815aa5") + # bar standard error end bits
  scale_x_date(limits = c(as.Date("2015-01-01"), as.Date("2015-12-31")), date_breaks = "1 month", date_minor_breaks = "1 week", date_labels = "%B") + # make xlabels pretty
  scale_y_continuous(limits = c(0, 13)) +  # add in some room in the y axis for the fish
  coord_polar(theta = "x", direction = 1, start = -1.57*1.5) + # comment this out to make the plot work in Cartesian coordinates first and then polarize
  facet_wrap(~species) +
  labs(x = "",
       y = "",
       title = "FISH IN HOT WATER",
       subtitle = "Under projected climate change scenarios, the American Shad and Striped Bass of the Hudson River Estuary are predicted to spawn 15 days earlier on average."
       # caption = "Data Source: Nack, C. et. al. (2019). https://doi.org/10.1002/mcf2.10076
       # Plot made by Ellie White, ewhite@usgs.gov 04/02/2023"
       ) +
  theme_bw()+
  theme(plot.title = element_text(vjust = 3, size = 14, face = "bold", family="sans"),
        plot.subtitle = element_text(vjust = 3, size = 12,family="sans"),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.border = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank())

ggsave("out/26_local-change_ewhite_base.png", width = 16, height = 9, units = "in", dpi = 1200)
```

```{r plotting_paper_figure, include = FALSE}
# recreating figure in paper to make sure I haven't accidentally messed up the data
ggplot(data = fish_data_ts_long[fish_data_ts_long$year > 1950, ], aes(x = year, y = value)) +
  geom_point(aes(shape = species, col = species)) +
  facet_wrap(~condition + spawning, ncol = 2, scales = "free") +
  labs(x="", y="Predicted Date of Spawning") +
  theme_usgs()
# it matches! Hooray
```

## Produce annual viz (final plot)
```{r plotting_annual}
# make the yearly data wide so onset and cessation columns are side by side
fish_data_ts_wide <- pivot_wider(fish_data_ts_long, names_from = "spawning", values_from = "value")

# filter by RCP
fish_data_ts_plot <- fish_data_ts_wide |>
  filter(condition == "RCP_85") |>
  filter(year != 1950) |> # the data for this year was off so I filtered it out
  mutate(year = as.integer(year)) 

# this business is for plot legend breaks and labels (they were messing up because the legend is a date)
df_as <- fish_data_ts_plot[fish_data_ts_plot$species=="American Shad", ]
df_sb <- fish_data_ts_plot[fish_data_ts_plot$species=="Striped Bass", ]
legend_date_breaks_as <- diff(range(df_as$Onset)) * 0:4 / 4 + min(df_as$Onset)
legend_date_labels_as <- format(legend_date_breaks_as, "%W") # this "%W" format is week of the year (0-53)
legend_date_breaks_sb <- diff(range(df_sb$Onset)) * 0:4 / 4 + min(df_sb$Onset)
legend_date_labels_sb <- format(legend_date_breaks_sb, "%W")

ggplot() +
  geom_hline(yintercept = 1950, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_hline(yintercept = 2012, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_hline(yintercept = 2099, linetype = "dashed", color = "grey50", linewidth = 0.5) + 
  geom_segment(data = df_as, aes(x = Onset, xend = Cessation, y = year, yend = year, col = Onset), linewidth = 0.5) +
  scale_color_scico(palette = "tokyo", guide = guide_colorbar(order = 1), labels = legend_date_labels_as, breaks = as.numeric(legend_date_breaks_as)) +
  new_scale_colour() +
  geom_segment(data = df_sb, aes(x = Onset, xend = Cessation, y = year, yend = year, col = Onset), linewidth = 0.5) +
  scale_color_scico(palette = "tokyo", guide = guide_colorbar(order = 2), labels = legend_date_labels_sb, breaks = as.numeric(legend_date_breaks_sb)) +
  scale_x_date(limits = c(as.Date("2015-01-01"), as.Date("2015-12-31")), date_breaks = "1 month", date_minor_breaks = "1 week", date_labels = "%B") +
  scale_y_continuous(limits = c(1900, 2120)) +
  coord_polar(theta = "x", direction = 1, start = -1.57*1.5) + # start is in radians, 90 Deg is Jan
  facet_wrap(~species) +
  labs(x = "",
       y = "",
       title = "FISH IN HOT WATER",
       subtitle = "RCP 8.5: Business-as-Usual."
       ) +
  theme_bw()+
  theme(plot.title = element_text(vjust = 3, size = 14, face = "bold", family="sans"),
        plot.subtitle = element_text(vjust = 3, size = 12,family="sans"),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.border = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank(), 
        legend.position = "bottom", 
        legend.title = element_blank()
        )
ggsave("out/26_local-change_ewhite_base2.png", width = 16, height = 9, units = "in", dpi = 1200)
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. The American Shad and the Striped Bass are migratory species needing both freshwater and marine habitats to complete their life cycle. This makes them particularly vulnerable to human activities. The Hudson River Shad has declined in stock so much that all its fisheries were closed in 2010. The Striped Bass, while declining in relative abundance, still remains the most important game fish in the Hudson River.

2. Under projected climate change scenarios, the American Shad and Striped Bass of the Hudson River Estuary are predicted to spawn 15 days earlier on average. 

3. Under RCP 8.5 (“business-as-usual”),  the American Shad and Striped Bass of the Hudson River Estuary are projected to start and stop spawning 20 and 26 days earlier on average respectively.

### Data source(s)

Paper is here: https://afspubs.onlinelibrary.wiley.com/doi/full/10.1002/mcf2.10076

Citation: Nack, C. C., Swaney, D. P., & Limburg, K. E. (2019). Historical and projected changes in spawning Phenologies of American Shad and Striped bass in the 
Hudson River Estuary. Marine and Coastal Fisheries, 11(3), 271-284.

DOI: https://doi.org/10.1002/mcf2.10076

### Process 

1. Produced base plots called `out/25_local-change_ewhite_base.png` and `out/25_local-change_ewhite_base2.png` with ggplot.

2. Made markups in PowerPoint. 

3. Saved final plot as picture and called them `out/25_local-change_ewhite_final.png` and `out/25_local-change_ewhite_final2.png`.




