---
title: "Green Energy Chart"
author: "Mandie Carr"
date: "2023-04-22"
output: html_document
---

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) # includes ggplot
library(readr) # best for reading and writing csvs


# These are used for layouts of viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo

```



## Load files


```{r load}

usa <- "in/ElectricityGeneration_US_2000_2020.csv"
us_data <- readr::read_csv(usa, col_types = cols())


```

## Get data ready for plotting

```{r processing}
# Copy 2020 data into 2021 to get full last "step" in plot
electricity <- us_data %>%
  dplyr::add_row(us_data |> filter(Year==2020) |> mutate(Year = 2021))

# Convert data to long format
green_energy <- electricity %>%
  tidyr::pivot_longer(!Year, names_to = "Technology", values_to = "GWh")


```

## Set up main plot

```{r plotting}
# Load some custom fonts and set some custom settings
title_font <- "Boogaloo"
sysfonts::font_add_google("Boogaloo")
supporting_font <- "Baloo 2"
sysfonts::font_add_google("Baloo 2")
annotation_font <- "Neucha"
sysfonts::font_add_google("Neucha")
showtext::showtext_opts(dpi = 300, regular.wt = 200, bold.wt = 700)
showtext::showtext_auto(enable = TRUE)
title_font_size <- 54
annotation_font_size <- 36
supporting_font_size <- 32

# Define colors
background_color = "#000000"
annotation_color = "white"
supporting_font_color = "#979597"
title_font_color = "#FFE700"
#colors <- c("#FFF9C4","#FFF176","#FFEB3B","#FBC02D","#F57F17")
colors <- c("#4DEEEA","#74EE15","#FFE700","#F000FF","#1685F8")
names(colors) <- c("Hydropower","Wind","Solar","Bioenergy","Geothermal")
# order data by color order
green_energy <- mutate(green_energy, Technology = factor(Technology, levels = names(colors)))


# The background canvas for your viz
canvas <- grid::rectGrob(
  x = 0, y = 0, 
  width = 16, height = 16,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color)
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_white.png") 

# Load in lightning bolt
lightning_bolt <- magick::image_read("in/lightning-bolt.png") 

# Main plot
(main_plot <- ggplot(green_energy) +
  geom_step(aes(Year, GWh, color = Technology),lwd = 5, alpha = 0.1) +
  geom_step(aes(Year, GWh, color = Technology),lwd = 3, alpha = 0.3) +
  geom_step(aes(Year, GWh, color = Technology),lwd = 1.5, alpha = 1) +
  scale_x_continuous(
    breaks = seq(2000,2020,by=5),
    expand = c(0,0),
    name = '') +
  scale_color_manual(values = colors,
                    name  = '') +
  scale_y_continuous(name = '',
                     limits = c(0, 350000),
                     breaks = c(50000, 100000, 150000, 200000, 250000, 300000, 350000), 
                     labels = c("50,000", "100,000", "150,000", "200,000", "250,000", 
                                "300,000", stringr::str_wrap('350,000 GWh', 8)),
                     expand = c(0,0)) +
  theme(axis.text.y = element_text(hjust = 1, color = supporting_font_color, size = supporting_font_size, family = supporting_font),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(color = supporting_font_color, size = supporting_font_size, family = supporting_font),
        plot.background = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.y = element_line(color = '#F3F3F3', linewidth = 0.5, linetype = 'dotted'),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = c(1.23, 0.7),
        legend.text = element_text(color = supporting_font_color, family = supporting_font, size = supporting_font_size, margin = margin(r = 17)),
        legend.spacing.y = unit(4.7, 'cm'),
        legend.spacing.x = unit(0.5, 'cm'),
        legend.key.width = unit(1, 'cm'),
        legend.key.height = unit(0.2, 'cm'),
        legend.background = element_blank(),
        legend.key = element_blank()) +
  guides(color = guide_legend(nrow=5, 
                             byrow = TRUE))
 )

# build sub plots
plot_list <- purrr::map(seq(1, length(colors)), function(tech) {
  factor_levels <- c(names(colors)[names(colors) !=  names(colors)[tech]], names(colors)[tech])

  green_energy <- mutate(green_energy, Technology = factor(Technology, levels = factor_levels))

  plot_colors <- rep(NA, length(colors))
  plot_colors[length(plot_colors)] <- colors[tech]

  ggplot(green_energy) +
    geom_step(aes(Year, GWh, color = Technology, linewidth = Technology))+
    scale_discrete_manual("linewidth",values=c(rep(0.5,4),1.5)) +
    scale_x_discrete(
    breaks = c(2000,2005,2010,2015,2020),
    expand = c(0,0),
    name = '') +
    scale_color_manual(values = plot_colors, name  = '', na.value = '#CBCACB') +
    scale_y_continuous(name = '',
                     limits = c(0, 350000),
                     breaks = c(50000, 100000, 150000, 200000, 250000, 300000, 350000), 
                     labels = c("50,000", "100,000", "150,000", "200,000", "250,000", 
                                "300,000", stringr::str_wrap('350,000 GWh', 8)),
                     expand = c(0,0)) +
  theme_void() +
    theme(legend.position = 'None',
          plot.margin = unit(c(1,0.6,1,0.6), "cm"))
})


```


## Produce final plot

```{r cowplot, fig.width = 16, fig.height = 16, warning = FALSE}
ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 16, width = 16,
            hjust = 0, vjust = 1) +
  # small multiples
  draw_plot(plot_grid(plotlist = plot_list, nrow = 5),
            x = 0.76,
            y = 0.145,
            width = 0.2,
            height = .7) +
  # the main plot
  draw_plot(main_plot,
            x = 0.028,
            y = 0.125,
            height = 0.7,
            width = 0.7) +
  # annotations
  # drought
draw_label("Extreme drought in the \nPacific Northwest",
             x = 0.21,
             y = 0.47,
             hjust = 0,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = annotation_font,
             color = annotation_color,
             size = annotation_font_size) +
  geom_curve(aes(xend = 0.185, x = 0.20,
                 yend = 0.555, y = 0.50),
             arrow = grid::arrow(length = unit(0.75, 'lines')),
             curvature = -0.45,
             angle = -25,
             color = annotation_color,
             lineend = 'round',
             linewidth = 1.5) +
    # more wind than hydro
  draw_label("More wind than hydro",
             x = 0.70,
             y = 0.83,
             hjust = 1,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = annotation_font,
             color = annotation_color,
             size = annotation_font_size) +
  geom_curve(aes(xend = 0.685, x = 0.66,
                 yend = 0.79, y = 0.82),
             arrow = grid::arrow(length = unit(0.75, 'lines')), 
             curvature = 0.25,
             angle = -55,
             color = annotation_color,
             lineend = 'round',
             linewidth = 1.5) +
  # Title
 draw_label(
    "Electricity generated by renewable energy in the U.S.",
             x = 0.04,
             y = 0.94,
             hjust = 0,
             vjust = 1,
             lineheight = 1,
             fontfamily = title_font,
             color = title_font_color,
             size = title_font_size) +
   # Add lightning bolt
  draw_image(lightning_bolt, 
             x = 0.91,
             y = 0.898,
             width = 0.025, 
             hjust = 0, vjust = 0, 
             halign = 0, valign = 0)+
  # explainer text
  draw_label(paste('Mandie Carr, USGS',"Data: https://www.irena.org/Publications/2022/Jul/Renewable-Energy-Statistics-2022", sep = '\n'),
             fontfamily = supporting_font,
             x = 0.96,   
             y = 0.04,
             size = 22,
             hjust = 1,
             vjust = 0,
             color = supporting_font_color)+
  # Add logo
  draw_image(usgs_logo, 
             x = 0.04,
             y = 0.04,
             width = 0.12, 
             hjust = 0, vjust = 0, 
             halign = 0, valign = 0)


# Save the final image in 16 by 16 format
# !! Use format for saving with the date of your prompt: 
#         YYYYMMDD_prompt_name ()
ggsave(filename = "out/20230422_green-energy_ancarr.png", 
       width = 16, height = 16, dpi = 300)
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. Electricity generated by wind and solar energy increased in the U.S. from 2000 to 2020. Hydropower went up and down, and bioenergy and geothermal energy did not change much.
2. Hydropower was the biggest renewable source of electricity in the U.S. until 2020, when wind took the lead.
3. An extreme drought in the Pacific Northwest in 2001 caused a big drop in electricity generated by hydropower. The drought had a big effect on energy production because much of the Nation's hydropower capacity is in the Pacific Northwest.

### Data source

Electricity generation data were downloaded from the International Renewable Energy Agency (IRENA) at https://pxweb.irena.org/pxweb/en/IRENASTAT/IRENASTAT__Power%20Capacity%20and%20Generation/RE-ELECGEN_2022_cycle2.px/

Data filters were as follows:

* Region/country/area = USA
* Technology = Hydropower, Wind, Solar, Bioenergy, Geothermal
* Year = 2000 - 2020

The same data were published in this report:
<p>IRENA, 2022, Renewable energy statistics 2022, International Renewable Energy Agency, Abu Dhabi. https://www.irena.org/Publications/2022/Jul/Renewable-Energy-Statistics-2022
