---
title: "Example for Chart Challenge"
author: "Althea Archer"
date: "2023-03-10"
output: html_document
---

## Instructions for using this template

> Make sure you have copied the template subdirectory and named it with your prompt's date, prompt's name and your name (e.g., `/01_part-to-whole_cnell`) before editing this document!

1. Put all comments either in the html text (like this) or in the `r` chunks (delineated with 3 "`" marks) to explain your process to Vizlab reviewers
2. Put all R code in the chunks. They're organized for you to (in order): do some set-up steps (load libraries), load in any data (manually or with built in packages), process your data for plotting, create plot using ggplot, and finally compose the twitter image with cowplot.
3. You can run individual code chunks while developing code and plots. Note: the plots that are created in the RStudio coding window *do not* match the plot as created through the ggsave function. 
4. To preview the final plot, use the `Knit` button above. That will create a preview of the final plot in the output `html` and will save the `png` in the `out/` folder. On many systems, the preview in the script document and the html will not match the output `png`, so please refer to the `png` for final proofing.
5. When you're happy with the final plot, fill out the supporting information at the bottom and push to gitlab for review. Note: only commit input data if it can't be downloaded directly with code (e.g., `sbtools` or `dataRetrieval`).

This website has great information on how to use RMarkdown: https://yihui.org/knitr/

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```



Define libraries here. 

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) # includes ggplot
library(readr) # best for reading and writing csvs

library(lubridate)
library(sf)

# These are used for the layout of the viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo
```



## Load files

Save any files that you are using directly in the `in` folder. Then read them in to the environment with this chunk of code. Remember, use the RStudio project and relative file pathways, and never include any personal computer file pathways.

This is also a good place to load any data from other built-in packages such as `dataRetrieval` or `spData`, etc.


```{r load}
# El Nino indices from NOAA (scraped from https://origin.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/ONI_v5.php)
elnino <- readr::read_csv("in/elnino_index.csv")

```

## Get data ready for plotting

First, process the el Niño/la Niña indices. The data come in with month as columns, years as rows. Need to make it long format with month and year as separate fields.

```{r processing}
# Rename months. They are 3-month windows centered on one month. Rename to that one month
elnino_rename <- elnino |>
  rename(January = DJF,
         February = JFM,
         March = FMA,
         April = MAM,
         May = AMJ,
         June = MJJ,
         July = JJA,
         August = JAS,
         September = ASO,
         October = SON,
         November = OND,
         December = NDJ)

# Gather into long format
elnino_long <- elnino_rename |>
  pivot_longer(cols = !Year, 
               names_to = "month",
               values_to = "index") |>
  # add month in numeric form and create date field
  mutate(month_numeric = match(month, month.name),
         date = lubridate::my(sprintf("%s-%s", month_numeric, Year)))

# Create subset for plotting dates as y-axis on the chart
year_labels <- data.frame(
  date = as.Date(sprintf("%s-01-01", seq(1950, 2020, by = 5))),
  index = rep(0, 15),
  label = seq(1950, 2020, by = 5)
)

```


## Set up main plot


```{r plotting}
# Load some custom fonts and set some custom settings
font_legend <- "Satisfy"
sysfonts::font_add_google("Satisfy")
supporting_font <- "Source Sans Pro"
sysfonts::font_add_google("Source Sans Pro")
handwriting_font <- "Shadows Into Light"
sysfonts::font_add_google("Shadows Into Light")
showtext::showtext_opts(dpi = 300, regular.wt = 200, bold.wt = 700)
showtext::showtext_auto(enable = TRUE)

# Define colors
background_color = "white"
font_color = "black"
elnino_color = "#7DaDB8"
elnino_dark = "#1194a7"
lanina_color = "#A280AE"
lanina_dark = "#A280AE"

# The background canvas for your viz
canvas <- grid::rectGrob(
  x = 0, y = 0, 
  width = 16, height = 9,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color)
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_white.png") |>
  image_colorize(100, "#36454F")

# Load background image and make it transparent
background_image <- magick::image_read("in/background_paint.png") |>
  magick::image_fx(expression = "0.45*a", channel = "alpha")

# Create the main plot 
(main_plot <- ggplot(elnino_long, 
                     aes(x = date, y = index, group = date)) +
    # main data
    geom_linerange(aes(ymin = 0, ymax = index, color = sign(index))) +
    scale_color_gradient2(low = lanina_color, mid = "black", high = elnino_color)+
    scale_x_date(breaks = "5 year", date_labels = "%Y", 
                 limits = as.Date(c("1949-01-01", "2023-09-01")))+
    # y-axis on chart
    geom_linerange(data = year_labels, aes(ymin = -0.135, ymax = 0.135), color = "#36454F") +
    geom_hline(yintercept = 0, color = "#ffffff", linewidth = 6)+
    geom_text(data = year_labels, aes(label = label), color = "#36454F", size = 5,
              fontface = "bold", family = supporting_font) +
    # thematic settings
    theme_minimal()+
    ylim(c(-2.6,2.6))+
    ylab("Oceanic\nNiño\nIndex\n(ONI)")+
    theme(legend.position = "none",
          panel.background = element_blank(),
          axis.title.x = element_blank(),
          axis.title.y = element_text(family = supporting_font, face = "bold",
                                      angle = 0, vjust = 0.42, hjust = 0,
                                      size = 18, color = "#36454F"),
          axis.text.y = element_text(family = supporting_font,
                                   size = 14, color = "#36454F", face = "bold"),
          axis.text.x = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
 )

```


## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

**Make sure to use the format for saving your png with the date of the prompt that you've been assigned to!** (e.g., `20230401_part-to-whole_cnell.png`)

```{r cowplot, fig.width = 16, fig.height = 9}
ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 9, width = 16,
            hjust = 0, vjust = 1) +
  # watercolored background
  draw_image(background_image, 
             x = 0.025,
             y = 0.152,
             height = 0.8,
             width = 0.98)+
  # the main plot
  draw_plot(main_plot,
            x = 0.01,
            y = 0.18,
            height = 0.68,
            width = 0.99) +
  # explainer text
  draw_label("Chart by Althea A. Archer.\nData from NOAA: origin.cpc.ncep.noaa.gov",
             fontfamily = supporting_font,
             x = 0.97,   
             y = 0.04,
             size = 14,
             hjust = 1,
             vjust = 0,
             color = "#36454F")+
  # el nino title
  draw_label("el Niño",
             x = 0.05,
             y = 0.95,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = elnino_dark,
             size = 60) +
  # el nino description
  draw_label("el Niño periods bring wetter weather in the Southern U.S.\n      and drier weather in the North.",
             x = 0.30,
             y = 0.93,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = handwriting_font,
             color = "#36454F",
             size = 24) +
  # la nina
  draw_label("la Niña",
             x = 0.05,
             y = 0.15,
             hjust = 0,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = lanina_dark,
             size = 60) +
  # la nina description
    draw_label("la Niña periods bring drier weather in the Southern U.S.\n      and wetter weather in the North.",
             x = 0.30,
             y = 0.20,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = handwriting_font,
             color = "#36454F",
             size = 24) +
  # Add logo
  draw_image(usgs_logo, 
             x = 0.03,
             y = 0.04,
             width = 0.08, 
             hjust = 0, vjust = 0, 
             halign = 0, valign = 0)

# Save the final image in Twitter's 16 by 9 format
# !! Use format for saving with the date of your prompt: 
#         YYYYMMDD_prompt_name ()
# e.g. `20230101_part-to-whole-cnell.png`
ggsave(filename = "out/20230421_down-upward_aaarcher.png", 
       width = 16, height = 9, dpi = 300)
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. Ocean currents cycle between warmer (el Niño) and cooler (la Niña) periods.   
2. In the U.S., el Niño causes wetter weather in the south and la Niña causes wetter weather in the north.  

### Data source(s)

NOAA Climate Prediction Center

https://www.climate.gov/news-features/understanding-climate/climate-variability-oceanic-niño-index 

