---
title: "Example for Chart Challenge"
author: "Althea Archer"
date: "2023-03-10"
output: html_document
---

## Instructions for using this template

> Make sure you have copied the template subdirectory and named it with your prompt's date, prompt's name and your name (e.g., `/01_part-to-whole_cnell`) before editing this document!

1. Put all comments either in the html text (like this) or in the `r` chunks (delineated with 3 "`" marks) to explain your process to Vizlab reviewers
2. Put all R code in the chunks. They're organized for you to (in order): do some set-up steps (load libraries), load in any data (manually or with built in packages), process your data for plotting, create plot using ggplot, and finally compose the twitter image with cowplot.
3. You can run individual code chunks while developing code and plots. Note: the plots that are created in the RStudio coding window *do not* match the plot as created through the ggsave function. 
4. To preview the final plot, use the `Knit` button above. That will create a preview of the final plot in the output `html` and will save the `png` in the `out/` folder. On many systems, the preview in the script document and the html will not match the output `png`, so please refer to the `png` for final proofing.
5. When you're happy with the final plot, fill out the supporting information at the bottom and push to gitlab for review. Note: only commit input data if it can't be downloaded directly with code (e.g., `sbtools` or `dataRetrieval`).

This website has great information on how to use RMarkdown: https://yihui.org/knitr/

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```



Define libraries here. 

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) # includes ggplot
library(readr) # best for reading and writing csvs

library(lubridate)
library(sf)

# These are used for the layout of the viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo
```



## Load files

Save any files that you are using directly in the `in` folder. Then read them in to the environment with this chunk of code. Remember, use the RStudio project and relative file pathways, and never include any personal computer file pathways.

This is also a good place to load any data from other built-in packages such as `dataRetrieval` or `spData`, etc.


```{r load}
# El Nino indices from NOAA (scraped from https://origin.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/ONI_v5.php)
elnino <- readr::read_csv("in/elnino_index.csv")

```

## Get data ready for plotting

First, process the El Nino/La Nina indices. The data come in with month as columns, years as rows. Need to make it long format with month and year as separate fields.

```{r processing}
# Rename months. They are 3-month windows centered on one month. Rename to that one month
elnino_rename <- elnino |>
  rename(January = DJF,
         February = JFM,
         March = FMA,
         April = MAM,
         May = AMJ,
         June = MJJ,
         July = JJA,
         August = JAS,
         September = ASO,
         October = SON,
         November = OND,
         December = NDJ)

# Gather into long format
elnino_long <- elnino_rename |>
  pivot_longer(cols = !Year, 
               names_to = "month",
               values_to = "index") |>
  # add month in numeric form and create date field
  mutate(month_numeric = match(month, month.name),
         date = lubridate::my(sprintf("%s-%s", month_numeric, Year)))


```

Process drought data by selecting the 100 month severe droughts by year



## Set up main plot

This chunk is where the main ggplot grob definition and set-up occurs. You may have several ggplot grobs, depending on your approach. For each one, if you define it with the `<-` assignment operator and surround the whole code statement in parentheses, you'll be able to preview here what the plot looks like before the next cowplot step.

> Note: The hardest part of this process is getting text to be the optimum size in the final `png` for Twitter. The font sizes here will **not** be the same as in the final image after the cowplot step. Make sure to check the output `png` for true font sizing and plot composition. 

```{r plotting}
# Load some custom fonts and set some custom settings
font_legend <- "Satisfy"
sysfonts::font_add_google("Satisfy")
supporting_font <- "Source Sans Pro"
sysfonts::font_add_google("Source Sans Pro")
showtext::showtext_opts(dpi = 300, regular.wt = 200, bold.wt = 700)
showtext::showtext_auto(enable = TRUE)

# Define colors
background_color = "white"
font_color = "black"

# The background canvas for your viz
canvas <- grid::rectGrob(
  x = 0, y = 0, 
  width = 16, height = 9,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color)
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_white.png") |>
  image_colorize(100, "#666666")
background_image <- magick::image_read("in/IMG_7007.png") |>
  magick::image_fx(expression = "0.45*a", channel = "alpha")

(main_plot <- ggplot(elnino_long, 
                     aes(x = date, y = index, group = date)) +
    geom_linerange(aes(ymin = 0, ymax = index, color = sign(index))) +
    scale_color_gradient2(low = "#de77ae", mid = "black", high = "#4292c6")+
    scale_x_date(breaks = "5 year", date_labels = "%Y", 
                 limits = as.Date(c("1949-01-01", "2020-01-01")))+
    geom_hline(yintercept = 0, color = "black")+
    theme_minimal()+
    ylim(c(-2.6,2.6))+
    ylab("Oceanic\nNiño\nIndex\n(ONI)")+
    theme(legend.position = "none",
          panel.background = element_blank(),
          axis.title.x = element_blank(),
          axis.title.y = element_text(family = supporting_font, face = "bold",
                                      angle = 0, vjust = 0.4, hjust = 0,
                                      size = 18),
          axis.text = element_text(family = supporting_font,
                                   size = 18, face = "bold"),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
 )

```


## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

**Make sure to use the format for saving your png with the date of the prompt that you've been assigned to!** (e.g., `20230401_part-to-whole_cnell.png`)

```{r cowplot, fig.width = 16, fig.height = 9}
ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 9, width = 16,
            hjust = 0, vjust = 1) +
  draw_image(background_image, 
             x = -0.01,
             y = -0.025,
             height = 1.00,
             width = 1.05)+
  # the main plot
  draw_plot(main_plot,
            x = 0.01,
            y = 0.10,
            height = 0.68,
            width = 0.96) +
  # explainer text
  draw_label("Chart by Althea A. Archer.\nData from NOAA: origin.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/ONI_v5.php",
             fontfamily = supporting_font,
             x = 0.97,   
             y = 0.04,
             size = 14,
             hjust = 1,
             vjust = 0,
             color = "#999999")+
  
  # Title
  draw_label("From 1950 to today, naturally occurring changes in oceanic currents\nhave caused warmer (el Niño) and cooler (la Niña) periods.",
             x = 0.03,
             y = 0.95,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = supporting_font,
             color = font_color,
             size = 35) +
  # Title
  draw_label("el Niño",
             x = 0.12,
             y = 0.80,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = "#084594",
             size = 60) +
  draw_label("In the U.S., el Niño events cause wetter weather in the South\nand drier weather in the North.",
             x = 0.36,
             y = 0.76,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = "#084594",
             size = 18) +
  # Title
  draw_label("la Niña",
             x = 0.12,
             y = 0.19,
             hjust = 0,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = "#c51b7d",
             size = 60) +
  draw_label("In the U.S., la Niña events cause drier weather in the South\nand wetter weather in the North.",
             x = 0.36,
             y = 0.22,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = "#c51b7d",
             size = 18) +
  # Add logo
  draw_image(usgs_logo, 
             x = 0.03,
             y = 0.04,
             width = 0.08, 
             hjust = 0, vjust = 0, 
             halign = 0, valign = 0)

# Save the final image in Twitter's 16 by 9 format
# !! Use format for saving with the date of your prompt: 
#         YYYYMMDD_prompt_name ()
# e.g. `20230101_part-to-whole-cnell.png`
ggsave(filename = "out/20230421_down-upward_aaarcher.png", 
       width = 16, height = 9, dpi = 300)
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. Key takeaways here

### Data source(s)

Data sources here.

