---
title: "Distribution - Humans"
author: "Hayley Corson-Dosch"
date: "2023-04-08"
output: html_document
---

## Instructions for using this template

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


Define libraries here. 

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) # includes ggplot
library(readr) # best for reading and writing csvs
library(sf)
library(tigris)
library(rmapshaper)

# These are used for the layout of the viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo
```



## Load files

Save any files that you are using directly in the `in` folder. Then read them in to the environment with this chunk of code. Remember, use the RStudio project and relative file pathways, and never include any personal computer file pathways.

This is also a good place to load any data from other built-in packages such as `dataRetrieval` or `spData`, etc.


```{r load}
# Add your code here
conus_proj <- '+proj=aea +lat_0=23 +lon_0=-96 +lat_1=29.5 +lat_2=45.5 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs +type=crs'

conus_sf <- tigris::states(cb = TRUE) %>% 
  st_transform(conus_proj) %>%
  filter(STUSPS %in% state.abb[!state.abb %in% c('AK', 'HI')]) %>%
  rmapshaper::ms_simplify(keep = 0.4)

counties_sf <- tigris::counties() %>%
  st_transform(crs=conus_proj) %>%
  filter(STATEFP %in% conus_sf$STATEFP) %>% 
  rmapshaper::ms_simplify(keep = 0.2) %>%
  st_intersection(st_union(conus_sf))

county_wu <- readr::read_csv('in/county_water_use.csv', col_types = cols())
national_wu <- readr::read_csv('in/national_water_use.csv', col_types = cols())

```

## Get data ready for plotting
  
This next section is for doing any pre-processing steps, data joins, etc, to get your data ready for plotting with ggplot. Remember that most of the time, ggplot prefers data in a "wide" format.


```{r processing}
# Add your processing code here
rankings_2015 <- national_wu %>%
  filter(year == 2015, !(wu_category == 'commercial')) %>%
  mutate(wu_category = case_when(
    wu_category == 'thermoelectric power' ~ 'thermoelectric',
    wu_category == 'self-supplied domestic' ~ 'domestic',
    wu_category == 'self-supplied industrial' ~ 'industrial',
    TRUE ~ wu_category
  )) %>%
  arrange(desc(wu_bgd)) %>%
  pull(wu_category)

county_wu_totals_2015 <- county_wu %>%
  filter(year == 2015) %>%
  filter(!(wu_category == 'commercial')) %>% # not measured in 2015
  rowwise() %>%
  mutate(GEOID = paste0(
    paste(rep('0', 2 - nchar(as.character(state_cd))), collapse = ''), 
    state_cd, 
    paste(rep('0', 3 - nchar(as.character(county_cd))), collapse = ''), 
    county_cd), 
    .after = county_cd) %>%
  group_by(GEOID, year) %>%
  mutate(county_total_mgd = sum(total_watersource_mgd, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(per_of_total = total_watersource_mgd/county_total_mgd * 100,
         wu_category = case_when(
           wu_category == 'thermoelectric power' ~ 'thermoelectric',
           wu_category == 'self-supplied domestic' ~ 'domestic',
           wu_category == 'self-supplied industrial' ~ 'industrial',
           TRUE ~ wu_category
         ),
         wu_category = factor(wu_category, levels = rankings_2015))

county_wu_2015_sf <- counties_sf %>%
  left_join(county_wu_totals_2015, by = 'GEOID')

```

## Set up main plot

This chunk is where the main ggplot grob definition and set-up occurs. You may have several ggplot grobs, depending on your approach. For each one, if you define it with the `<-` assignment operator and surround the whole code statement in parentheses, you'll be able to preview here what the plot looks like before the next cowplot step.

> Note: The hardest part of this process is getting text to be the optimum size in the final `png` for Twitter. The font sizes here will **not** be the same as in the final image after the cowplot step. Make sure to check the output `png` for true font sizing and plot composition. 

```{r plotting}
# Load some custom fonts and set some custom settings
supporting_font <- "Source Sans Pro"
sysfonts::font_add_google("Source Sans Pro")
showtext::showtext_opts(dpi = 300, regular.wt = 200, bold.wt = 700)
showtext::showtext_auto(enable = TRUE)

# Define colors
background_color = "white"
font_color = "grey10"
label_font_size <- 14
category_colors <-  c(RColorBrewer::brewer.pal(7, 'Dark2'), '#A0BFEC') #, '#E2E2E2') # add blue for PS and light grey for NA values
names(category_colors) <- c('aquaculture','domestic','industrial','livestock','thermoelectric','mining','irrigation','public supply')

# The background canvas for your viz
plot_width <- 9
plot_height <- 9
canvas <- grid::rectGrob(
  x = 0, y = 0, 
  width = plot_width, height = plot_height,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color)
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_black.png") 

# Main plot
(main_plot <- ggplot() +
  geom_sf(data = conus_sf,
          fill = 'white',
          color = 'white',
          size = 0.6) +
  geom_sf(data = county_wu_2015_sf,
          aes(alpha = per_of_total,
              fill = wu_category),
          color = NA,
          size = 0.1) +
  geom_sf(data = conus_sf,
          fill = NA,
          color = background_color,
          size = 3) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_alpha(range = c(0.1,1), limits = c(0, 100), name = '') +
  scale_fill_manual(name = '', values = category_colors) +

  theme_void() +
  facet_wrap(~wu_category, labeller = labeller(wu_category = label_wrap_gen(25))) +
  theme(
    legend.position = 'None',
    strip.text.x = element_text(size = label_font_size, color = font_color, family = supporting_font, margin = margin(15, 0, 3, 0))
  )
 )

```


```{r plotting2}
# approach modified from https://timogrossenbacher.ch/2019/04/bivariate-maps-with-ggplot2-and-sf/#create-a-bivariate-choropleth
bivariate_color_scale <- purrr::map2_df(names(category_colors), category_colors, function(category, category_color) {
  tibble(
    wu_category = rep(category, 5),
    alpha = c(100, 75, 50, 25, 0),
    fill = rep(category_color, 5)
  )
}) %>%
  mutate(wu_category = factor(wu_category, levels = rev(rankings_2015)))

(legend <- ggplot() +
  geom_tile(
    data = bivariate_color_scale,
    mapping = aes(
      x = alpha,
      y = wu_category,
      fill = fill,
      alpha = alpha)
  ) + 
  scale_fill_identity() +
  scale_alpha(range = c(0.1,1), limits = c(0, 100), name = '') +
  scale_x_continuous(breaks = c(-12.5, 12.5, 37.5, 62.5, 87.5), labels = c('0', '25', '50', '75', '100% of withdrawals'), expand = c(0,0)) +
  scale_y_discrete(position = "right", expand = c(0,0)) +
  # labs(x = "Percent of total water use") +
  theme_void() +
  theme(
    legend.position = 'none',
    # axis.title.x = element_text(size = 16, hjust = 0),
    axis.text = element_text(size = label_font_size, color = font_color, family = supporting_font),
    axis.text.y = element_text(hjust = 0, margin = margin(0, 0, 0, 7)),
    axis.text.x = element_text(margin = margin(7, 0, 0, 0), hjust = 0)
  ))

```

## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

**Make sure to use the format for saving your png with the date of the prompt that you've been assigned to!** (e.g., `20230401_part-to-whole_cnell.png`)

```{r cowplot, fig.width = 16, fig.height = 9}
ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = plot_height, width = plot_width,
            hjust = 0, vjust = 1) +
  # the main plot
  draw_plot(main_plot,
            x = 0.005,
            y = 0.1,
            height = 0.75) +
  # the legend
  draw_plot(legend,
            x = 0.68,
            y = 0.07,
            width = 0.28,
            height = 0.25) +
  # explainer text
  draw_label("Data: https://waterdata.usgs.gov/nwis/wu",
             fontfamily = supporting_font,
             fontface = 'italic',
             x = 0.96,   
             y = 0.03,
             size = 10,
             hjust = 1,
             vjust = 0,
             color = font_color)+
  # Title
  draw_label("2015 water withdrawals",
             x = 0.04,
             y = 0.95,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = supporting_font,
             fontface = 'bold',
             color = font_color,
             size = 45) +
  # Add logo
  draw_image(usgs_logo, 
             x = 0.03,
             y = 0.02,
             width = 0.1, 
             hjust = 0, vjust = 0, 
             halign = 0, valign = 0)

# Save the final image in Twitter's 16 by 9 format
# !! Use format for saving with the date of your prompt: 
#         YYYYMMDD_prompt_name ()
# e.g. `20230101_part-to-whole-cnell.png`
ggsave(filename = "out/20230408_humans_hcorson-dosch.png", 
       width = plot_width, height = plot_height, dpi = 300)
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. Can see the dominance of different categories of water use by county. Irrigation very high percentage of water withdrawals in many western states. 

### Data source(s)

[https://waterdata.usgs.gov/nwis/wu](https://waterdata.usgs.gov/nwis/wu)

