---
title: "Global change"
author: "Katie Nuessly"
date: "2023-03-25"
output: html_document
---

## Instructions for using this template

> Make sure you have copied the template subdirectory and named it with your prompt's date, prompt's name and your name (e.g., `/01_part-to-whole_cnell`) before editing this document!

1. Put all comments either in the html text (like this) or in the `r` chunks (delineated with 3 "`" marks) to explain your process to Vizlab reviewers
2. Put all R code in the chunks. They're organized for you to (in order): do some set-up steps (load libraries), load in any data (manually or with built in packages), process your data for plotting, create plot using ggplot, and finally compose the twitter image with cowplot.
3. You can run individual code chunks while developing code and plots. Note: the plots that are created in the RStudio coding window *do not* match the plot as created through the ggsave function. 
4. To preview the final plot, use the `Knit` button above. That will create a preview of the final plot in the output `html` and will save the `png` in the `out/` folder. On many systems, the preview in the script document and the html will not match the output `png`, so please refer to the `png` for final proofing.
5. When you're happy with the final plot, fill out the supporting information at the bottom and push to gitlab for review. Note: only commit input data if it can't be downloaded directly with code (e.g., `sbtools` or `dataRetrieval`).

This website has great information on how to use RMarkdown: https://yihui.org/knitr/

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```



Define libraries here. 

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) # includes ggplot
library(readr) # best for reading and writing csvs

# These are used for the layout of the viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo
```



## Load files

Save any files that you are using directly in the `in` folder. Then read them in to the environment with this chunk of code. Remember, use the RStudio project and relative file pathways, and never include any personal computer file pathways.

This is also a good place to load any data from other built-in packages such as `dataRetrieval` or `spData`, etc.


```{r load}
# Predicted historic and current grassland biome spatial datasets were downloaded 
# from NatureServe, with use approval from the authors of Comer et al. 2018. 
# See details on how to access dataset in the `Data source(s)` section, below
# 
# Grassland area loss was calculated using the con function in 
# the raster calculator tool of ArcGIS Pro (Version 2.9.5). Grassland loss is the 
# difference in area between predicted historic and current spatial extent for 
# each of the 12 main grassland types identified in Comer et. al 2018. Zonal 
# statistics were used to extract the existing and lost grassland area by state.
data <- read.csv("in/state_long.csv")
```

## Get data ready for plotting
  
This next section is for doing any pre-processing steps, data joins, etc, to get your data ready for plotting with ggplot. Remember that most of the time, ggplot prefers data in a "wide" format.


```{r processing}
##set up figure parameters
#create empty bars for each group to space out the groups from one another in the plot
empty_bar <- 2
nObsType <- nlevels(as.factor(data$Category))
to_add <- data.frame( matrix(NA, empty_bar*nlevels(as.factor(data$Region))*nObsType, ncol(data)) )
colnames(to_add) <- colnames(data)
to_add$Region <- rep(levels(as.factor(data$Region)), each=empty_bar*nObsType )
data <- rbind(data, to_add)
data <- data %>% arrange(Region, State)
data$id <- rep( seq(1, nrow(data)/nObsType) , each=nObsType)
#generate the angle and the y position of each stacked bar label and create data frame for reference
label_data <- data %>% group_by(id, State) %>% summarize(tot=sum(Area))
number_of_bar <- nrow(label_data)
angle <- 90 - 360 * (label_data$id-0.5) /number_of_bar 
  #subtract 0.5 from label id so that the label has the angle of the center of the bar
  #(not extreme right(1) or extreme left (0))
label_data$hjust <- ifelse( angle < -90, 1, 0)
label_data$angle <- ifelse(angle < -90, angle+180, angle)
#generate the base lines ("0" y-axis line) for each stacked bar group and create data frame for reference
base_data <- data %>% 
  group_by(Region) %>% 
  summarize(start=min(id), end=max(id) - empty_bar) %>% 
  rowwise() %>% 
  mutate(title=mean(c(start, end)))
#generate y-axis break line locations and create data frame for reference
grid_data <- base_data
grid_data$end <- grid_data$end[ c( nrow(grid_data), 1:nrow(grid_data)-1)] + 1
grid_data$start <- grid_data$start - 1
grid_data <- grid_data[-1,]


```

## Set up main plot

This chunk is where the main ggplot grob definition and set-up occurs. You may have several ggplot grobs, depending on your approach. For each one, if you define it with the `<-` assignment operator and surround the whole code statement in parentheses, you'll be able to preview here what the plot looks like before the next cowplot step.

> Note: The hardest part of this process is getting text to be the optimum size in the final `png` for Twitter. The font sizes here will **not** be the same as in the final image after the cowplot step. Make sure to check the output `png` for true font sizing and plot composition. 

```{r plotting}
# Build stacked circular bar chart
(ggplot(data) +      
  #add the data (stacked bars)
  geom_bar(aes(x=as.factor(id), y=Area, fill=Category), stat="identity", alpha=1) +
  #change color of stacked bars
  scale_fill_manual(values=c("#607FBF","#F4683D")) +
  #create y-axis lines at desired data breaks
  geom_segment(data=grid_data, aes(x = end, y = 0, xend = start, yend = 0), 
               colour = "grey", alpha=1, size=1 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 50000, xend = start, yend = 50000), 
               colour = "grey", alpha=1, size=1 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 100000, xend = start, yend = 100000), 
               colour = "grey", alpha=1, size=1 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 250000, xend = start, yend = 250000), 
               colour = "grey", alpha=1, size=1 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 350000, xend = start, yend = 350000), 
               colour = "grey", alpha=1, size=1 , inherit.aes = FALSE ) +
  #label the y-axis lines
  ggplot2::annotate("text", x = rep(max(data$id),5), y = c(0, 50000, 100000, 250000, 350000),
                    label = c("0", "50,000", "100,000", "250,000", "350,000") , color="grey", size=3,
                    angle=0, hjust=1) +
  #create "0" y-axis line for each group
  geom_segment(data=base_data, aes(x = start, y = 0, xend = end, yend = 0), 
               colour = "grey", alpha=1, size=0.5, inherit.aes = FALSE )  +
  #set internal circle size to accommodate group labels (first number) 
  #and set external circle size to accommodate greatest y-axis line (second number)
  ylim(-150000,max(label_data$tot+1000, na.rm=T)) +
  #set theme
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-1,4), "cm") 
  ) +
  coord_polar() +
  #label each stacked bar
  geom_text(data=label_data, aes(x=id, y=tot+100, label=State, hjust=hjust), 
             color="grey", fontface="bold",alpha=1, size=3, angle=label_data$angle, 
             inherit.aes = FALSE ) +
  #label each group
  geom_text(data=base_data, aes(x = title, y = -100, label=Region), hjust=c(0.75,1.1,-.1,0), 
          vjust=c(2,0,0,1.5), colour = "grey", alpha=1, size=3, fontface="bold", 
          inherit.aes = FALSE))



```


## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

**Make sure to use the format for saving your png with the date of the prompt that you've been assigned to!** (e.g., `20230401_part-to-whole_cnell.png`)

```{r cowplot, fig.width = 16, fig.height = 9}
# Map generated in ArcPro
# Final plot compiled in Illustrator
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. Once spanning more than 580 million acres, the Central Grasslands are the world’s most imperiled and least conserved ecosystem. Grasslands are being lost at a far faster pace than they are being conserved due to agricultural conversion, invasive species, and climate change.

### Data source(s)

* Comer, P.J., J.C. Hak, K. Kindscher, E. Muldavin, and J. Singhurst. 2018. Continent-Scale Landscape Conservation Design for Temperate Grasslands of the Great Plains and Chihuahuan Desert. Natural Areas Journal 38(2), 196-211. https://doi.org/10.3375/043.038.0209

* NatureServe data link: [https://transfer.natureserve.org/download/Longterm/Ecosystem_Americas/Maps/](https://transfer.natureserve.org/download/Longterm/Ecosystem_Americas/Maps/).File name: `'North American Ecosystems_IVC Distributions_2021update'`. Also check out the `'READ ME: READ ME - NatureServe Macrogroups and Terrestrial Ecological Systems of the Americas'` file.
