---
title: 'Fish In Hot Water: Made for Chart Challenge'
author: "Ellie White"
date: "2023-04-03"
output:
  pdf_document: default
  html_document: default
---

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) 
library(readr) 
library(scales)
```

## Load files
Copy-pasted data from paper to a csv. Paper is here: https://afspubs.onlinelibrary.wiley.com/doi/full/10.1002/mcf2.10076

```{r load}
fish_data <- read_csv("in/fish_data.csv")
```

This extra data came from the author in personal communications 04/03/2023

I modified the spreadsheet to be more R friendly

It's a time series record spanning 1950-2099 of fish spawning onset and cessation dates (modeled values)

Maybe we can use this later to make a probability of spawning on the y axis. for now, we don't need it.

```{r load_extra_data}
fish_data_ts <- read_csv("in/fish_data_from_nack.csv")

# pulled the 1951 values as origin dates and saved to this file for easy reference
# note that the year date is set to 2015 to be consistent with the rest of the reference dates but the year in the date doesn't matter; we care about month and day. 
fish_data_ts[fish_data_ts$year == 1951, ]
fish_data_origin_dates <- read_csv("in/fish_data_origin_dates.csv")
```

## Get data ready for plotting
```{r processing}
# change to factors 
fish_data <- fish_data |>
  mutate(species = factor(species, levels = c("American Shad", "Striped Bass")), 
         variable = factor(variable, levels = c("Onset", "Cessation", "Duration")),
         period = factor(period, levels = c("Historical", "Future")))

# take out duration and confidence intervals
fish_data <- fish_data[fish_data$variable %in% c("Onset", "Cessation"), ]
fish_data_uncertainty <- select(fish_data, -c("RCP_26", "RCP_45", "RCP_60", "RCP_85")) # save this to bring back in later
fish_data <- select(fish_data, -c("RCP_26_CI", "RCP_45_CI", "RCP_60_CI", "RCP_85_CI"))

# prep origin dates 
fish_data_origin_dates <- gather(fish_data_origin_dates, condition, origin_date, RCP_26:RCP_85)
fish_data_origin_dates$origin_date <- as.Date(fish_data_origin_dates$origin_date, format = "%m/%d/%Y")

# add in origin dates
fish_data_long <- gather(fish_data, condition, value, RCP_26:RCP_85)
fish_data_long <- full_join(fish_data_long, fish_data_origin_dates, by = c("species", "variable", "condition"))
fish_data_long$end_date <- fish_data_long$origin_date + fish_data_long$value

# prep uncertainty in dates to be brought in later
fish_data_uncertainty_long <- gather(fish_data_uncertainty, condition, confidence_interval, RCP_26_CI:RCP_85_CI)
fish_data_uncertainty_long$condition <- substr(fish_data_uncertainty_long$condition, 1, 6)
```

```{r processing_extra_data, include = FALSE}
# change to factors
fish_data_ts <- fish_data_ts|>
  mutate(species = factor(species, levels = c("American Shad", "Striped Bass")),
         condition = factor(condition, levels = c("RCP_26", "RCP_45", "RCP_60", "RCP_85", labels = c("RCP 2.6", "RCP 4.5", "RCP 6.0", "RCP 8.5"))),
         period = factor(ifelse(year <= 2012, "Historical", "Future")),
         )

# fix date column, the minus one is to match the numbers Nack gave in his spreadsheet.
fish_data_ts$Onset <- as.Date(fish_data_ts$start_date_julianday, origin = as.Date("2015-01-01"))-1
fish_data_ts$Cessation <- as.Date(fish_data_ts$end_date_julianday, origin = as.Date("2015-01-01"))-1

# get rid of duration and extra dates
fish_data_ranked <- fish_data_ts |>      # use this to calculate probabilities
  select(-c("duration", "start_date", "end_date"))

fish_data_ts <- fish_data_ts |>
  select(-c("duration", "start_date_julianday", "end_date_julianday", "start_date", "end_date"))

# make long format
fish_data_ts_long <- gather(fish_data_ts, spawning, value, Onset:Cessation)
fish_data_ts_long$spawning <- factor(fish_data_ts_long$spawning, levels = c("Onset", "Cessation"))
```

## Set up main plot
Theme:
```{r plotting_setup}
theme_usgs <- function(legend.position = "right"){
  theme(
    plot.title = element_text(vjust = 3, size = 14, face = "bold", family="sans"),
    plot.subtitle = element_text(vjust = 3, size = 12,family="sans"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white"),
    legend.background = element_blank(),
    legend.justification=c(0, 0),
    legend.position = legend.position,
    legend.key = element_blank(),
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    axis.title.x = element_text(size = 10, family="sans"),
    axis.title.y = element_text(vjust = 1, angle = 90, size = 9, family="sans"),
    axis.text.x = element_text(size = 10, vjust = -0.25, colour = "black", 
                               family="sans", margin=margin(10,5,20,5,"pt")),
    axis.text.y = element_text(size = 10, hjust = 1, colour = "black", 
                               family="sans", margin=margin(5,10,10,5,"pt")),
    axis.ticks = element_line(colour = "black", linewidth = 0.1),
    axis.ticks.length = unit(-0.25 , "cm")
  )
}
```

Prep data for plot:
```{r plotting_add_extra_info}
# make into wide format 
fish_data_long <- fish_data_long |> 
  select(-c(origin_date, value))
fish_data_wide <- spread(fish_data_long, variable, end_date)
names(fish_data_wide)[names(fish_data_wide)=="Onset"] <- "onset"
names(fish_data_wide)[names(fish_data_wide)=="Cessation"] <- "cessation"

# bring in uncertainty that was prepped before
fish_data_uncertainty_wide <- spread(fish_data_uncertainty_long, variable, confidence_interval)
names(fish_data_uncertainty_wide)[names(fish_data_uncertainty_wide)=="Onset"] <- "onset_ci"
names(fish_data_uncertainty_wide)[names(fish_data_uncertainty_wide)=="Cessation"] <- "cessation_ci"

fish_data_wide <- full_join(fish_data_wide, fish_data_uncertainty_wide, by = c("species", "period", "condition"))

# add in y location for segment plot
y_location <- tibble(condition = c("RCP_26", "RCP_45", "RCP_60", "RCP_85"),
                     y = rep(c(1:4)*2 + 5))
fish_data_wide <- full_join(fish_data_wide, y_location, by = c("condition"))

# offset the y locations and mess with spacing for historical vs. future so they slightly overlap but don't overlap across the RCPs
fish_data_wide <- fish_data_wide |>
  mutate(y_offset = ifelse(period == "Historical", y - 0.5, y))
```

## Produce plots
```{r plotting, include = FALSE}
# First, a dot chart for my own sanity
ggplot(data = fish_data_long, aes(x = end_date, y = variable)) +
  geom_point(aes(col = condition)) + 
  facet_wrap(~species) +
  theme_usgs()
```

```{r plotting_main}
# Main base plot, trying out blurring edges
library(ggnewscale)
ggplot() +
  geom_segment(data = fish_data_wide[fish_data_wide$period == "Historical", ], aes(x = onset, xend = cessation, y = y_offset, yend = y_offset), alpha = 0.7, linewidth = 8 , show.legend = FALSE, col = "grey") +
  geom_segment(data = fish_data_wide[fish_data_wide$period == "Historical", ], aes(x = onset-onset_ci, xend = cessation + cessation_ci, y = y_offset, yend = y_offset), alpha = 0.25, linewidth = 8 , show.legend = FALSE, col = "grey") +
  new_scale_colour() +
  geom_segment(data = fish_data_wide[fish_data_wide$period == "Future", ], aes(x = onset, xend = cessation, y = y_offset, yend = y_offset), alpha = 0.7, linewidth = 8 , show.legend = FALSE, col = "#815aa5") +
  geom_segment(data = fish_data_wide[fish_data_wide$period == "Future", ], aes(x = onset-onset_ci, xend = cessation + cessation_ci, y = y_offset, yend = y_offset), alpha = 0.25, linewidth = 8 , show.legend = FALSE, col = "#815aa5") +
  scale_x_date(limits = c(as.Date("2015-01-01"), as.Date("2015-12-31")), date_breaks = "1 month", date_minor_breaks = "1 week", date_labels = "%B") +
  scale_y_continuous(limits = c(0, 13)) +
  coord_polar(theta = "x", direction = 1, start = -1.57*1.5) + # start is in radians, 90 Deg is Jan
  facet_wrap(~species) +
  labs(x = "",
       y = "",
       title = "FISH IN HOT WATER",
       subtitle = "Under projected climate change scenarios, the American Shad and Striped Bass of the Hudson River Estuary are predicted to spawn 15 days earlier on average."
       # caption = "Data Source: Nack, C. et. al. (2019). https://doi.org/10.1002/mcf2.10076
       # Plot made by Ellie White, ewhite@usgs.gov 04/02/2023"
       ) +
  theme_bw()+
  theme(plot.title = element_text(vjust = 3, size = 14, face = "bold", family="sans"),
        plot.subtitle = element_text(vjust = 3, size = 12,family="sans"),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.border = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank())

ggsave("out/11_circular_ewhite_base.png", width = 16, height = 9, units = "in", dpi = 1200)
```

```{r checking_original_data_with_extra_data, include = FALSE}
# recreating figure in paper for sanity
ggplot(data = fish_data_ts_long[fish_data_ts_long$year > 1950, ], aes(x = year, y = value)) +
  geom_point(aes(shape = species, col = species)) +
  facet_wrap(~condition + spawning, ncol = 2, scales = "free") +
  labs(x="", y="Predicted Date of Spawning") +
  theme_usgs()
# it matches! Hooray


# is the mean what is reported for onset and cessation deviations in the linear modeling? 
fish_data_ts_summarized <- fish_data_ts_long |>
  group_by(species, condition, period, spawning) |>
  summarize(mean_value = mean(value), 
            se_value = sd(value)/sqrt(length(value))) # calculate standard error
  
fish_data_summarized <- full_join(fish_data_long, fish_data_ts_summarized, by = join_by(species == species, period == period, variable == spawning, condition == condition)) |>
  mutate(date_diff = end_date - mean_value)

# It's not the mean value of the data, it's slightly off!
fish_data_summarized_summarized <- fish_data_summarized |>
  group_by(period, variable) |>
  summarize(mean(mean_value))

# cessation 
fish_data_summarized_summarized$`mean(mean_value)`[1] - fish_data_summarized_summarized$`mean(mean_value)`[3]

# onset
fish_data_summarized_summarized$`mean(mean_value)`[2] - fish_data_summarized_summarized$`mean(mean_value)`[4]
```

Let's make a gif with the extra data instead?
```{r gif}
# make the yearly data wide so onset and cessation columns are side by side
fish_data_ts_wide <- spread(fish_data_ts_long, spawning, value)

# filter by RCP
fish_data_ts_plot <- fish_data_ts_wide |>
  filter(condition == "RCP_85") |>
  filter(year != 1950) |>
  mutate(year = as.integer(year))

library(gganimate)
library(scico)
animation <- ggplot(data = fish_data_ts_plot) + 
  geom_segment(aes(x = Onset, xend = Cessation, y = year, yend = year, col = Onset), linewidth = 1 , show.legend = FALSE) + 
  scale_color_scico(palette = "tokyo") +
  scale_x_date(limits = c(as.Date("2015-01-01"), as.Date("2015-12-31")), date_breaks = "1 month", date_minor_breaks = "1 week", date_labels = "%B") +  
  scale_y_continuous(limits = c(1900, 2120)) +
  coord_polar(theta = "x", direction = 1, start = -1.57*1.5) + # start is in radians, 90 Deg is Jan
  facet_wrap(~species) +
  labs(x = "", 
       y = "", 
       title = "FISH IN HOT WATER", 
       subtitle = "Under projected the climate change scenario RCP 8.5, the American Shad and Striped Bass of the Hudson River Estuary are predicted to spawn XX days earlier on average.
       Year: {frame_time}
       "
       # caption = "Data Source: Nack, C. et. al. (2019). https://doi.org/10.1002/mcf2.10076
       # Plot made by Ellie White, ewhite@usgs.gov 04/02/2023"
       ) +
  theme_bw()+
  theme(plot.title = element_text(vjust = 3, size = 14, face = "bold", family="sans"),
        plot.subtitle = element_text(vjust = 3, size = 12,family="sans"), 
        # axis.text.y = element_blank(), 
        axis.ticks = element_blank(), 
        panel.border = element_blank(), 
        strip.background = element_blank(), 
        strip.text.x = element_blank()) +
  # Here comes the gganimate specific bits
  transition_time(year) +
  shadow_mark(past = TRUE, future = FALSE, alpha = 0.3) 

anim_save("out/11_circular_ewhite_animated.gif", animation, width = 16, height = 9, units = "in", res = 900)
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. The American Shad and the Striped Bass are migratory species needing both freshwater and marine habitats to complete their life cycle. This makes them particularly vulnerable to human activities. The Hudson River Shad has declined in stock so much that all its fisheries were closed in 2010. The Striped Bass, while declining in relative abundance, still remains the most important game fish in the Hudson River.


### Data source(s)

Paper is here: https://afspubs.onlinelibrary.wiley.com/doi/full/10.1002/mcf2.10076

Citation: Nack, C. C., Swaney, D. P., & Limburg, K. E. (2019). Historical and projected changes in spawning Phenologies of American Shad and Striped bass in the 
Hudson River Estuary. Marine and Coastal Fisheries, 11(3), 271-284.

DOI: https://doi.org/10.1002/mcf2.10076

### Process 

1) produced `out/[day]_ewhite_base.png` with ggplot

2) made markups in powerpoint 

3) final plot is called `out/[day]_ewhite_final.png`




