---
title: "OWID chart"
author: "Hayley Corson-Dosch"
date: "2023-03-29"
output: html_document
---

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

Define libraries here. 

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) # includes ggplot
library(readr) # best for reading and writing csvs
library(ggalluvial)
library(MetBrewer)

# These are used for the layout of the viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo
```

## Load files

```{r load, warning=FALSE}
# Add your code here
natural_disasters <- readr::read_csv('in/natural-disasters.csv', col_types = cols())
```

## Get data ready for plotting

```{r processing}
# Munge US data here
us_natural_disaster_deaths <- natural_disasters %>%
  filter(`Country name` == 'United States') %>% # filter to US
  dplyr::select(country=`Country name`, year = Year, contains('Number of deaths')) %>% # Select death totals
  dplyr::select(where(~sum(!is.na(.x)) > 0)) %>% # drop columns w/ NA values for all years
  pivot_longer(starts_with('Number of deaths'), names_to = 'natural_hazard', values_to = 'deaths',
               names_pattern = 'Number of deaths from (.*)') %>% # pivot to long format
  filter(!(natural_hazard == 'disasters')) %>% # filter out the category for all disasters
  mutate(date = as.Date(sprintf('%s-01-01', year)), # add date column
         deaths = ifelse(is.na(deaths), 0, deaths)) %>% # set deaths to 0 if NA
  filter(date > as.Date('1971-12-31')) # filter to last 50 years

```

## Set up main plot

```{r plotting}
# Load some custom fonts and set some custom settings
annotation_font <- "Shadows Into Light"
sysfonts::font_add_google(annotation_font)
annotation_font_size <- 22
supporting_font <- "Source Sans Pro"
sysfonts::font_add_google(supporting_font)
showtext::showtext_opts(dpi = 300, regular.wt = 200, bold.wt = 700)
showtext::showtext_auto(enable = TRUE)
plot_font_size <- 16

# Define colors
background_color <- "#ffffff"
font_color <- "#6E6E6E"
annotation_font_color <- "#1A1A1A"
title_font_color <- "#4F4F4F"
colors <- MetBrewer::met.brewer('Tiepolo', 8)
names(colors) <- c('wildfires', 'drought', 'extreme temperatures', 'landslides', 'volcanic activity', 'earthquakes', 'storms', 'floods')

# order data by color order
us_natural_disaster_deaths <- mutate(us_natural_disaster_deaths, natural_hazard = factor(natural_hazard, levels = names(colors)))

# The background canvas for your viz
canvas <- grid::rectGrob(
  x = 0, y = 0, 
  width = 16, height = 9,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color)
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_black.png") 

# Build main plot
(main_plot <- ggplot(us_natural_disaster_deaths) +
  geom_alluvium(aes(date, deaths, fill = natural_hazard, alluvium = natural_hazard), decreasing = FALSE, alpha = 0.7) +
  scale_x_date(
    breaks = scales::date_breaks(width = '5 years'),
    labels = scales::label_date_short(),
    expand = c(0,0),
    name = '') +
  scale_fill_manual(values = colors,
                    name  = '') +
  scale_y_continuous(name = '',
                     limits = c(0, 2100),
                     breaks = c(500, 1000, 1500, 2000), 
                     labels = c(500, 1000, 1500, stringr::str_wrap('2000 deaths', 10)),
                     expand = c(0,0)) +
  theme(axis.text.y = element_text(hjust = 1, color = font_color, size = plot_font_size, family = supporting_font),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(color = font_color, size = plot_font_size, family = supporting_font),
        plot.background = element_blank(),
        panel.background = element_blank(),
        panel.grid.major.y = element_line(color = 'grey70', linewidth = 0.25, linetype = 'dotted'),
        panel.grid.major.x = element_blank(),
        legend.position = c(1.29, 0.71),
        legend.text = element_text(color = font_color, family = supporting_font, size = plot_font_size, margin = margin(r = 17)),
        legend.spacing.y = unit(3.2, 'cm'),
        legend.spacing.x = unit(0.18, 'cm'),
        legend.key.width = unit(0.5, 'cm'),
        legend.key.height = unit(0.2, 'cm'),
        legend.background = element_blank()) +
  guides(fill = guide_legend(nrow=4, 
                             byrow = TRUE))
 )

# build sub plots
plot_list <- purrr::map(seq(1, length(colors)), function(hazard_id) {
  factor_levels <- c(names(colors)[names(colors) !=  names(colors)[hazard_id]], names(colors)[hazard_id])

  us_natural_disaster_deaths <- mutate(us_natural_disaster_deaths, natural_hazard = factor(natural_hazard, levels = factor_levels))

  plot_colors <- rep(NA, length(colors))
  plot_colors[length(plot_colors)] <- colors[hazard_id]

  ggplot(us_natural_disaster_deaths) +
    geom_alluvium(aes(date, deaths, fill = natural_hazard, alluvium = natural_hazard), decreasing = FALSE, alpha = 0.8) +
    scale_x_date(
      breaks = scales::date_breaks(width = '5 years'),
      labels = scales::label_date_short(),
      expand = c(0,0),
      name = '') +
    scale_fill_manual(values = plot_colors, name  = '', na.value = 'grey90') +
    scale_y_continuous(name = '',
                       limits = c(0, 2100),
                       breaks = c(500, 1000, 1500, 2000),
                       labels = c(500, 1000, 1500, stringr::str_wrap('2000 deaths', 10)),
                       expand = c(0,0)) +
    theme_void() +
    theme(legend.position = 'None',
          plot.margin = unit(c(0.3,0.3,0.3,0.3), "cm"))
})

```


## Produce final plot

```{r cowplot, fig.width = 16, fig.height = 9, warning=FALSE}
white_arrow_linewidth <- 1.3
grey_arrow_linewidth <- 0.7
arrow_length <- 0.75

ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 9, width = 16,
            hjust = 0, vjust = 1) +
  # small multiples
  draw_plot(plot_grid(plotlist = plot_list, nrow = 4),
            x = 0.64,
            y = 0.190,
            width = 0.325,
            height = 0.66) +
  # the main plot
  draw_plot(main_plot,
            x = 0.028,
            y = 0.15,
            height = 0.7,
            width = 0.6) +
  # annotations
  # heat wave
draw_label("1980 heat wave",
             x = 0.195,
             y = 0.57,
             hjust = 0,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = annotation_font,
             color = annotation_font_color,
             size = annotation_font_size) +
  geom_curve(aes(xend = 0.184, x = 0.212,
                 yend = 0.52, y = 0.555),
             arrow = grid::arrow(length = unit(arrow_length, 'lines')),
             curvature = -0.35,
             angle = -25,
             color = title_font_color,
             lineend = 'round',
             linewidth = grey_arrow_linewidth) +
  # mount st helens
  draw_label(stringr::str_wrap("Mount St. Helens eruption", 20),
             x = 0.195,
             y = 0.42,
             hjust = 0,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = annotation_font,
             color = annotation_font_color,
             size = annotation_font_size) +
  geom_curve(aes(xend = 0.174, x = 0.188,
                 yend = 0.255, y = 0.435),
             arrow = grid::arrow(length = unit(arrow_length, 'lines')), 
             curvature = 0.22,
             angle = -25,
             color = background_color,
             lineend = 'round',
             linewidth = white_arrow_linewidth) +
  geom_curve(aes(xend = 0.174, x = 0.188,
                 yend = 0.255, y = 0.435),
             arrow = grid::arrow(length = unit(arrow_length, 'lines')), 
             curvature = 0.22,
             angle = -25,
             color = title_font_color,
             lineend = 'round',
             linewidth = grey_arrow_linewidth) +
  # hurricane katrina
  draw_label("Hurricane Katrina",
             x = 0.42,
             y = 0.7,
             hjust = 1,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = annotation_font,
             color = annotation_font_color,
             size = annotation_font_size) +
  geom_curve(aes(xend = 0.43, x = 0.4,
                 yend = 0.62, y = 0.685),
             arrow = grid::arrow(length = unit(arrow_length, 'lines')), 
             curvature = 0.25,
             angle = -55,
             color = title_font_color,
             lineend = 'round',
             linewidth = grey_arrow_linewidth) +
  # Oso mudslide
  draw_label("2014 Oso mudslide",
             x = 0.47,
             y = 0.47,
             hjust = 0,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = annotation_font,
             color = annotation_font_color,
             size = annotation_font_size) +
   geom_curve(aes(xend = 0.532, x = 0.485,
                 yend = 0.225, y = 0.453),
             arrow = grid::arrow(length = unit(arrow_length, 'lines')), 
             curvature = 0.35,
             angle = -25,
             color = background_color,
             lineend = 'round',
             linewidth = white_arrow_linewidth) +
  geom_curve(aes(xend = 0.532, x = 0.485,
                 yend = 0.225, y = 0.453),
             arrow = grid::arrow(length = unit(arrow_length, 'lines')), 
             curvature = 0.35,
             angle = -25,
             color = title_font_color,
             lineend = 'round',
             linewidth = grey_arrow_linewidth) +
  # Camp fire
  draw_label(paste("2018", "Camp fire", sep = '\n'),
             x = 0.52,
             y = 0.325,
             hjust = 0,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = annotation_font,
             color = annotation_font_color,
             size = annotation_font_size) +
  geom_curve(aes(xend = 0.571, x = 0.54,
                 yend = 0.24, y = 0.307),
             arrow = grid::arrow(length = unit(arrow_length, 'lines')), 
             curvature = 0.3,
             angle = -45,
             color = background_color,
             lineend = 'round',
             linewidth = white_arrow_linewidth) +
  geom_curve(aes(xend = 0.571, x = 0.54,
                 yend = 0.24, y = 0.307),
             arrow = grid::arrow(length = unit(arrow_length, 'lines')), 
             curvature = 0.3,
             angle = -45,
             color = title_font_color,
             lineend = 'round',
             linewidth = grey_arrow_linewidth) +
  # explainer text
  draw_label("Data: https://ourworldindata.org/natural-disasters",
             fontfamily = supporting_font,
             x = 0.96,   
             y = 0.05,
             size = 14,
             hjust = 1,
             vjust = 0,
             color = font_color)+
  # Title
  draw_label("Lives lost in U.S. natural disasters",
             x = 0.04,
             y = 0.95,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = supporting_font,
             fontface = 'bold',
             color = title_font_color,
             size = 45) +
  # Add logo
  draw_image(usgs_logo, 
             x = 0.04,
             y = 0.05,
             width = 0.1, 
             hjust = 0, vjust = 0, 
             halign = 0, valign = 0)

# Save the final image in Twitter's 16 by 9 format
# !! Use format for saving with the date of your prompt: 
#         YYYYMMDD_prompt_name ()
# e.g. `20230101_part-to-whole-cnell.png`
ggsave(filename = "out/20230406_OWID_hcorson-dosch.png", 
       width = 16, height = 9, dpi = 300)
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. Key historic events stand out in OWID record of deaths from U.S. natural hazards - 1977 floods in Kentucky and Pennsylvania, 1980 heat wave, Mt. Saint Helens eruption in 1980, Hurricane Katrina 2005, 2014 Oso Mudslide, 2018 Camp Fire. 

### Data source(s)

[Our World in Data (OWID) natural disasters data](https://ourworldindata.org/natural-disasters)

