---
title: "Example for Chart Challenge"
author: "Althea Archer"
date: "2023-03-10"
output: html_document
---

## Instructions for using this template



## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```



Define libraries here. 

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) # includes ggplot
library(readr) # best for reading and writing csvs

library(sbtools)
library(lubridate)
library(gganimate)
library(hms)

# These are used for the layout of the viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo
```



## Load files

Save any files that you are using directly in the `in` folder. Then read them in to the environment with this chunk of code. Remember, use the RStudio project and relative file pathways, and never include any personal computer file pathways.

This is also a good place to load any data from other built-in packages such as `dataRetrieval` or `spData`, etc.


```{r load}
# Load drought properties data from ScienceBase
#sbtools::authenticate_sb()

file_in_path_discrete <- "in/EcoDrought Discrete_VA.csv"
if(!file.exists(file_in_path_discrete)){ # if files don't exist, download
  sbtools::item_file_download(sb_id = "624decacd34e21f82766c9da",
                              names = "EcoDrought Discrete_VA.csv",
                              destinations =  "in/EcoDrought Discrete_VA.csv",
                              overwrite_file = F)
}
discrete <- readr::read_csv(file_in_path_discrete)

file_in_path_continuous <- "in/EcoDrought Continuous_VA.csv"
if(!file.exists(file_in_path_continuous)){ # if files don't exist, download
  sbtools::item_file_download(sb_id = "624decacd34e21f82766c9da",
                              names = "EcoDrought Continuous_VA.csv",
                              destinations =  "in/EcoDrought Continuous_VA.csv",
                              overwrite_file = F)
}
continuous <- readr::read_csv(file_in_path_continuous)

metadata_in_path <- "in/VA_site_info.csv"
if(!file.exists(metadata_in_path)){ # if files don't exist, download
  sbtools::item_file_download(sb_id = "624decacd34e21f82766c9da",
                              names = "VA_site_info.csv",
                              destinations =  "in/VA_site_info.csv",
                              overwrite_file = F)
}
metadata <- readr::read_csv(metadata_in_path)

```

## Get data ready for plotting
  
This next section is for doing any pre-processing steps, data joins, etc, to get your data ready for plotting with ggplot. Remember that most of the time, ggplot prefers data in a "wide" format.


```{r processing}
# Isolate example of hysteresis
hyst_example <- continuous |> 
  filter(year(DateTime_EST) == 2019,
         yday(DateTime_EST) == 117,
         !is.na(AirTemperature_HOBO_degF),
         !is.na(WaterTemperature_HOBO_DegF)) |>
  rename(dt = DateTime_EST,
         waterT = WaterTemperature_HOBO_DegF,
         airT = AirTemperature_HOBO_degF) |>
  mutate(time = hms::as_hms(dt),
         am_logic = am(time),
         daytime = case_when(time %in% as_hms(3600*seq(5.5, 16, by = 0.5)) ~ TRUE,
                             TRUE ~ FALSE)) |>
  arrange(time) |>
  left_join(metadata, by = "Site_ID")

hyst_exampleSummer <- continuous |> 
  filter(year(DateTime_EST) == 2021,
         yday(DateTime_EST) == 241,
         !is.na(AirTemperature_HOBO_degF),
         !is.na(WaterTemperature_HOBO_DegF)) |>
  rename(dt = DateTime_EST,
         waterT = WaterTemperature_HOBO_DegF,
         airT = AirTemperature_HOBO_degF) |>
  mutate(time = hms::as_hms(dt),
         am_logic = am(time),
         daytime = case_when(time %in% as_hms(3600*seq(5.5, 16, by = 0.5)) ~ TRUE,
                             TRUE ~ FALSE)) |>
  arrange(time) |>
  left_join(metadata, by = "Site_ID")


whole_month <- continuous |> 
  filter(year(DateTime_EST) == 2019,
         month(DateTime_EST) == 4,
         !is.na(AirTemperature_HOBO_degF),
         !is.na(WaterTemperature_HOBO_DegF)) |>
  rename(dt = DateTime_EST,
         waterT = WaterTemperature_HOBO_DegF,
         airT = AirTemperature_HOBO_degF) |>
  mutate(time = hms::as_hms(dt),
         am_logic = am(time)) |>
  arrange(dt) |>
  left_join(metadata, by = "Site_ID")

full_record <- continuous |> 
  filter(!is.na(AirTemperature_HOBO_degF),
         !is.na(WaterTemperature_HOBO_DegF)) |>
  rename(dt = DateTime_EST,
         waterT = WaterTemperature_HOBO_DegF,
         airT = AirTemperature_HOBO_degF) |>
  mutate(time = hms::as_hms(dt),
         am_logic = am(time)) |>
  arrange(dt) |>
  left_join(metadata, by = "Site_ID")

test2021 <- full_record |> filter(year(dt) == 2021)
ggplot(test2021 |> filter(month(dt) == 4), aes(x = airT, y = waterT, group = yday(dt))) + geom_path(linewidth = 0.2) + facet_wrap(~ yday(dt)) + coord_fixed(ratio = 1)#+ theme_void()

ggplot(hyst_exampleSummer, aes(x = airT, y = waterT, group = yday(dt))) + geom_point() + facet_wrap(~ yday(dt)) #+ theme_void()

```

## Set up main plot

This chunk is where the main ggplot grob definition and set-up occurs. You may have several ggplot grobs, depending on your approach. For each one, if you define it with the `<-` assignment operator and surround the whole code statement in parentheses, you'll be able to preview here what the plot looks like before the next cowplot step.

> Note: The hardest part of this process is getting text to be the optimum size in the final `png` for Twitter. The font sizes here will **not** be the same as in the final image after the cowplot step. Make sure to check the output `png` for true font sizing and plot composition. 

```{r plotting}
# Load some custom fonts and set some custom settings
font_legend <- "Share Tech Mono"  #"Special Elite"
sysfonts::font_add_google("Share Tech Mono")
supporting_font <- "Source Sans Pro"
sysfonts::font_add_google("Source Sans Pro")
showtext::showtext_opts(dpi = 300, regular.wt = 200, bold.wt = 700)
showtext::showtext_auto(enable = TRUE)

# Define colors
background_color = "#dddfd4"
font_color = "#173e43"
pop_color = "#3fb0ac"
highlight_color = "#fae596"
point_size = 7
inset_point_size = 4

# The background canvas for your viz
canvas <- grid::rectGrob(
  x = 0, y = 0, 
  width = 16, height = 9,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color)
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_black.png") |>
  magick::image_colorize(100, font_color)

# Main plot
(main_plot <- ggplot(hyst_example, aes(x = airT, y = waterT)) +
    geom_smooth(method = "lm", se = FALSE, color = "white",
                linetype = "dashed")+
    geom_point(color = pop_color, size = point_size)+
    geom_point(color = "white", size = inset_point_size, 
               data = hyst_example |> filter(daytime))+
    geom_point(color = highlight_color, size = inset_point_size, 
               data = hyst_example |> filter(time %in% c(hms::as_hms(3600*6),
                                                         hms::as_hms(3600*11))))+
    theme_minimal()+
    ylab("Water Temperature (F)\n")+
    xlab("\nAir Temperature (F)")+
    ylim(c(40, 58))+
    xlim(c(44, 68))+
    geom_point(data = data.frame(airT = c(45, 45), waterT = c(54, 54.5)),
              color = pop_color, size = point_size)+
    geom_point(data = data.frame(airT = 45, waterT = 54.5),
              color = "white", size = inset_point_size)+
    geom_text(label = "day", y = 54.5, x = 45.5, hjust = -0,
             color = pop_color, size = 5)+
    geom_text(label = "6 a.m.", y = 48.9, x = 45.5, hjust = -0,
             color = font_color, size = 5)+
    geom_text(label = "night", y = 54, x = 45.5, hjust = -0,
              color = pop_color, size = 5)+
    geom_text(label = "11 a.m.", y = 50, x = 56, hjust = -0,
             color = font_color, size = 5)+
    theme(panel.background = element_blank(),
          axis.text = element_text(size = 20, color = font_color),
          axis.title = element_text(size = 20, color = font_color))+
    coord_fixed(ratio = 1)+
    scale_x_continuous(breaks = seq(44, 68, by = 2))+
    scale_y_continuous(breaks = seq(40, 58, by = 2))
 )
```



```{r animation}
(animated_plot <- ggplot(hyst_example, aes(x = airT, y = waterT)) +
    geom_point(color = pop_color, size = point_size)+
    theme_minimal()+
    transition_states(time)+
    shadow_mark(alpha = 0.5, size = inset_point_size, color = pop_color)+
    labs(title = "April 27, 2019 {closest_state}")+
    ylab("Water Temperature (F)")+
    xlab("Air Temperature (F)")+
    theme(plot.background = element_rect(fill = "white", color = NA),
          axis.text = element_text(size = 18, color = font_color),
          axis.title = element_text(size = 18, color = font_color, face = "bold"),
          plot.title = element_text(size = 24, family = font_legend))+
    coord_fixed(ratio = 1)+
    scale_x_continuous(breaks = seq(44, 68, by = 2))+
    scale_y_continuous(breaks = seq(40, 58, by = 2))
 )
animation_to_save <- animate(animated_plot,
          width = 1600, height = 800, res = 200)  + exit_shrink()
anim_save("out/20230420_correlation_aaarcher.gif", bg = background_color)

```


## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

**Make sure to use the format for saving your png with the date of the prompt that you've been assigned to!** (e.g., `20230401_part-to-whole_cnell.png`)

```{r cowplot, fig.width = 16, fig.height = 9}
ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 9, width = 16,
            hjust = 0, vjust = 1) +
  # the main plot
  draw_plot(main_plot,
            x = 0.034,
            y = -0.05,
            width = 0.95) +
  # explainer text
  draw_label("Chart by Althea A. Archer.\nApril 27, 2019 data from Paine Run, Shenandoah National Park.\nData from https://doi.org/10.5066/P97ULKSM",
             fontfamily = supporting_font,
             x = 0.96,   
             y = 0.04,
             size = 20,
             hjust = 1,
             vjust = 0,
             color = font_color)+
  # Title
  draw_label("Hysteresis",
             x = 0.04,
             y = 0.95,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = font_color,
             size = 60) +
  # Title
  draw_label("(noun) - A phenomenon in which the value of a physical property\n  lags behind changes in the effect causing it.",
             x = 0.10,
             y = 0.86,
             hjust = 0,
             vjust = 1,
             lineheight = 0.85,
             fontfamily = font_legend,
             color = font_color,
             size = 24) +
  # draw_label("In the morning, it takes longer\nfor the stream to warm than the air.",
  #            x = 0.56,
  #            y = 0.375,
  #            hjust = 0,
  #            vjust = 1,
  #            lineheight = 0.75,
  #            fontfamily = font_legend,
  #            color = font_color,
  #            size = 24)+
  # Add logo
  draw_image(usgs_logo, 
             x = 0.04,
             y = 0.04,
             width = 0.12, 
             hjust = 0, vjust = 0, 
             halign = 0, valign = 0)

# Save the final image in Twitter's 16 by 9 format
# !! Use format for saving with the date of your prompt: 
#         YYYYMMDD_prompt_name ()
# e.g. `20230101_part-to-whole-cnell.png`
ggsave(filename = "out/20230420_correlation_aaarcher.png", 
       width = 16, height = 9, dpi = 300)
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. Key takeaways here

### Data source(s)

Data sources here.

