---
title: "Not your parents' baselines"
author: "Althea Archer"
date: "2023-03-16"
output: html_document
---

## Instructions for using this template

1. Put all comments either in the html text (like this) or in the `r` chunks (delineated with 3 "`" marks) to explain your process to Vizlab reviewers
2. Put all R code in the chunks. They're organized for you to (in order): do some set-up steps (load libraries), load in any data (manually or with built in packages), process your data for plotting, create plot using ggplot, and finally compose the twitter image with cowplot.
3. You can run individual code chunks while developing code and plots. Note: the plots that are created in the RStudio coding window *do not* match the plot as created through the ggsave function. 
4. To preview the final plot, use the `Knit` button above. That will create a preview of the final plot in the output `html` and will save the `png` in the `out/` folder. On many systems, the preview in the script document and the html will not match the output `png`, so please refer to the `png` for final proofing.
5. When you're happy with the final plot, fill out the supporting information at the bottom and push to gitlab for review. Note: only commit input data if it can't be downloaded directly with code (e.g., `sbtools` or `dataRetrieval`).

This website has great information on how to use RMarkdown: https://yihui.org/knitr/

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```



Define libraries here. 

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) # includes ggplot
library(readr) # best for reading and writing csvs

library(sbtools)
library(sf)
library(readr)
library(scales)
library(zoo)

# These are used for the layout of the viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo
```



## Load files

Save any files that you are using directly in the `in` folder. Then read them in to the environment with this chunk of code. Remember, use the RStudio project and relative file pathways, and never include any personal computer file pathways.

This is also a good place to load any data from other built-in packages such as `dataRetrieval` or `spData`, etc.


```{r load}
# Load drought properties data from ScienceBase
#authenticate_sb()
# 1921-2020
file_in_1921 <- "in/Streamflow_percentiles_national_1921/01031500.csv"
if(!file.exists(file_in_1921)){ # if files don't exist, download
  sbtools::item_file_download(sb_id = "62f463c5d34eacf5397395d9",
                              names = "Streamflow_percentiles_national_1921.zip",
                              destinations = file_in_1921,
                              overwrite_file = F)
}

file_in_metadata <- "in/study_watersheds_metadata.csv"
if(!file.exists(file_in_metadata)){ # if files don't exist, download
  sbtools::item_file_download(sb_id = "62793493d34e8d45aa6e3ba9",
                              names = "study_watersheds_metadata.csv",
                              destinations = file_in_metadata,
                              overwrite_file = F)
}
metadata <- readr::read_csv(file_in_metadata, show_col_types = FALSE) |>
  select(StaID, STANAME, DRAIN_SQKM, HUC02, LAT_GAGE, LNG_GAGE, STATE,
         national_1981, national_1951, national_1921, HCDN_2009)

# Download US State boundaries as sf object
states_sf <- spData::us_states
# add in state names, too
st_crosswalk <- data_frame(NAME = state.name,
                           STATE = state.abb)
# Merge in CASC info
states_CASC <- states_sf |>
  mutate("CASC" = case_when(NAME %in% c("Minnesota", "Iowa", "Missouri", 
                                       "Wisconsin", "Illinois", "Indiana", 
                                       "Michigan", "Ohio") ~ "Midwest",
                          NAME %in% c("Montana", "Wyoming", "Colorado", "North Dakota", 
                                       "South Dakota", "Nebraska", "Kansas") ~ "North Central",
                          NAME %in% c("Maine", "New Hampshire", "Vermont", "Massachusetts", 
                                       "Connecticut", "Rhode Island",
                                       "New York", "New Jersey", "Pennsylvania", 
                                       "Delaware", "Maryland", "West Virginia", 
                                       "Virginia", "Kentucky", "District of Columbia") ~ "Northeast",
                          NAME %in% c("Washington", "Oregon", "Idaho") ~ "Northwest",
                          NAME %in% c("New Mexico", "Texas", 
                                      "Oklahoma", "Louisiana") ~ "South Central",
                          NAME %in% c("North Carolina", "South Carolina", "Georgia", "Alabama", 
                                       "Mississippi", "Florida", "Tennessee", 
                                       "Arkansas") ~ "Southeast",
                          NAME %in% c("Arizona",
                                       "Utah", "Nevada") ~ "Southwest",
                          NAME == "California" ~ "California",
                          TRUE ~ "not sorted")) |>
  left_join(st_crosswalk, by = "NAME")

```

## Get data ready for plotting
  
This next section is for doing any pre-processing steps, data joins, etc, to get your data ready for plotting with ggplot. Remember that most of the time, ggplot prefers data in a "wide" format.


```{r processing}
# Determine what sites are in both datasets using metadata
metadata <- metadata |>
  mutate(in_both_L = case_when(national_1981 & national_1921 ~ TRUE,
                               TRUE ~ FALSE)) |>
  # add in Climate area 
  mutate("CASC" = case_when(STATE %in% c("MN", "IA", "MO", "WI", "IL", "IN", "MI", "OH") ~ "Midwest",
                          STATE %in% c("MT", "WY", "CO", "ND", "SD", "NE", "KS") ~ "North Central",
                          STATE %in% c("ME", "NH", "VT", "MA", "CT", "RI", "DC",
                                       "NY", "NJ", "PA", "DE", "MD", "WV", "VA", "KY") ~ "Northeast",
                          STATE %in% c("WA", "OR", "ID") ~ "Northwest",
                          STATE %in% c("NM", "TX", "OK", "LA") ~ "South Central",
                          STATE %in% c("NC", "SC", "GA", "AL", "MS", "FL", "TN", "AR") ~ "Southeast",
                          STATE %in% c("AZ", "UT", "NV") ~ "Southwest",
                          STATE == "CA" ~ "California",
                          TRUE ~ "not sorted"))

# Determine focal sites (these 331 sites have data for 100 years)
focal_sites <- metadata$StaID[metadata$in_both_L]

# Read in all focal sites and rowbind together for each timeframe
files_1921 <- sprintf("in/Streamflow_percentiles_national_1921/%s.csv", focal_sites)
t1921_data <- purrr::map(files_1921[1], 
                         ~ readr::read_csv(files_1921, 
                                           show_col_types = FALSE,
                                           col_types = list("StaID" = col_character()))) |>

  reduce(rbind) |>
  # only keep fields we need (value = mean daily streamflow in cfs)
  select(StaID, dt, year, value)



# Calculate EPA's climate change metrics for each station
#  (https://www.epa.gov/climate-indicators/climate-change-indicators-streamflow)
climate_metrics_site <- t1921_data |> 
  filter(year %in% 1920:2019) |>
  group_by(StaID) |>
  # calculate 7-day minimum streamflow by site (returns Inf for empty sets)
  mutate(min_flow_7d = zoo::rollapply(value, width = 7, 
                                      function(x){min(x, na.rm = TRUE)}, fill = NA)) |>
  # convert infinities to missing values
  mutate(min_flow_7d = case_when(min_flow_7d %in% c(Inf, -Inf) ~ NA,
                                 TRUE ~ min_flow_7d))

# Extract MIN values by station and year to get "Annual Minimum Streamflow"
climate_metrics_year_site <- climate_metrics_site |>
  group_by(year, StaID) |> 
  summarise(minflow_7d = min(min_flow_7d, na.rm = TRUE)) |>
  # convert infinities to missing values
  mutate(minflow_7d = case_when(minflow_7d %in% c(Inf, -Inf) ~ NA,
                                 TRUE ~ minflow_7d))


# Add in generations 
climate_metrics_year_site_join <- climate_metrics_year_site  |>
  mutate(generation = case_when(year %in% 1920:1945 ~ "0 silent generation",
                                year %in% 1946:1964 ~ "1 baby boomers",
                                year %in% 1965:1980 ~ "2 gen x",
                                year %in% 1981:1996 ~ "3 millenials",
                                year %in% 1997:2021 ~ "4 gen z"))  

# Get average values by station and generation
climate_metrics_generation_site_temp <- climate_metrics_year_site_join |>
  group_by(generation, StaID) |> 
  summarise(minflow_7d = mean(minflow_7d, na.rm = TRUE)) 

# Remove sites that had 0 flow in 1920s (three sites)
climate_metrics_generation_site_remove <- climate_metrics_generation_site_temp |>
  filter(generation == "0 silent generation",
         minflow_7d == 0)
climate_metrics_generation_site <- climate_metrics_generation_site_temp |>
  filter(! StaID %in% climate_metrics_generation_site_remove$StaID)

# Calculate average across all sites by generation
climate_metrics_CONUS_temp <- climate_metrics_generation_site |> 
  group_by(generation) |> 
  summarise(CONUSmean_minflow_7d = mean(minflow_7d, na.rm = TRUE),
            n = n())

# Add in baseline flows for comparisons
climate_metrics_CONUS <- climate_metrics_CONUS_temp |>
  mutate(baseline_flow = 
           climate_metrics_CONUS_temp$CONUSmean_minflow_7d[
             climate_metrics_CONUS_temp$generation == "0 silent generation"],
         change_in_minflow = ((CONUSmean_minflow_7d - baseline_flow)/baseline_flow)*100,
         is_change_negative = case_when(change_in_minflow < 0 ~ "yes",
                                        change_in_minflow == 0 ~ "no change",
                                        change_in_minflow > 0 ~ "no"))
CONUS_statistic <- round(
  climate_metrics_CONUS$change_in_minflow[climate_metrics_CONUS$generation == "4 gen z"],0)

# Calculate averages by CASC
climate_metrics_CASC_temp <- climate_metrics_generation_site |> 
  left_join(metadata, by = "StaID") |>
  group_by(generation, CASC) |> 
  summarise(CASCmean_minflow_7d = mean(minflow_7d, na.rm = TRUE),
            n = n())

# Add in baselines
climate_metrics_CASC <- climate_metrics_CASC_temp |> 
  left_join(climate_metrics_CASC_temp |> 
              rename(baseline_flow = CASCmean_minflow_7d,
                     baseline_generation = generation) |>
              filter(baseline_generation == "0 silent generation") |>
              select(baseline_generation, CASC, baseline_flow),
            by = "CASC") |>
  mutate(change_in_minflow = ((CASCmean_minflow_7d - baseline_flow)/baseline_flow)*100,
         is_change_negative = case_when(change_in_minflow < 0 ~ "yes",
                                        change_in_minflow == 0 ~ "no change",
                                        change_in_minflow > 0 ~ "no"))
CASC_stats <- climate_metrics_CASC |> filter(generation == "4 gen z") |> 
  select(change_in_minflow, generation, CASC)

# Set up plotting data - CONUS
plotting_data_CONUS <- climate_metrics_year_site |> 
  filter(year %in% 1920:2019) |>
  group_by(year) |>
  summarise(CONUS_annual_minflow_7d = mean(minflow_7d, na.rm = TRUE)) |>
  mutate(generation = case_when(year %in% 1920:1945 ~ "0 silent generation",
                                year %in% 1946:1964 ~ "1 baby boomers",
                                year %in% 1965:1980 ~ "2 gen x",
                                year %in% 1981:1996 ~ "3 millenials",
                                year %in% 1997:2021 ~ "4 gen z")) |>
  left_join(climate_metrics_CONUS, by = c("generation"))

plotting_data_CASC <- climate_metrics_year_site |>
  filter(year %in% 1920:2019) |>
  left_join(metadata, by = "StaID") |>
  group_by(year, CASC) |> 
  summarise(CASC_annual_minflow_7d = mean(minflow_7d, na.rm = TRUE)) |>
  mutate(generation = case_when(year %in% 1920:1945 ~ "0 silent generation",
                                year %in% 1946:1964 ~ "1 baby boomers",
                                year %in% 1965:1980 ~ "2 gen x",
                                year %in% 1981:1996 ~ "3 millenials",
                                year %in% 1997:2021 ~ "4 gen z")) |>
  left_join(climate_metrics_CASC, by = c("generation", "CASC"))



mapping_data <- states_CASC |> 
  left_join(climate_metrics_CASC |> filter(generation == "4 gen z"),
                                         by = "CASC")

# convert id to CASC for mapping just regions to make them look "exploded"
mapping_data_regions <- mapping_data |>
  group_by(CASC) |>
  summarise(id = unique(CASC))

```

## Set up main plot

This chunk is where the main ggplot grob definition and set-up occurs. You may have several ggplot grobs, depending on your approach. For each one, if you define it with the `<-` assignment operator and surround the whole code statement in parentheses, you'll be able to preview here what the plot looks like before the next cowplot step.

> Note: The hardest part of this process is getting text to be the optimum size in the final `png` for Twitter. The font sizes here will **not** be the same as in the final image after the cowplot step. Make sure to check the output `png` for true font sizing and plot composition. 

```{r plotting}
# Load some custom fonts and set some custom settings
font_legend <- "Comfortaa"
sysfonts::font_add_google("Comfortaa")
supporting_font <- "Source Sans Pro"
sysfonts::font_add_google("Source Sans Pro")
showtext::showtext_opts(dpi = 300, regular.wt = 200, bold.wt = 700)
showtext::showtext_auto(enable = TRUE)

# Define colors
background_color = "#ffffff"
font_color = "black"

# The background canvas for your viz
canvas <- grid::rectGrob(
  x = 0, y = 0, 
  width = 16, height = 9,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color)
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_black.png") |>
  image_colorize(100, "#36454F")

# Set default chart
ggplot2::theme_set(theme_minimal())

# Main plot
(main_plot <- ggplot(data = plotting_data_CONUS, 
                     aes(x = year, y = CONUS_annual_minflow_7d)) +
    geom_line(color = "#ADAEB3") +
    geom_line(data = plotting_data_CONUS |> filter(change_in_minflow >= 0),
              aes(y = CONUSmean_minflow_7d, 
                  color = change_in_minflow, group = generation),
              linewidth = 2) +
    scale_color_viridis_c(limits = c(0,100), direction = -1, end = 0.6) +
    geom_line(data = plotting_data_CONUS |> 
                filter(change_in_minflow < 0),
              aes(y = CONUSmean_minflow_7d, group = generation),
              color = "#ef8a62",
              linewidth = 2) +
    geom_line(data = plotting_data_CONUS |> 
                filter(generation == "0 silent generation"),
              aes(y = CONUSmean_minflow_7d, group = generation),
              color = "#999999",
              linewidth = 2) +
    ylab("Minimum Annual Streamflow (cfs)") +
    theme(legend.position = "none",
          panel.grid = element_blank(),
          axis.title.x = element_blank(),
          axis.title.y = element_text(size = 14),
          axis.ticks = element_line()) +
    scale_x_continuous(breaks = c(1920, 1946, 1964, 1980, 1997, 2020))
)
# Chart settings
Ymargin <-  0.04
Xmargin <- 0.03

# CASC Plots
CASC_names <- unique(plotting_data_CASC$CASC)
Xbase <- 1 - Xmargin
Ybase <- Ymargin*2 + 0.02
Yjitter <- 0.25
Xjitter <- 0.2 

# Function to make regional plots
draw_CASC_plots <- map(1:length(CASC_names), function(x){
  plot_temp <- plotting_data_CASC |>
    filter(CASC == CASC_names[x]) |> 
    ggplot(aes(x = year, y = CASC_annual_minflow_7d)) +
    geom_line(linewidth = 0.5, color = "#ADAEB3") +
    geom_line(data = plotting_data_CASC |> 
                    filter(CASC == CASC_names[x]),
              aes(y = CASCmean_minflow_7d, group = generation, 
                  color = change_in_minflow),
              linewidth = 2) +
    scale_color_viridis_c(limits = c(0,100), direction = -1, end = 0.7) +
    {if(CASC_names[x] == "South Central"){
              geom_line(data = plotting_data_CASC |> 
                    filter(CASC == CASC_names[x], generation == "3 millenials"),
                  aes(y = CASCmean_minflow_7d, group = generation),
                  color = "blue",
              linewidth = 2) 
    }} +
        geom_line(data = plotting_data_CASC |> 
                    filter(CASC == CASC_names[x], generation == "0 silent generation"),
                  aes(y = CASCmean_minflow_7d, group = generation),
                  color = "#999999",
              linewidth = 2) +
        geom_line(data = plotting_data_CASC |> 
                    filter(CASC == CASC_names[x], change_in_minflow < 0),
                  aes(y = CASCmean_minflow_7d, group = generation),
                  color = "#ef8a62",
              linewidth = 2) +
    ggtitle(CASC_names[x]) + 
    theme(legend.position = "none",
        panel.grid = element_blank(),
        axis.title = element_blank()) +
    ylab("Minimum Annual Streamflow") +
    scale_x_continuous(breaks = c(1920, 1946, 1964, 1980, 1997, 2020)) #order: above, baseline, below
  
  if(CASC_names[x] %in% c("Southwest", "Southeast", "South Central")){
    Y_position <- Ybase
  } else if(CASC_names[x] %in% c("California", "Northeast")){
    Y_position <- Ybase + Yjitter
  } else {
    Y_position <- Ybase + 2*Yjitter
  }
  
  if(CASC_names[x] %in% c("Northwest", "California", "Southwest")){
    X_position <- Xbase - (2*Xjitter)
  } else if(CASC_names[x] %in% c("North Central", "South Central")){
    X_position <- Xbase - Xjitter
  } else {
    X_position <- Xbase
  } 
  
  draw_plot(plot_temp,
            x = X_position,
            y = Y_position,
            hjust = 1,
            height = 0.2,
            width = 0.2) 

})

# Function to make the labels for each regional plot
draw_CASC_labels <- map(1:length(CASC_names), function(x){

  if(CASC_names[x] %in% c("Southwest", "Southeast", "South Central")){
    Y_position <- Ybase + Yjitter - 0.09
  } else if(CASC_names[x] %in% c("California", "Northeast")){
    Y_position <- Ybase + 2*Yjitter - 0.09
  } else {
    Y_position <- Ybase + 3*Yjitter - 0.09
  }
  
  if(CASC_names[x] %in% c("Northwest", "California", "Southwest")){
    X_position <- Xbase - (2*Xjitter) -0.03
  } else if(CASC_names[x] %in% c("North Central", "South Central")){
    X_position <- Xbase - Xjitter - 0.03
  } else {
    X_position <- Xbase - 0.03
  } 

  draw_label(sprintf("%s%%", round(
    CASC_stats$change_in_minflow[CASC_stats$CASC == CASC_names[x]]), 0
    ),
            x = X_position,
            y = Y_position,
    fontfamily = font_legend,
    size = 18,
    color = ifelse(CASC_names[x] %in% c("Southwest", "Southeast"), "#ef8a62", "#1F9E89FF"))
})

# Regional map
(map_plot <- ggplot(data = mapping_data) +
    geom_sf(aes(fill = change_in_minflow), color = "white") +
    theme_void()+
    scale_fill_viridis_c(limits = c(0,100), direction = -1, end = 0.7) +
    geom_sf(data = mapping_data |> filter(change_in_minflow < 0), 
            fill = "#ef8a62", color = "white")+ 
    geom_sf(data = mapping_data_regions, color = "white", fill = NA, linewidth = 1)+
    geom_segment(aes(x = -125, xend = -127, 
                     y = 50, yend = 52), 
                 arrow = arrow(length = unit(0.03, "npc")))+
    geom_segment(aes(x = -104, xend = -104, 
                     y = 50, yend = 52), 
                 arrow = arrow(length = unit(0.03, "npc")))+
    geom_segment(aes(x = -85, xend = -75, 
                     y = 48, yend = 52), 
                 arrow = arrow(length = unit(0.03, "npc")))+
    geom_segment(aes(x = -74, xend = -70, 
                     y = 39, yend = 39), 
                 arrow = arrow(length = unit(0.03, "npc")))+
    geom_segment(aes(x = -124, xend = -127, 
                     y = 37, yend = 37), 
                 arrow = arrow(length = unit(0.03, "npc")))+
    geom_segment(aes(x = -113, xend = -127, 
                     y = 30, yend = 24), 
                 arrow = arrow(length = unit(0.03, "npc")))+
    geom_segment(aes(x = -101, xend = -101, 
                     y = 26, yend = 24), 
                 arrow = arrow(length = unit(0.03, "npc")))+
    geom_segment(aes(x = -80, xend = -70, 
                     y = 30, yend = 24), 
                 arrow = arrow(length = unit(0.03, "npc")))+
    theme(legend.position = "none")
)

```


## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

**Make sure to use the format for saving your png with the date of the prompt that you've been assigned to!** (e.g., `20230401_part-to-whole_cnell.png`)

```{r cowplot, fig.width = 16, fig.height = 9}
ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 9, width = 16,
            hjust = 0, vjust = 1) +
  # the main plot
  draw_plot(main_plot,
            x = Xmargin,
            y = 0.15,
            height = 0.35,
            width = 0.3,
            hjust = 0) +
  # Generation names
  draw_text("Silent\nGeneration\n(baseline)", 
            x = Xmargin + 0.08, y = 0.5, size = 12) +
  draw_text("Baby\nBoomers\n-5%", 
            x = Xmargin + 0.135, y = 0.4, size = 12) +
  draw_text("Gen X\n+14%", 
            x = Xmargin + 0.175, y = 0.5, size = 12) +
  draw_text("Millenials\n+18%", 
            x = Xmargin + 0.215, y = 0.51, size = 12) +
  draw_text("Gen Z\n+22%", 
            x = Xmargin + 0.26, y = 0.515, size = 12) +
  # map plot
  draw_plot(map_plot,
            x = Xbase - (2*Xjitter),
            y = Ybase + Yjitter,
            height = 0.2,
            width = 0.2) +
  draw_CASC_plots + 
  draw_CASC_labels +
  # explainer text
  draw_label("Minimum Annual Streamflow values are calculated by averaging annual 7-day low flows from 331 gages for which there are 100 years of data.\nData from doi.org/10.5066/P92FAASD. Chart by A.A. Archer.",
             fontfamily = supporting_font,
             x = 1-Xmargin,   
             y = Ymargin,
             size = 14,
             hjust = 1,
             vjust = 0,
             color = "#999999")+
  # Title
  draw_label("Not your parents' streamflows!",
             x = Xmargin,
             y = 1-Ymargin,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = "#31688EFF",
             size = 45,
             fontface="bold") +
    # Statistics
  draw_label("Across CONUS,\nMinimum Annual Streamflow\nduring Gen Z was\n\n\n\non average than during the\nSilent Generation",
             x = Xmargin + 0.15,
             y = 0.86,
             hjust = 0.5,
             vjust = 1,
             lineheight = 0.8,
             fontfamily = supporting_font,
             color = font_color,
             size = 25) +
  draw_label(sprintf("%s%% higher", CONUS_statistic), x = Xmargin + 0.15, y = 0.67,
            hjust = 0.5, vjust = 0, size = 52, fontfamily = font_legend, color = "#1F9E89FF")+
  draw_label("These changes in Minimum Annual Streamflow vary by region",
             x = Xmargin + 0.35,
             y = 0.85,
             hjust = 0,
             vjust = 1,
             lineheight = 0.85,
             fontfamily = supporting_font,
             color = font_color,
             size = 25)+
  # Add logo
  draw_image(usgs_logo, 
             x = Xmargin,
             y = Ymargin,
             width = 0.1, 
             hjust = 0, vjust = 0, 
             halign = 0, valign = 0)+
  draw_line(x = c(Xmargin, 1 - Xmargin),
           y = c(0.9, 0.9),
           color = "#31688EFF")+
  draw_line(x = c(0.135, 1 - Xmargin),
            y = c(0.09, 0.09),
            color = "#36454F")

# Save the final image in Twitter's 16 by 9 format
# !! Use format for saving with the date of your prompt: 
#         YYYYMMDD_prompt_name ()
# e.g. `20230101_part-to-whole-cnell.png`
ggsave(filename = "out/20230404_historical_aaarcher.png", 
       width = 16, height = 9, dpi = 300)

```



## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. Shows the minimum annual streamflow averaged by generation for 331 USGS gages that have 100 years of near complete data, by CONUS and by region 
2. Almost everywhere in the US, minimum annual streamflow levels have increased during the last few generations compared to the “Silent Generation” 
3. Minimum annual streamflow during Gen Z generation was 22% higher on average across the CONUS than it was during the Silent generation 
4. Two regions, southwest and southeast have slightly lower minimum annual streamflow levels now than they did during the silent generation 
5. Data from doi.org/10.5066/P92FAASD 
6. Uses #rstats and #OpenData 

### Data source(s)

Simeone, C.E., 2022, Streamflow Drought Metrics for Select United States Geological Survey Streamgages for Three Different Time Periods from 1921 - 2020: U.S. Geological Survey data release, https://doi.org/10.5066/P92FAASD.

