---
title: "Chart Challenge 2023: April 2, Waffle" 
author: "Elmera Azadpour"
date: "2023-03-10"
output: html_document
---

## Instructions for using this template

> Make sure you have copied the template subdirectory and named it with your prompt's date, prompt's name and your name (e.g., `/01_part-to-whole_cnell`) before editing this document!

1.  Put all comments either in the html text (like this) or in the `r` chunks (delineated with 3 "\`" marks) to explain your process to Vizlab reviewers
2.  Put all R code in the chunks. They're organized for you to (in order): do some set-up steps (load libraries), load in any data (manually or with built in packages), process your data for plotting, create plot using ggplot, and finally compose the twitter image with cowplot.
3.  You can run individual code chunks while developing code and plots. Note: the plots that are created in the RStudio coding window *do not* match the plot as created through the ggsave function.
4.  To preview the final plot, use the `Knit` button above. That will create a preview of the final plot in the output `html` and will save the `png` in the `out/` folder. On many systems, the preview in the script document and the html will not match the output `png`, so please refer to the `png` for final proofing.
5.  When you're happy with the final plot, fill out the supporting information at the bottom and push to gitlab for review. Note: only commit input data if it can't be downloaded directly with code (e.g., `sbtools` or `dataRetrieval`).

This website has great information on how to use RMarkdown: <https://yihui.org/knitr/>

# adjust working dir for targets

```{r}
library(here)
setwd(here("00_template_elmera_azadpour"))
```

### To get all data:

-   lets first run a targets pipeline to 1) get the data from sbtools and 2) do some munging/wrangling
-   to do this open `_targets.R` and run lines 1-10, if you do not have targets installed, run `install.packages("targets")`
-   now run `tar_make()`, you should see the pipeline buukd wuth outputs in `00_template_elmera_azadpour/out` folder

## Set up - lets load

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Define libraries here.

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(targets) # for water quality data
library(tidyverse) # includes ggplot
library(readr) # best for reading and writing csvs

# These are used for the layout of the viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo

# for waffle charts
# devtools::install_github("hrbrmstr/waffle")
library(waffle)
# for region maps
library(spData)
library(sf)
library(raster)
library(scico)
```

## Load files

Save any files that you are using directly in the `in` folder. Then read them in to the environment with this chunk of code. Remember, use the RStudio project and relative file pathways, and never include any personal computer file pathways.

This is also a good place to load any data from other built-in packages such as `dataRetrieval` or `spData`, etc.

```{r load}
tar_make() # run pipeline
tar_load(threshold_decadal_gw) #load threshold data target for plotting
tar_load(wells_centroid_csv) #load spatial data for plotting 
```

## Get data ready for plotting

This next section is for doing any pre-processing steps, data joins, etc, to get your data ready for plotting with ggplot. Remember that most of the time, ggplot prefers data in a "wide" format.

```{r processing}
# lets add a column in `p2_threshold_decadal_gw` using case when to then group networks by 4 contiguous U.S regions: West, Central, Midwest, East
threshold_decadal_gw_reg <- threshold_decadal_gw |> 
  mutate(region = case_when(study_unit_abbreviation == 'acad'~ 'central',
                            study_unit_abbreviation == 'acfb'~ "east",
                            study_unit_abbreviation == 'albe' ~ 'east',
                            study_unit_abbreviation == 'cazb' ~ 'west',
                            study_unit_abbreviation == 'ccpt' ~ 'west',
                            study_unit_abbreviation == 'cnbr' ~ 'central',
                            study_unit_abbreviation == 'conn' ~ 'east',
                            study_unit_abbreviation == 'dlmv' ~ 'east',
                            study_unit_abbreviation == 'eiwa' ~ 'midwest',
                            study_unit_abbreviation == 'gafl' ~ 'east',
                            study_unit_abbreviation == 'grsl' ~ 'west',
                            study_unit_abbreviation == 'hpgw' ~ 'central',
                            study_unit_abbreviation == 'leri' ~ 'midwest',
                            study_unit_abbreviation == 'linj' ~ 'east',
                            study_unit_abbreviation == 'lirb' ~ 'midwest',
                            study_unit_abbreviation == 'miam' ~ 'midwest',
                            study_unit_abbreviation == 'mise' ~ 'midwest',
                            study_unit_abbreviation == 'mobl' ~ 'east',
                            study_unit_abbreviation == 'necb' ~ 'east',
                            study_unit_abbreviation == 'nvbr' ~ 'west',
                            study_unit_abbreviation == 'ozrk' ~ 'central',
                            study_unit_abbreviation == 'podl' ~ 'east',
                            study_unit_abbreviation == 'poto' ~ 'east',
                            study_unit_abbreviation == 'pugt' ~ 'west',
                            study_unit_abbreviation == 'riog' ~ 'west',
                            study_unit_abbreviation == 'sacr' ~ 'west',
                            study_unit_abbreviation == 'sana' ~ 'west',
                            study_unit_abbreviation == 'sanj' ~ 'west',
                            study_unit_abbreviation == 'sant' ~ 'east',
                            study_unit_abbreviation == 'sctx' ~ 'central',
                            study_unit_abbreviation == 'sofl' ~ 'east',
                            study_unit_abbreviation == 'splt' ~ 'central',
                            study_unit_abbreviation == 'trin' ~ 'central',
                            study_unit_abbreviation == 'uirb' ~ 'midwest',
                            study_unit_abbreviation == 'umis' ~ 'midwest',
                            study_unit_abbreviation == 'usnk' ~ 'west',
                            study_unit_abbreviation == 'whit' ~ 'midwest',
                            study_unit_abbreviation == 'wmic' ~ 'midwest'))

# make ratio columns 
threshold_decadal_gw_reg_grp <- threshold_decadal_gw_reg |> 
  group_by(parameter, region, bins) |> 
  summarise(count_bins_sum=sum(count_bins),
            count_obs_sum = sum(count_obs))  |> 
  mutate(ratio = round(count_bins_sum/count_obs_sum*100)) |> 
  filter(parameter %in% c("pH", "DO","As", "Fe", "Li_dw", "Li_hhb", "Mo", "NO3", "Pb", "Sr")) |>
  #capitalize regions
  mutate(region = str_to_title(gsub(",", " ", region))) |> 
  arrange(match(bins, c("high", "moderate", "low")))

## sf regions 
wells_centroid <-readr::read_csv(wells_centroid_csv,show_col_types = FALSE)  |> 
  janitor::clean_names() |> 
  filter(!is.na(longitude_nad83_dd)) |> 
  sf::st_as_sf(coords = c("longitude_nad83_dd", "latitude_nad83_dd"), crs = "epsg:4269") |>  
  rename(su_code = network)

threshold_decadal_gw_reg_grp_sf <- threshold_decadal_gw_reg |> 
  left_join(wells_centroid, by = c("su_code")) |> 
  dplyr::select(su_code, region, geometry) |> 
  st_as_sf(crs = "epsg:4269") 

# Get state outlines
states_outline <-  spData::us_states |>
  st_transform(crs = st_crs(wells_centroid))

states_thresholds <- spData::us_states |> 
  st_transform(crs = st_crs(wells_centroid)) |> 
  st_intersection(threshold_decadal_gw_reg_grp_sf) 
```

## Set up main plot

This chunk is where the main ggplot grob definition and set-up occurs. You may have several ggplot grobs, depending on your approach. For each one, if you define it with the `<-` assignment operator and surround the whole code statement in parentheses, you'll be able to preview here what the plot looks like before the next cowplot step.

> Note: The hardest part of this process is getting text to be the optimum size in the final `png` for Twitter. The font sizes here will **not** be the same as in the final image after the cowplot step. Make sure to check the output `png` for true font sizing and plot composition.

```{r plotting}
# unique regions
uniq_reg <- unique(states_thresholds$region)

# loop sf region plots for each unique region
for (i in seq_along(uniq_reg)) {
    data_sub = subset(
      states_thresholds, 
      region==uniq_reg[[i]])
      
      data_crop <- states_outline |> st_crop(data_sub)
      
    if (nrow(data) == 0) {
      next
    }
# sf plots 
ggplot() + 
  geom_sf(data = data_crop ,
          fill = NA,
          color = 'black',
          linewidth = 0.3) +
  geom_sf(data = data_sub,
          fill = '#e45c5c',
            color = '#e45c5c',
          size = 1.5,
          alpha = 0.5) +
  theme_void() 

ggsave(file = paste0("out/",uniq_reg[i], ".svg"),
       width = 1.6, 
       height = 1.6,
       units = "in",
       bg = "white")

}
# Load some custom fonts and set some custom settings
font_legend <- 'Merriweather Sans'
font_add_google(font_legend)
showtext_opts(dpi = 300, regular.wt = 300, bold.wt = 800)
showtext_auto(enable = TRUE)

  
# waffle charts - bug on facet_grid see here: https://github.com/hrbrmstr/waffle/issues/66
# save as svgs and rearrange in adobe illustrator
plt <- threshold_decadal_gw_reg_grp |> 
  split(threshold_decadal_gw_reg_grp$parameter) |> 
  map(~ggplot(.x, aes(values = ratio, fill = bins)) +
  geom_waffle(color = "white", size=1.125, n_rows = 10, 
              make_proportional = TRUE,
              stat = "identity") +
  facet_wrap(~factor(region, levels = c("West", "Central", "Midwest", "East"))) +
  coord_equal() + 
  scale_x_discrete(expand=c(0,0)) +
  scale_y_discrete(expand=c(0,0)) +
  scale_fill_manual(name=NULL, values = c('#36161a', "#b54a56","#e9c9cc"), breaks=c('high', 'moderate', 'low'),
                    labels = c("High", "Moderate", "Low")) +
  # pink scale, iron: '#36161a', "#b54a56","#e9c9cc"
  # blue scale, strontium : '#162336', "#3b5d91","#b7c8e1"  
  # brown sclae, lead : '#52402d', "#947352","#c9b39c" 
  # green scale, nitrate : "#37402a", "#798c5b","#b9c5a6"
  # gray scale, legend: '#3A3A3A', "#808080","#BDBDBD"
  labs(
    title = .x |> pull(parameter) |> unique(),
    subtitle = "Proportion of study area" 
  ) +
  theme_void() +
      theme(strip.text = element_text(size = 26,
                                      margin=margin(b=5)),
            legend.key.width = unit(0.75, "cm"),
            legend.key.height = unit(0.75, "cm"),
            legend.spacing.x = unit(0.5, 'cm'),
            legend.text = element_text(size = 16),
            legend.title = element_text(size = 26),
            legend.direction = "horizontal",
            legend.position = "top", 
            text = element_text(size=22,family = font_legend, color = "black", face= "bold"),
            plot.title = element_text(hjust = 0.5),
            plot.subtitle = element_text(hjust = 0.5, size = 20)) +
      guides(fill = guide_legend(title.position = "top", title.hjust=0.5))) 

  pwalk(list(paste0("out/", "region", "_", stringr::str_c(names(plt), ".svg")), plt),
        ggsave, 
        width=12, 
        height=12,
        units = "in",
        bg = "white",
        path = ".")

# Plotting just one constituent concentrations regionally  
# Define colors
background_color = "#ffffff"
font_color =  "#000000"

# The background canvas for your viz
canvas <- grid::rectGrob(
  x = 0, y = 0,
  width = 16, height = 9,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color)
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_black.png")

# Main plot
# For the purposes of this markdown, I'm just displaying one constituents regional variation. Final plot with small multiples is craeted in illustrator due to faceting bug. 
annotate_text <-"Shadows Into Light"
font_add_google(annotate_text)
showtext_opts(dpi = 300, regular.wt = 300, bold.wt = 800)
showtext_auto(enable = TRUE)

(main_plot <- plt$Fe + theme(plot.margin = unit(c(2, -25, 2, -25), "cm"),
                             strip.text.x = element_text(hjust = 0.25)) +
      # Add iron label and arrow
       geom_curve(data =threshold_decadal_gw_reg_grp |>  filter(region %in% c("East")),
                  aes(x = 14.8,
                  xend = 11.6,
                  yend = 3.2, y = 5.4),
               arrow = grid::arrow(length = unit(0.5, 'lines')), 
               curvature = -0.5,
               angle = 34,
               color ='black') +
    geom_text(data =threshold_decadal_gw_reg_grp |>  filter(region %in% c("East")),
              aes(x = 16, y = 7.2),
                  label = "Fe deposits in\nnearby limestone\nand dolomite",
              family = annotate_text,
              hjust = 1, vjust = 0.5,
              color = 'black', size = 5)
  )
```

## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

**Make sure to use the format for saving your png with the date of the prompt that you've been assigned to!** (e.g., `20230401_part-to-whole_cnell.png`)

```{r cowplot, fig.width = 16, fig.height = 9}
ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 9, width = 16,
            hjust = 0, vjust = 1) +
  # the main plot
  draw_plot(main_plot,
            x = 0.01,
            y = 0.01,
            height = 1) +
  # explainer text
  draw_label("Elmera Azadpour, USGS\nData from doi.org/10.5066/P9FZT1WO &\ndoi.org/10.1021/acsestwater.2c00390",
             fontfamily = font_legend,
             x = 0.96,
             y = 0.05,
             size = 14,
             hjust = 1,
             vjust = 0,
             color = font_color) +
#   # Title
  draw_label("Groundwater\ncontaminant\nconcentrations\n(2012 - 2021)",
             x = 0.04,
             y = 0.95,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = font_color,
             size = 34) +
  # Add logo
  draw_image(usgs_logo,
             x = 0.04,
             y = 0.05,
             width = 0.1,
             hjust = 0, vjust = 0,
             halign = 0, valign = 0)

# Save the final image in Twitter's 16 by 9 format
# !! Use format for saving with the date of your prompt: 
#         YYYYMMDD_prompt_name ()
# e.g. `20230101_part-to-whole-cnell.png`
ggsave(filename = "out/20230402_waffle_eazadpour.png",
       width = 16, height = 9, dpi = 300)
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. High DO in Western region - thought not necessarily a bad thing. High DO levels improve teh taste of drinking water but may cause corrosion in pipes. Alternatively, low DO levels may create hypoxic conditions.
2. High Pb in the East - Pb contamination may occur from metal plumbing; [Jurgens et al., 2019](https://pubs.acs.org/doi/10.1021/acs.est.8b04475) found that the majority of elevated and high Pb tended to occur in clusters in the Eastern U.S.
3. High NO3- in Central U.S. - shallow aquifers, [like the Ogallala](https://sites.tufts.edu/gis/files/2013/11/Gill_John.pdf), that underlie agricultural areas can lead to high NO3- concentrations due to high levels of fertilizer use and permeable soils.
4. High Fe in the East - Fe deposits in surrounding [limestone and dolomite](https://longsecowater.com/blog/effects-of-iron-in-water) can lead to increased Fe concentrations.

### Data source(s)

Lindsey, B.D., May, A.N., and Johnson, T.D., 2022, Data from Decadal Change in Groundwater Quality Web Site, 1988-2021: U.S. Geological Survey data release, https://doi.org/10.5066/P9FZT1WO.

Belitz, K., Fram, M. S., Lindsey, B. D., Stackelberg, P. E., Bexfield, L. M., Johnson, T. D., ... & Dubrovsky, N. M. (2022). Quality of Groundwater Used for Public Supply in the Continental United States: A Comprehensive Assessment. ACS ES&T Water, 2(12), 2645-2656. https://doi.org/10.1021/acs.est.2c08061 
