---
title: "Chart Challenge 2023: April 2, Pop Culture" 
author: "Elmera Azadpour"
date: "2023-04-10"
output: html_document
---

## Instructions for using this template

> Make sure you have copied the template subdirectory and named it with your prompt's date, prompt's name and your name (e.g., `/01_part-to-whole_cnell`) before editing this document!

1.  Put all comments either in the html text (like this) or in the `r` chunks (delineated with 3 "\`" marks) to explain your process to Vizlab reviewers
2.  Put all R code in the chunks. They're organized for you to (in order): do some set-up steps (load libraries), load in any data (manually or with built in packages), process your data for plotting, create plot using ggplot, and finally compose the twitter image with cowplot.
3.  You can run individual code chunks while developing code and plots. Note: the plots that are created in the RStudio coding window *do not* match the plot as created through the ggsave function.
4.  To preview the final plot, use the `Knit` button above. That will create a preview of the final plot in the output `html` and will save the `png` in the `out/` folder. On many systems, the preview in the script document and the html will not match the output `png`, so please refer to the `png` for final proofing.
5.  When you're happy with the final plot, fill out the supporting information at the bottom and push to gitlab for review. Note: only commit input data if it can't be downloaded directly with code (e.g., `sbtools` or `dataRetrieval`).

This website has great information on how to use RMarkdown: <https://yihui.org/knitr/>

# adjust working dir for targets

```{r}

```

### To get all data:

-   lets first run a targets pipeline to 1) get the data from sbtools and 2) do some munging/wrangling
-   to do this open `_targets.R` and run lines 1-10, if you do not have targets installed, run `install.packages("targets")`
-   now run `tar_make()`, you should see the pipeline build with outputs in `02_waffle_eazadpour/out` folder

## Set up - lets load

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Define libraries here.

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) # includes ggplot
library(dataRetrieval) # pull gage data
library(ggridges) # make ridge plots
# library(gridExtra) #annotating
# library(grid)

# These are used for the layout of the viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo
```

## Load files

Save any files that you are using directly in the `in` folder. Then read them in to the environment with this chunk of code. Remember, use the RStudio project and relative file pathways, and never include any personal computer file pathways.

This is also a good place to load any data from other built-in packages such as `dataRetrieval` or `spData`, etc.

```{r load}
# select timeframe to plot by
start_date <- "2011-01-01"
end_date <- "2022-12-31"

# select gage of interest
# I've selected Rio Grande at Embudo, NM - https://waterdata.usgs.gov/monitoring-location/08279500/#parameterCode=00065&period=P7D  
site_no = "08279500"

# pull NWIS data
stream_data_raw <- readNWISuv(siteNumbers = site_no, parameterCd = c("00060", "00065"),
                         startDate = as.Date(start_date), endDate = as.Date(end_date)) |>
  renameNWISColumns()
```

## Get data ready for plotting

This next section is for doing any pre-processing steps, data joins, etc, to get your data ready for plotting with ggplot. Remember that most of the time, ggplot prefers data in a "wide" format.

```{r processing}
# munge dates and get mean daily gage height (f)
stream_data <- stream_data_raw  %>%
  filter(dateTime >= as.POSIXct(start_date), dateTime <= as.POSIXct(end_date)) |>
  mutate(julian_day = yday(dateTime),
         year = as.numeric(format(dateTime, "%Y")),
         day = as.numeric(format(dateTime, "%d")),
         month = as.numeric(format(dateTime, "%m"))) |>
  group_by(site_no, year, julian_day) |>
  # users can select what parameter they would like to summarize by (flow, gage height, etc.)
  # more on parameters here: https://help.waterdata.usgs.gov/code/parameter_cd_query?fmt=rdb&inline=true&group_cd=%
  summarise(mean_daily_gh_f = mean(GH_Inst))
```

## Set up main plot

This chunk is where the main ggplot grob definition and set-up occurs. You may have several ggplot grobs, depending on your approach. For each one, if you define it with the `<-` assignment operator and surround the whole code statement in parentheses, you'll be able to preview here what the plot looks like before the next cowplot step.

> Note: The hardest part of this process is getting text to be the optimum size in the final `png` for Twitter. The font sizes here will **not** be the same as in the final image after the cowplot step. Make sure to check the output `png` for true font sizing and plot composition.

```{r plotting}
# Load some custom fonts and set some custom settings
font_legend <- 'Yantramanav'
font_add_google(font_legend)
showtext_opts(dpi = 300, regular.wt = 300, bold.wt = 800)
showtext_auto(enable = TRUE)

  
main_plot <- ggplot(stream_data, aes(x = julian_day,  y= year, group = year, height = mean_daily_gh_f)) +
  geom_density_ridges(stat = "identity", scale = 4,
                      fill = "black", color = 'white',
                      alpha=0.5) +
  coord_cartesian(clip="off") +
  geom_text(x = 183, y = 2011, label = "GAGE HEIGHT", color = "white", size = 8, hjust = 0.5) +
  scale_x_continuous(name = NULL) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "black"),
        axis.line = element_blank(),
        axis.text.x = element_blank(),
        plot.background = element_rect(fill = "black"),
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(size=22,family = font_legend, color = "white"),
        plot.title=element_text(hjust=0.5, vjust = -0.5),
        plot.margin = unit(c(1,1,1,1), "lines")) +
  labs(title = "RIO GRANDE RIVER, NEW MEXICO")

# Plotting just one constituent concentrations regionally  
# Define colors
font_color = "#ffffff"
background_color =  "#000000"

# The background canvas for your viz
canvas <- grid::rectGrob(
  x = 0, y = 0,
  width = 16, height = 9,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color)
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_white.png")

```

## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

**Make sure to use the format for saving your png with the date of the prompt that you've been assigned to!** (e.g., `20230401_part-to-whole_cnell.png`)

```{r cowplot, fig.width = 16, fig.height = 9}
ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 8, width = 8,
            hjust = 0, vjust = 1) +
  # the main plot
  draw_plot(main_plot,
            x = 0.01,
            y = 0.01,
            height = 1) +
  # explainer text
  draw_label("Elmera Azadpour, USGS\nData: NWIS",
             fontfamily = font_legend,
             x = 0.96,
             y = 0.05,
             size = 10,
             hjust = 1,
             vjust = 0,
             color = font_color) +
  # Add logo
  draw_image(usgs_logo,
             x = 0.04,
             y = 0.05,
             width = 0.1,
             hjust = 0, vjust = 0,
             halign = 0, valign = 0)

# Save the final image in Twitter's 16 by 9 format
# !! Use format for saving with the date of your prompt: 
#         YYYYMMDD_prompt_name ()
# e.g. `20230101_part-to-whole-cnell.png`
ggsave(filename = "out/20230410_pop-culture_eazadpour.png",
       width = 8, height = 8, dpi = 300,
       bg = "black")
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)
1) Ridge lines display daily mean gage height form 2011 to 2022 for Rio Grande at Embudo, NM (08279500) 
2) Play on Joy Division - Unknown Pleasures album cover 
Note: we recognize the history of `geom_joy()`and renaming to `geom_density_ridgline()`, more on this from Claus O. Wilke: https://clauswilke.com/blog/2017/09/15/goodbye-joyplots/ 

### Data source(s)
U.S. Geological Survey, 2016, National Water Information System data available on the World Wide Web (USGS Water Data for the Nation), accessed [April 10, 2023], at URL [http://waterdata.usgs.gov/nwis/].

