---
title: "Day 28: Trends"
author: "Julie Padilla"
date: "2023-04-24"
output: html_document
---

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```



Define libraries here. 

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(targets)
library(tidyverse) # includes ggplot
library(readr) # best for reading and writing csvs
library(terra) # here, used for raster processing
library(spData) # state boundaries as sf object
library(sf) # general spatial processing
library(scico) # good color themes for mapping
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo
```



## Load files

Load the appropriate targets pipeline outputs


```{r load}

tar_load(p3_great_lakes_ice_bar_png)

```

## Set up main plot

This chunk is where the output from the targets pipeline is combined with standard VizLab formatting

```{r plotting}
# Load some custom fonts and set some custom settings
font_legend <- "Lato"
sysfonts::font_add_google("Lato")
supporting_font <- "Source Sans Pro"
sysfonts::font_add_google("Source Sans Pro")
showtext::showtext_opts(dpi = 300, regular.wt = 200, bold.wt = 700)
showtext::showtext_auto(enable = TRUE)

# Define colors
background_color = "white"
font_color = "gray15"

# The background canvas for your viz
canvas <- grid::rectGrob(
  x = 0, y = 0, 
  width = 16, height = 9,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color)
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_black.png") 

# Main plot
(main_plot <- p3_great_lakes_ice_bar_png)

```


## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

**Make sure to use the format for saving your png with the date of the prompt that you've been assigned to!** (e.g., `20230101_part-to-whole_cnell.png`)

```{r cowplot, fig.width = 16, fig.height = 9}
out <- 
  ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
              xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 9, width = 16,
            hjust = 0, vjust = 1) +
  # the main plot
  draw_plot(main_plot,
            x = - 0.01,
            y = 0.11,
            height = 0.85) +
  # explainer text
  draw_label("Julie Padilla, USGS\nData from NOAA Great Lakes Environmental Research Lab (GLERL)\nU.S. Geological Survey",
             fontfamily = font_legend,
             fontface = 'italic',
             x = 0.96,
             y = 0.04,
             size = 14,
             hjust = 1,
             vjust = 0,
             color = font_color) +
  # Add logo
draw_image(usgs_logo,
           x = 0.04,
           y = 0.04,
           width = 0.1,
           hjust = 0, vjust = 0,
           halign = 0, valign = 0)

ggsave(plot = out, filename = "out/20230428_trends_greatlakesice.png", 
       width = 16, height = 9, dpi = 300, units = "in")
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. 2023 ice coverage on all 5 of the Great Lakes was significantly below the 50 year average
2. The greatest deviation from the average was on Lake Superior (- 67% below average) and the smallest deviation was on Lake Ontario (- 45%).
3. Basin-wide 2023 ice coverage was 59% below average

### Data source(s)

Ice data is from the Great Lakes Environmental Research Lab 

* [press release](https://research.noaa.gov/article/ArtMID/587/ArticleID/2941/Low-ice-on-the-Great-Lakes-this-winter)
* [Daily ice data from 1973-2022](https://www.glerl.noaa.gov/data/ice/glicd/daily/)
* [Daily ice data from 2023](https://coastwatch.glerl.noaa.gov/statistic/ice/dat/g2022_2023_ice.dat)

Great Lakes spatial data is from [US Geological Survey](https://www.sciencebase.gov/catalog/item/530f8a0ee4b0e7e46bd300dd)


