---
title: "Chart Challenge 2023: April 28, Trend" 
author: "Elmera Azadpour"
date: "2023-04-21"
output: html_document
---

## Instructions for using this template

> Make sure you have copied the template subdirectory and named it with your prompt's date, prompt's name and your name (e.g., `/01_part-to-whole_cnell`) before editing this document!

1.  Put all comments either in the html text (like this) or in the `r` chunks (delineated with 3 "\`" marks) to explain your process to Vizlab reviewers
2.  Put all R code in the chunks. They're organized for you to (in order): do some set-up steps (load libraries), load in any data (manually or with built in packages), process your data for plotting, create plot using ggplot, and finally compose the twitter image with cowplot.
3.  You can run individual code chunks while developing code and plots. Note: the plots that are created in the RStudio coding window *do not* match the plot as created through the ggsave function.
4.  To preview the final plot, use the `Knit` button above. That will create a preview of the final plot in the output `html` and will save the `png` in the `out/` folder. On many systems, the preview in the script document and the html will not match the output `png`, so please refer to the `png` for final proofing.
5.  When you're happy with the final plot, fill out the supporting information at the bottom and push to gitlab for review. Note: only commit input data if it can't be downloaded directly with code (e.g., `sbtools` or `dataRetrieval`).

This website has great information on how to use RMarkdown: <https://yihui.org/knitr/>

## Set up - lets load

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Define libraries here.

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) # includes ggplot
library(sbtools) # for pulling sciencebase data
library(readr) # for reading csv 
library(janitor) # for cleaning dataframe column names
library(geofacet) # for the facet map of the U.S. 
library(grid) # plot layout
library(png) # for adding oak tree bckground img
library(here) # for easy assessing oak tree background img

# These are used for the layout of the viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo
```

## Load files

Save any files that you are using directly in the `in` folder. Then read them in to the environment with this chunk of code. Remember, use the RStudio project and relative file pathways, and never include any personal computer file pathways.

This is also a good place to load any data from other built-in packages such as `dataRetrieval` or `spData`, etc.

```{r load}
# Load forest area across CONUS 1985-2020 
#sbtools::authenticate_sb()

file_in_path_continuous <- "in/summary_forest_area_by_year.csv"
if(!file.exists(file_in_path_continuous)){ # if files don't exist, download
  sbtools::item_file_download(sb_id = "6287f350d34e4fef2ec3c2b2",
                              names = "summary_forest_area_by_year.csv",
                              destinations =  "in/summary_forest_area_by_year.csv",
                              overwrite_file = F)
}

annual_forest_area_raw <- readr::read_csv(file_in_path_continuous) |> 
  janitor::clean_names() |> 
  rename(name = state)
```

## Get data ready for plotting

This next section is for doing any pre-processing steps, data joins, etc, to get your data ready for plotting with ggplot. Remember that most of the time, ggplot prefers data in a "wide" format.

```{r processing}
# get all unique states to then filter us grid by 
state_names <- unique(annual_forest_area_raw$name)

# get state grid minus Alaska, Hawaii, Wisonsin for legend plt
my_us_grid_dropWI <- us_state_grid1[c(-2, -11 , -49), ]

# join state (minus WI, AK, WI) grid data with annual forest data, filter states to plot with geofacet, and calculate zscore from 35 yr mean for each state
my_grid_zscore_dropWI <- my_us_grid_dropWI |> 
  filter(!code %in% c("AK", "HI", "WI")) |> 
  left_join(annual_forest_area_raw, by = "name") |> 
  filter(name %in% state_names) |> 
  group_by(code) |>  
  mutate(z_score = as.numeric(scale(area))) |> 
  ungroup() |> 
  mutate(area_change = case_when(z_score > 0 ~ "POSITIVE",
                                        z_score < 0 ~ "NEGATIVE"))

# join state (minus AK, HI) grid data with annual forest data, filter states to plot with geofacet, and calculate zscore from 35 yr mean for each state
my_grid_zscore_conus <- us_state_grid1 |> 
  filter(!code %in% c("AK", "HI")) |> 
  left_join(annual_forest_area_raw, by = "name") |> 
  filter(name %in% state_names) |> 
  group_by(code) |>  
  mutate(z_score = as.numeric(scale(area))) |> 
  ungroup() |> 
  mutate(area_change = case_when(z_score > 0 ~ "POSITIVE",
                                        z_score < 0 ~ "NEGATIVE"))
```

## Set up main plot

This chunk is where the main ggplot grob definition and set-up occurs. You may have several ggplot grobs, depending on your approach. For each one, if you define it with the `<-` assignment operator and surround the whole code statement in parentheses, you'll be able to preview here what the plot looks like before the next cowplot step.

> Note: The hardest part of this process is getting text to be the optimum size in the final `png` for Twitter. The font sizes here will **not** be the same as in the final image after the cowplot step. Make sure to check the output `png` for true font sizing and plot composition.

```{r plotting}
# Load some custom fonts and set some custom settings
font_legend <- 'PT Sans'
font_add_google(font_legend)
showtext_opts(dpi = 300, regular.wt = 300, bold.wt = 800)
showtext_auto(enable = TRUE)

postive_col = "#083d21"
negative_col = "#CC5500"

(main_plot <- ggplot(my_grid_zscore_dropWI, aes(x = year, y = z_score)) +
  # Make positive zscores darker green 
  geom_segment(data = my_grid_zscore_dropWI |> filter(area_change %in% c("POSITIVE")), 
                 aes(x=year, xend=year, y=0, yend=z_score), color= postive_col,
               linewidth = 0.5) +
    geom_point(data = my_grid_zscore_dropWI |> filter(area_change %in% c("POSITIVE")),
             color=postive_col, size=0.5) + 
    # Make negative zscores burnt orange 
    geom_segment(data = my_grid_zscore_dropWI |> filter(area_change %in% c("NEGATIVE")), 
                 aes(x=year, xend=year, y=0, yend=z_score), color= negative_col,
               linewidth = 0.5) +
    geom_point(data = my_grid_zscore_dropWI |> filter(area_change %in% c("NEGATIVE")),
             color= negative_col, size=0.5) + 
  scale_x_continuous(limits = c(1985, 2020), breaks = seq(1985, 2020, by = 10)) + 
  facet_geo(~ code, grid = my_us_grid_dropWI) +
  labs(x = NULL, y= NULL) +
    theme_classic(base_size = 12) + 
    theme(strip.background = element_blank(),
          strip.text = element_text(vjust = 1,
                                    size = 12,
                                    family = font_legend,
                                    margin = margin(b = -2),
                                    color = "black"),
          strip.placement = "inside",
          strip.background.x = element_blank(),
          axis.line = element_blank(),
          axis.ticks = element_blank(),
          axis.text = element_blank(),
          axis.title = element_blank(),
          panel.border = element_blank(),
          plot.background = element_rect(fill = 'white', color = NA),
          panel.background = element_blank(),
          panel.spacing.x = unit(4, "pt"),
          panel.spacing.y = unit(7, "pt"),
          plot.margin = margin(50, 10, 10, 10, "pt"),
          legend.background = element_rect(fill = 'white', color = NA))
)


(legend_plot <- ggplot(my_grid_zscore_conus |> filter(name %in% c('Wisconsin')), aes(x = year, y = z_score)) +
  # Make positive zscores darker green 
  geom_segment(data = my_grid_zscore_conus |> filter(area_change %in% c("POSITIVE") & 
                                                name %in% c('Wisconsin') ), 
                 aes(x=year, xend=year, y=0, yend=z_score), color= postive_col,
               linewidth = 0.5) +
    geom_point(data = my_grid_zscore_conus |> filter(area_change %in% c("POSITIVE") & 
                                                name %in% c('Wisconsin')),
             color= postive_col, size=0.5) + 
    # Make negative zscores burnt orange 
    geom_segment(data = my_grid_zscore_conus |> filter(area_change %in% c("NEGATIVE") &
                                                  name %in% c('Wisconsin')), 
                 aes(x=year, xend=year, y=0, yend=z_score), color= negative_col,
               linewidth = 0.5) +
    geom_point(data = my_grid_zscore_conus |> filter(area_change %in% c("NEGATIVE") &
                                                name %in% c('Wisconsin')),
             color= negative_col, size=0.5) +
  facet_wrap(~code) +
  scale_y_continuous(limits = c(-3, 3), breaks = c(-3, 0, 3)) +
  scale_x_discrete(limits = c(1985, 2020), labels = c("1985", "2020")) +
  labs(x = NULL, y = "Z-Score") +
  theme_minimal() +
  theme(panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_blank(),
        text = element_text(family = font_legend, size = 10),
        strip.text = element_text(size = 12,
                                  margin = margin(b = -2),
                                  vjust = 1, hjust = 0.365,
                                  family = font_legend,
                                  color = "black"),
        axis.text = element_text(size = 12), 
        panel.background = element_rect(fill = 'white', color = "white")
        )) 

# Define colors
background_color = "#ffffff"
font_color =  "#000000"

# The background canvas for your viz
canvas <- grid::rectGrob(
  x = 0, y = 0,
  width = 16, height = 9,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color
                  )
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_black.png")

```

## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

**Make sure to use the format for saving your png with the date of the prompt that you've been assigned to!** (e.g., `20230401_part-to-whole_cnell.png`)

```{r cowplot, fig.width = 16, fig.height = 9}
custom_plt <- ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 8, width = 8,
            hjust = 0, vjust = 1) +
  # the main plot
  draw_plot(main_plot,
            x = 0.001,
            y = 0.01,
            height = 1) +
  # # Wisconsin legend plot
  draw_plot(legend_plot, .4450, .81, .115, .15) +
  # explainer text
  draw_label("Elmera Azadpour, USGS\nData: doi.org/10.5066/P9PRH8DQ",
             fontfamily = font_legend,
             x = 0.983,
             y = 0.039,
             size = 10,
             hjust = 1,
             vjust = 0,
             color = font_color) +
  #  Title
  # uggh! fix expression pasting issue 
  draw_label(expression(paste("Forest area"~(km^2))),
             x = 0.04,
             y = 0.99,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = font_color,
             size = 26) +
    draw_label("difference",
             x = 0.215,
             y = 0.98,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = font_color,
             size = 26) +
      draw_label("from 35 year mean (1985-2020)",
             x = 0.04,
             y = 0.93,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = font_color,
             size = 26) +
        draw_label("across CONUS",
             x = 0.04,
             y = 0.88,
             hjust = 0,
             vjust = 1,
             lineheight = 0.75,
             fontfamily = font_legend,
             color = font_color,
             size = 26) +
  # Add logo
  draw_image(usgs_logo,
             x = 0.04,
             y = 0.034,
             width = 0.1,
             hjust = 0, vjust = 0,
             halign = 0, valign = 0)

# add oak tree png 
im <- png::readPNG(here("28_trend_eazadpour/in/oak-tree.png"))
im2 <- matrix(rgb(im[,,1],im[,,2],im[,,3], im[,,4] * 0.05), nrow=dim(im)[1]) ## can adjust 0.05 alpha 

final_plt <- custom_plt +
    annotation_custom(rasterGrob(im2, 
                               width = unit(1,"npc"), 
                               height = unit(1,"npc")), 
                    0.6, 1.21, -Inf, 1.5) 

# Save the final image in Twitter's 16 by 9 format
# !! Use format for saving with the date of your prompt: 
#         YYYYMMDD_prompt_name ()
# e.g. `20230101_part-to-whole-cnell.png`
ggsave(filename = "out/20230428_trend_eazadpour.png",
       width = 16, height = 9, dpi = 300,
       bg = "white")
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)
1) Forest area is declining across the U.S., however states such as Kansas, Nebraska, Iowa, and North Dakota have seen an increase in forest area in recent years. 

### Data source(s)
Sleeter, B.M., and Selmants, P.C., 2022, Frequency of forest change across the conterminous United States from 1985-2020: U.S. Geological Survey data release, https://doi.org/10.5066/P9PRH8DQ.

