---
title: "Water use waffle"
author: "Hayley Corson-Dosch"
date: "2023-03-27"
output: html_document
---

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```



Define libraries here. 

```{r libraries, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) # includes ggplot
library(rvest)
library(ggwaffle)

# These are used for the layout of the viz
library(cowplot) # for laying out the final plot
library(sysfonts) # for text editing
library(showtext) # for adding in google fonts
library(magick) # for adding logo
```



## Load files

Scrape water use data


```{r load}
# Scrape water use data
national_wu_url <- 'https://waterdata.usgs.gov/nwis/water_use?format=html_table&rdb_compression=file&wu_year=ALL&wu_category=ALL&wu_category_nms=--ALL%2BCategories--'
# start session
session <- rvest::session(national_wu_url)
# pull data tables
site_tables <- rvest::html_table(session)
# wu data is in table 2
wu_table <- site_tables[[2]]

```

## Get data ready for plotting
  
This next section is for doing any pre-processing steps, data joins, etc, to get your data ready for plotting with ggplot. Remember that most of the time, ggplot prefers data in a "wide" format.


```{r processing}
# Munge water use data
# pivot to long format
wu_long <- wu_table %>%
  pivot_longer(!`National Totals`, names_to = 'year', values_to = 'vol_bgd', values_transform = list(vol_bgd = as.character))

# Filter out population and water source rows
wu_all <- wu_long %>%
  filter(!(`National Totals`) %in% c('Population, in millions', 'Total withdrawals, in Bgal/d',
                                     'Total Groundwater, fresh, in Bgal/d', 'Total Groundwater, saline, in Bgal/d',
                                     'Total Surface water, fresh, in Bgal/d', 'Total Surface water, saline, in Bgal/d')) %>%
  rename(wu_category = `National Totals`) %>%
  mutate(wu_category = gsub(', in Bgal/d|, In Bgal/d', '', wu_category))

# Get 2015 subset of data, omitting the commercial water use category, which has no data in 2015
wu_2015 <- wu_all %>%
  filter(year == 2015) %>%
  filter(!(wu_category == 'Commercial')) %>%
  mutate(vol_bgd = as.numeric(vol_bgd))

# Get an ordered list of water use categories (from high to low total volume of use)
wu_category_order <- wu_2015 %>%
  arrange(desc(vol_bgd)) %>%
  pull(wu_category)

# Use the ordered list to make water use category a factor in the 2015 water use data
wu_2015 <- mutate(wu_2015, wu_category = factor(wu_category, levels = wu_category_order))

# For use with ggwaffle, expand the data, so that the # of rows for each category = bgd of use (rounded to nearest whole value)
wu_2015_expanded <- wu_2015 %>%
  mutate(vol_bgd_rounded = round(vol_bgd)) %>%
  uncount(vol_bgd_rounded, .remove = FALSE)

# Use the ggwaffle function `waffle_uron()` to prep the data for plotting
wu_waffle <- waffle_iron(wu_2015_expanded, mapping = aes_d(group = wu_category), row = 13)

```

## Set up main plot

This chunk is where the main ggplot grob definition and set-up occurs. You may have several ggplot grobs, depending on your approach. For each one, if you define it with the `<-` assignment operator and surround the whole code statement in parentheses, you'll be able to preview here what the plot looks like before the next cowplot step.

> Note: The hardest part of this process is getting text to be the optimum size in the final `png` for Twitter. The font sizes here will **not** be the same as in the final image after the cowplot step. Make sure to check the output `png` for true font sizing and plot composition. 

```{r plotting, warning=FALSE}

# Define colors
background_color = "#ffffff"
font_color = "grey20"
waffle_colors <- RColorBrewer::brewer.pal(8, 'Paired')

# The background canvas for your viz
canvas <- grid::rectGrob(
  x = 0, y = 0, 
  width = 16, height = 9,
  gp = grid::gpar(fill = background_color, alpha = 1, col = background_color)
)

# Load in USGS logo (also a black logo available)
usgs_logo <- magick::image_read("../usgs_logo_black.png") 

# Generate sections of waffle chart for each category of water use, coloring one category per ggplot grob
(plot_list <- purrr::map(seq(1:length(wu_category_order)), function(category_id) {
  color_list <- rep(NA, length(wu_category_order))
  color_list[category_id] <- waffle_colors[category_id]
  ggplot(wu_waffle, aes(x, y, fill = group)) +
    geom_waffle() +
    coord_equal() +
    scale_fill_manual(values = color_list, na.value = NA) +
    theme_void() +
    theme(
      legend.position = 'None'
    )
})
 )



```


## Produce final plot

Here, use `cowplot` and `ggsave` to create the final viz for sharing out on Twitter. This template includes adding the USGS logo, title, text, etc.

**Make sure to use the format for saving your png with the date of the prompt that you've been assigned to!** (e.g., `20230401_part-to-whole_cnell.png`)

```{r cowplot, fig.width = 16, fig.height = 9}

# Load some custom fonts and set some custom settings
annotation_font <- 'Shadows Into Light'
sysfonts::font_add_google(annotation_font)
annotation_font_size <- 22
supporting_font <- "Source Sans Pro"
sysfonts::font_add_google(supporting_font)
showtext::showtext_opts(dpi = 300, regular.wt = 200, bold.wt = 900)
showtext::showtext_auto(enable = TRUE)

get_annotation_text <- function(category) {
  sprintf('%s (%s)', category, filter(wu_2015, wu_category == category) %>% pull(vol_bgd) %>% round(1))
}

# set plot spacing parameters
element_width <- 0.7
spacing <- 0.015
left_margin <- 0.1

# build plot
ggdraw(ylim = c(0,1), # 0-1 scale makes it easy to place viz items on canvas
       xlim = c(0,1)) +
  # a background
  draw_grob(canvas,
            x = 0, y = 1,
            height = 9, width = 16,
            hjust = 0, vjust = 1) +
  # the main plot
  purrr::map(seq(1:length(plot_list)), function(plot_id) {
    draw_plot(plot_list[[plot_id]],
              x = left_margin + spacing*(plot_id - 1),
              y = -0.03,
              width = element_width)
  }) +
  # legend
  geom_curve(aes(xend = 0.125, x = left_margin/2,
                 yend = 0.74, y = 0.69),
             arrow = grid::arrow(length = unit(1, 'lines')), 
             curvature = -0.3,
             angle = -35,
             color = font_color,
             lineend = 'round') +
  draw_label(paste('1 square =','1 billion gallons','per day', sep = "\n"),
             fontfamily = annotation_font,
             x = left_margin/4,   
             y = 0.63, 
             size = annotation_font_size,
             hjust = 0, 
             vjust = 0.5,
             color = font_color) +
  # give total
  draw_label(sprintf('Total water use = %s billion gallons per day', sum(wu_2015$vol_bgd) %>% round(0)),
             fontfamily = annotation_font,
             x = 0.5,   
             y = 0.14, 
             size = annotation_font_size,
             hjust = 0.5, 
             vjust = 0.5,
             color = font_color) +
  draw_line(
      x = c(0.126, 0.126, 0.88, 0.88),
      y = c(0.18, 0.17, 0.17, 0.18),
      color = 'grey40', 
      size = 0.5,
      linetype = 'solid'
  ) +
  # label thermo
  draw_label(get_annotation_text(wu_category_order[[1]]),
             fontfamily = annotation_font,
             x = 0.26,   
             y = 0.8, 
             size = annotation_font_size,
             hjust = 0.5, 
             vjust = 0.5,
             color = font_color) +
  # label irrigation
  draw_label(get_annotation_text(wu_category_order[[2]]),
             fontfamily = annotation_font,
             x = 0.515,   
             y = 0.8, 
             size = annotation_font_size,
             hjust = 0.5, 
             vjust = 0.5,
             color = font_color) +
  # label public supply
  geom_curve(aes(xend = 0.685, x = 0.715,
                 yend = 0.775, y = 0.885),
             arrow = grid::arrow(length = unit(1, 'lines')), 
             curvature = 0.25,
             angle = 55,
             color = font_color,
             lineend = 'round') +
  draw_label(get_annotation_text(wu_category_order[[3]]),
             fontfamily = annotation_font,
             x = 0.722,   
             y = 0.89, 
             size = annotation_font_size,
             hjust = 0, 
             vjust = 0.5,
             color = font_color) +
  # label industrial
  geom_curve(aes(xend = 0.75, x = 0.775,
                 yend = 0.775, y = 0.83),
             arrow = grid::arrow(length = unit(1, 'lines')), 
             curvature = 0.25,
             angle = 55,
             color = font_color,
             lineend = 'round') +
  draw_label(get_annotation_text(wu_category_order[[4]]),
             fontfamily = annotation_font,
             x = 0.782,   
             y = 0.83, 
             size = annotation_font_size,
             hjust = 0, 
             vjust = 0.5,
             color = font_color) +
  # label Aquaculture
  geom_curve(aes(xend = 0.81, x = 0.85,
                 yend = 0.693, y = 0.733),
             arrow = grid::arrow(length = unit(1, 'lines')), 
             curvature = -0.3,
             angle = -35,
             color = font_color,
             lineend = 'round') +
  draw_label(get_annotation_text(wu_category_order[[5]]),
             fontfamily = annotation_font,
             x = 0.83,   
             y = 0.758, 
             size = annotation_font_size,
             hjust = 0, 
             vjust = 0.5,
             color = font_color) +
  # label mining
  geom_curve(aes(xend = 0.85, x = 0.89,
                 yend = 0.285, y = 0.325),
             arrow = grid::arrow(length = unit(1, 'lines')), 
             curvature = -0.3,
             angle = -35,
             color = font_color,
             lineend = 'round') +
  draw_label(get_annotation_text(wu_category_order[[6]]),
             fontfamily = annotation_font,
             x = 0.87,   
             y = 0.35, 
             size = annotation_font_size,
             hjust = 0, 
             vjust = 0.5,
             color = font_color) +
  # label domestic
  geom_curve(aes(xend = 0.865, x = 0.905,
                 yend = 0.425, y = 0.465),
             arrow = grid::arrow(length = unit(1, 'lines')), 
             curvature = -0.3,
             angle = -35,
             color = font_color,
             lineend = 'round') +
  draw_label(stringr::str_wrap(get_annotation_text(wu_category_order[[7]]), 20),
             fontfamily = annotation_font,
             x = 0.885,   
             y = 0.51, 
             size = annotation_font_size,
             hjust = 0, 
             vjust = 0.5,
             color = font_color) +
  # label livestock
  geom_curve(aes(xend = 0.88, x = 0.92,
                 yend = 0.6, y = 0.64),
             arrow = grid::arrow(length = unit(1, 'lines')), 
             curvature = -0.3,
             angle = -35,
             color = font_color,
             lineend = 'round') +
  draw_label(get_annotation_text(wu_category_order[[8]]),
             fontfamily = annotation_font,
             x = 0.89,   
             y = 0.665, 
             size = annotation_font_size,
             hjust = 0, 
             vjust = 0.5,
             color = font_color) +
  # explainer text
  draw_label("National water use data: https://waterdata.usgs.gov/nwis/wu",
             fontfamily = supporting_font,
             x = 0.96,   
             y = 0.05,
             size = 14,
             hjust = 1,
             vjust = 0,
             color = '#6E6E6E')+
  # Title
  draw_label("U.S. water use - 2015",
             x = 0.04,
             y = 0.885,
             hjust = 0,
             vjust = 0,
             lineheight = 0.75,
             fontfamily = supporting_font,
             fontface = 'bold',
             color = font_color,
             size = 65) +
  # Add logo
  draw_image(usgs_logo, 
             x = 0.04,
             y = 0.05,
             width = 0.1, 
             hjust = 0, vjust = 0, 
             halign = 0, valign = 0)

# Save the final image in Twitter's 16 by 9 format
# !! Use format for saving with the date of your prompt: 
#         YYYYMMDD_prompt_name ()
# e.g. `20230101_part-to-whole-cnell.png`
ggsave(filename = "out/20230402_waffle_hcorson-dosch.png", 
       width = 16, height = 9, dpi = 300)
```

## Supporting information

### Key takeaways of this viz (1-2 sentences each)

1. As of 2015, thermoelectric power generation was the single largest category of water use in the United States, using 133 billion gallons per day. Water for thermoelectric power is used in the process of generating electricity with steam-driven turbine generators.

### Data source(s)

[National water use data](https://waterdata.usgs.gov/nwis/wu)

